\markright{Executive Summary}

# Executive summary {-}

## Stock{-}

This is an assessment of widow rockfish (*Sebastes entomelas*) that reside in the waters off California, Oregon, and Washington, from the U.S.-Canada border in the north to the U.S.-Mexico border in the south. Widow rockfish inhabit water depths of 25–370 m from northern Baja California, Mexico, to Southeastern Alaska. Although catches north of the U.S.-Canada border and south of the U.S.-Mexico border were not included in this assessment, it is not certain if those populations contribute to the biomass of widow rockfish off the U.S. West Coast, possibly through adult migration and/or larval dispersion. Following the 2015 benchmark assessment [@hicks_status_2015], this assessment is based on a single coastwide area model.


## Catches{-}

Historically, fisheries have caught widow rockfish since the turn of the 20th century. Landings in the trawl fishery are estimated to have increased into the 1940s and remained relatively constant and small (below 1,000 mt per year) throughout the 1950s and into the 1960s, before catches increased from a foreign trawl fleet in the 1970s, with a peak at almost 5,000 mt in 1967. Catches by a midwater trawl fleet increased rapidly in the late 1970s following the discovery that widow rockfish form large aggregations at night.

Total landings of widow rockfish peaked in the early 1980s, increasing from approximately 1,000 metric tons (mt) in 1978 to over 25,000 mt in 1981. The large landings in the early 1980s were curtailed with trip limits beginning in 1982, which resulted in a decline in landings throughout the 1980s and 1990s, following sequential reductions in the trip limits. From 2000 to 2003, landings of widow rockfish dropped from over 4,000 mt to about 40 mt and remained low through 2016. Landings increased rapidly following the quota share reallocation in 2017, and have been near or above 10,000 mt in all years between 2018 and 2024. Midwater trawl gears targeting rockfish and bycatch in the Pacific hake/whiting (*Merluccius productus*, hereafter "hake") fisheries account for the majority of the recent catch.

Widow rockfish are a desirable market species, and it is believed that discarding was low historically. However, management restrictions (e.g., trip limits) resulted in a substantial amount of discarding beginning in the early 1980s. Trawl rationalization was introduced in 2011. Between 2011 and 2024, very little discarding of widow rockfish has occurred. Catches for the past ten years are in @tbl-a and for the entire time series in @fig-a.

{{< pagebreak >}} 

```{r}
#| label: tbl-a
#| echo: false
#| warning: false
#| tbl-cap: "Recent total catch for the bottom trawl, midwater trawl, hake fleets. Catch combines landings and discards. The bottom trawl catch includes catches from fleets that were previously separated as net, and hook-and-line fisheries as described in the model bridging section."

library("flextable")
library("dplyr")
library("officer")
library("here")



exec_tab_style <- function(ft) {
  ft %>% 
  #font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 9, part = "all") %>% 
  bold(part="header") %>%
  hline(part="header") %>%
  hline_top(part="header") %>% hline_bottom() %>%
  align(align = "center", part = "header") %>%
  align(align = "center", part = "body")
}

# Function to read in lenght-1 .Rdata / .rda with name of choice
read_rdata <- function(path) {load(path); get(ls()[ls() != "path"])}

#Old table 12
table_A <- read_rdata(here("report", "tables", "exec_summ_tables", "catches_es.rda"))

# command below removes caption and converts list to just the dataframe
table_A <- table_A$table |> 
  mutate(across(is.numeric, round, digits = 1)) |> 
  dplyr::select(-dplyr::starts_with("ignore")) # filter out fleets 4 and 5 (formerly net and HnL)

table_A <- round(table_A, 2) # manually round bc flex table isn't working well

colnames_A <- c(
  "Year", "Bottom Trawl","Midwater Trawl","Hake", #"Net","Hook-and-line", 
  #"Total Landings", # no longer separate
  "Total Mortality"
)

flextable(table_A, col_keys = names(table_A)) %>%
  # set_header_labels(names = names(table_A), values = colnames_A) %>% 
  colformat_num(j = 1, big.mark = "", digits = 0)%>%
  fontsize(size = 9, part = "all") %>%
  set_table_properties(layout = "autofit") %>% 
  exec_tab_style


```


```{r}
#| label: fig-a
#| echo: false
#| warning: false
#| eval: true
#| fig-cap: "Total catches (landings and discards combined) of widow rockfish from 1916 to 2024 for bottom trawl, midwater trawl, and hake fleets."
#| out.width: "90%"
#| fig-align: "center"

knitr::include_graphics(
    file.path(here::here(),"figures/2025 base model r4ss plots/plots/catch2_landings_stacked.png")
)

```

## Data and Assessment{-}

```{r}
#| include: false

model_path <- here::here("models", "2025 base model")
report <- r4ss::SS_output(
  here("models", "2025 base model"), 
  printstats = FALSE, 
  verbose = FALSE,
  SpawnOutputLabel = "Spawning output (billions of eggs)"
)

```

The most recent benchmark assessment was conducted in 2015 [@hicks_status_2015], which was then updated in 2019 [@adams_status_2019]. This assessment began as a second update of the 2015 benchmark stock assessment and was initially reviewed by the Pacific Fishery Management Council (PFMC) Science and Statistical Committee (SSC) Groundfish Subcommittee (GFSC) in August 2025. The GFSC requested additional explorations and changes to be reviewed in October 2025. For the October 2025 supplemental review, the stock assessment team (STAT) provided all the requested materials, but due to the United States federal government shutdown that lasted from October 1, 2025, to November 12, 2025, the STAT and SSC members who are federal employees were unable to participate in the review meeting. At the subsequent November 2025 PFMC meeting, the SSC requested another review of the widow rockfish assessment to allow for full participation of the STAT and SSC members. Many of the changes made after the August 2025 model are outside the scope of the update assessment Terms of Reference, which resulted in a more parsimonious model raised to the current best practices in West Coast groundfish stock assessments.

This assessment uses the length- and age-structured modeling software Stock Synthesis (version 3.30.24.1). The coastwide population was modeled assuming separate growth and mortality parameters for females and males from 1916 to 2024.

The fleet structure in this assessment was streamlined compared to the 2015 benchmark and the 2019 update assessments to focus on major fisheries that remove more than 97% of the catch. These include: 1) a coastwide shore-based bottom trawl fleet (1916–2024), 2) a coastwide shore-based midwater trawl fleet that targets rockfish (1979–2024), and 3) a mostly midwater trawl fleet that targets hake and includes an at-sea fleet (1975–2024) and a domestic shore-based fleet (1991–2024). The bottom trawl fleet includes minimal catches made by net and hook-and-line gears, as well as widow rockfish catches made within the historical foreign fishery targeting Pacific Ocean Perch between 1966–1976. The model includes catch, age, and length data for all three fishing fleets. Catch in each fleet represents total removals. The discard amount for bottom trawl and midwater trawl fleets was reconstructed outside the model and added to the landings within the corresponding fleets.

There are three historical fishery-dependent CPUE indices retained from the 2015 benchmark: 1) Oregon bottom trawl (1984–1999), 2) at-sea foreign hake (1977–1988), and 3) at-sea domestic hake fleets (1983–1998). Data from three fishery-independent surveys were also included in the model: 1) length composition and an index for the \gls{s-tri} were retained from the 2015 benchmark (1977–2004), 2) length and conditional age-at-length compositions and an index for the \gls{s-wcgbt} (2003–2024), and 3) a recruitment index from the \gls{nmfs} \gls{s-juv} (2004–2024).

The base model estimated parameters for length-based selectivity for all fleets and surveys, a sex-specific length-at-age relationship, sex-specific natural mortality, and recruitment deviations. A Beverton-Holt stock-recruitment function was used to model productivity, and the steepness parameter was fixed at `r report$parameters["SR_BH_steep","Value"]` based on a steepness meta-analysis [@thorson_steep_2019] for West Coast rockfishes. The model also uses widow rockfish fecundity-at-length relationships from Dick et al. (2017) instead of assuming spawning output is proportional to body weight as in the 2015 and 2019 models. 

Natural mortality is the major source of uncertainty that is proposed as the axis of uncertainty to define states of nature in a decision table.

## Stock biomass and dynamics{-}

```{r}
#| include: false

if (!exists("report")) {
  model_path <- here::here("models", "2025 base model")
  report <- r4ss::SS_output(
    here("models", "2025 base model"), 
    printstats = FALSE, 
    verbose = FALSE,
    SpawnOutputLabel = "Spawning output (billions of eggs)"
  )
}

catch_sum <- aggregate(report$catch$Exp, by = list(Yr = report$catch$Yr), sum)

table_b <- read_rdata(here::here("report", "tables", "exec_summ_tables", "ssb_es.rda"))$table
names(table_b) <- c("Year", "SSB", "SSB_Lwr", "SSB_Upr", "Bratio", "Bratio_Lwr", "Bratio_Upr")

require("dplyr")

SSB <- report$derived_quants[report$derived_quants$Label %in% paste0("SSB_", 1916:2036),]
SSB$Year <- as.integer(gsub("SSB_", "", SSB$Label))

Bratio <- report$derived_quants[report$derived_quants$Label %in% paste0("Bratio_", 1916:2036),]
Bratio$Year <- as.integer(gsub("Bratio_", "", Bratio$Label))

```


The time series of estimated spawning output and the fraction of unfished spawning output are in @fig-b and @fig-c, respectively, and for the most recent year in @tbl-b. The spawning output declined rapidly with the developing domestic midwater fishery in the late 1970s and early 1980s, and is estimated to have been below the 25% of $B_0$ overfished reference point from `r Bratio |> filter(Value < 0.25) |> pull(Year) |> min()` to `r Bratio |> filter(Value < 0.25) |> pull(Year) |> max()` with a minimum of `r Bratio |> pull(Value) |> min() |> round(2) * 100`% in `r Bratio |> filter(Value == min(Value)) |> pull(Year)`. A combination of strong recruitment and low catches resulted in a steady increase in spawning output from 2000 through 2016, increasing above the 40% of $B_0$ target in `r Bratio |> filter(Value > 0.4 & Year > 2000) |> pull(Year) |> min()`. A target fishery for widow rockfish was reestablished in 2017. The stock exhibits a decline in the most recent years with the increased catches since 2017.

The 2025 spawning output relative to unfished equilibrium spawning output is `r paste0(round(table_b$Bratio[table_b$Year == 2025] * 100), "%")`, above the target of 40% (95% confidence interval of `r round(table_b$Bratio_Lwr[table_b$Year == 2025] * 100)` - `r round(table_b$Bratio_Upr[table_b$Year == 2025] * 100)`%). 

Spawning output is estimated to be at `r format(SSB$Value[SSB$Year == 2025], big.mark = ",", digits = 1)` billion eggs in 2025 (95% confidence interval of `r paste(format(table_b[table_b$Year == 2025, c("SSB_Lwr", "SSB_Upr")], big.mark = ",", digits = 1), collapse = " - ")` billions of eggs). The uncertainty in the estimated spawning output is high, especially in the early years and at the end of the time series (currently and in projections).

```{r}
#| label: fig-b
#| echo: false
#| warning: false
#| eval: true
#| fig-cap: "Estimated female spawning output time-series from the base model (solid line) with asymptotic 95% confidence interval (dashed lines)."
#| out.width: "90%"
#| fig-align: "center"

knitr::include_graphics(
    file.path(here::here(),"figures/2025 base model r4ss plots/plots/ts7_Spawning_output_with_95_intervals.png")
)

```

```{r}
#| label: fig-c
#| echo: false
#| warning: false
#| eval: true
#| fig-cap: "Estimated relative spawning output (depletion) with asymptotic 95% confidence interval (dashed lines) for the base model."
#| out.width: "90%"
#| fig-align: "center"

knitr::include_graphics(
    file.path(here::here(),"figures/2025 base model r4ss plots/plots/ts9_Fraction_of_unfished_spawning_output_with_95_asymptotic_intervals_intervals.png")
)

```
{{< pagebreak >}} 
```{r}
#| label: tbl-b
#| echo: false
#| warning: false
#| tbl-cap: "Recent trend in estimated female spawning output (billions of eggs) and fraction of unfished spawning output (relative spawning output)."

#Old table 12
table_B <- read_rdata(here::here("report", "tables", "exec_summ_tables", "ssb_es.rda"))$table

names(table_B)[2] <- "Spawning output"

table_B <- table_B |> 
  mutate(across(2:4, ~ round(.x, 0))) |> 
  mutate(across(tidyselect::last_col(3):last_col(), ~ round(.x, 3)))

flextable(table_B) |> 
  colformat_num(j = 1, big.mark = "", digits = 0)%>%
  exec_tab_style %>% 
  width(j = 1, width = 0.4) %>% 
  width(j = 2, width = 0.6) %>%
  width(j = 3:4, width = 0.95) %>%
  width(j = 5:7, width = 0.8)
```

{{< pagebreak >}} 

## Recruitment{-}

Recruitment deviations were estimated for the entire time series modeled. There is little information regarding recruitment prior to 1965, and the uncertainty in these estimates is expressed in the model. Estimates of recruitment appear to be episodic and characterized by periods of low recruitment, with several very large, but uncertain, year classes (@fig-d, @fig-e).

The 2019 update assessment [@adams_status_2019] estimated the largest recruitment on record in 2013, and above-average recruitment in 2014. With a more parsimonious fleet structure in this assessment, and the additional years of data, 2013 is now estimated as only the tenth-largest recruitment on record, and 2014 is estimated as below average.


```{r}
#| label: fig-d
#| echo: false
#| warning: false
#| eval: true
#| fig-cap: "Time-series of estimated recruitments (medians as open circles) for the base model with asymptotic 95% confidence interval (vertical bars). Estimated unfished equilibrium recruitment (R~0~) is indicated with a filled circle."
#| out.width: "90%"
#| fig-align: "center"

knitr::include_graphics(
    file.path(here::here(),"figures/2025 base model r4ss plots/plots/ts11_Age-0_recruits_(1000s)_with_95_asymptotic_intervals.png")
)

```

{{< pagebreak >}} 

```{r}
#| label: tbl-c
#| echo: false
#| warning: false
#| tbl-cap: "Recent estimated trend in widow rockfish recruitment with asymptotic 95% confidence intervals determined from the base model."
#| out.width: "90%"
#| fig-align: "center"

table_c <- read_rdata(here::here("report", "tables", "exec_summ_tables", "recr_es.rda"))$table

colnames(table_c) <- c(
  "Year", 
  "Recruitment (1,000s)", "Lower Interval (1,000s)", "Upper Interval (1,000s)", 
  "Recruitment Deviations", "Lower Interval", "Upper Interval" 
)

table_c <- table_c %>%
  mutate(across(2:4, ~ round(.x, 0)))|>
  mutate(across(tidyselect::last_col(3):last_col(), ~ round(.x, 3))) 

flextable(table_c) %>%
  colformat_num(j = 1, big.mark = "") %>%
  colformat_num(j = 3, big.mark = ",")%>%
  exec_tab_style %>%
  width(j = 1, width = 0.4) %>%
  width(j = 2:7, width = 0.8)

```
{{< pagebreak >}} 

## Exploitation status{-}

The population declined throughout the 1980s and 1990s when fishing intensity was above the management target. The population then increased between 2002 and 2016 when fishing intensity was well below the management target. Fishing intensity increased substantially in 2017, and is estimated to have been above the management target from 2018 onward, though the relative spawning output remains above the management target of 40%. Recent fishing intensities are in @tbl-d, and a full time series of relative spawning output and fishing intensity is in @fig-e.

```{r}
#| label: tbl-d
#| echo: false
#| warning: false
#| tbl-cap: "Estimated recent trend in relative fishing intensity ((1-SPR)/(1-SPR~50%~)) and exploitation rate (as the proportion of age 4+ biomass) with asymptotic 95% confidence intervals for both quantities."

table_d <- read_rdata(here("report", "tables", "exec_summ_tables", "spr_es.rda"))$table

table_d <- table_d|>
mutate(across(2:7, ~ round(.x, 3)))

flextable(table_d) %>%
  colformat_num(j = 1, big.mark = "") %>%
  exec_tab_style %>% 
  width(j = 1, width = 0.4) %>%
  width(j = 2, width = 0.8) %>%
  width(j = 3:7, width = 0.9)
  
```

{{< pagebreak >}} 

```{r}
#| label: fig-e
#| echo: false
#| warning: false
#| eval: true
#| fig-cap: "Phase plot of relative fishing intensity ((1-SPR)/(1-SPR~50%~)) versus fraction unfished (spawning output, (B/B~0~). Lines through the final point show marginal asymptotic 95% intervals for each dimension. The shaded ellipse is an approximate 95% confidence region which accounts for the estimated covariance between the two quantities."
#| out.width: "90%"
#| fig-align: "center"

knitr::include_graphics(
    file.path(here::here(),"figures/2025 base model r4ss plots/plots/SPR4_phase.png")
)

```

{{< pagebreak >}} 

## Ecosystem considerations{-}

Recruitment is a key mechanism by which the ecosystem may directly impact the population dynamics of widow rockfish; however, the specific pathways through which environmental conditions exert influence on widow rockfish dynamics are unclear. Changes in the environment may result in changes in other population processes as well. Unfortunately, there are few data available for widow rockfish that provide insights into these effects. 

Fishing has effects on habitats, in addition to the population itself. Rockfish are often associated with habitats containing living structures such as sponges and corals, which fishing may threaten, although the midwater fishery currently targeting widow rockfish is expected to have fewer impacts on living structures than the historical bottom trawl fishery. Recent studies on essential fish habitat are beginning to characterize important locations for rockfish throughout their life history; however, there is little current information available to evaluate the specific effects of fishing on the ecosystem characteristics pertinent to the management of widow rockfish.

## Reference points{-}

```{r}
#| include: false

table_e <- read_rdata(here::here("report", "tables", "exec_summ_tables", "reference_points.rda"))$table

ref_pts <- table_e
colnames(ref_pts) <- c("Label", "Est", "Lwr", "Upr")

# Convert the reference points data frame into a named list of lists,
# where each entry contains the formatted estimate, lower, and upper values for each reference point label.
ref_pts <- setNames(
  lapply(seq_len(nrow(ref_pts)), function(i) {
    vals <- ref_pts[i, c("Est", "Lwr", "Upr")]
    vals <- format(round(as.numeric(vals)), big.mark = ",", scientific = FALSE)
    setNames(as.list(vals), c("Est", "Lwr", "Upr"))
  }),
  nm = gsub("\\\\%", "%", ref_pts$Label)
)
```

Reference points were calculated using the estimated selectivity parameters and average catch distribution among fleets in the most recent five years of the model (2019--2024). A list of estimates of the current state of the population, as well as reference points based on 1) a target unfished spawning output of 40%, 2) a \gls{spr} of 0.5, and 3) the model estimate of maximum sustainable yield (MSY), are all listed in @tbl-e. The \gls{spr} is the expected lifetime reproductive output of a female recruit under a given fishing intensity as a proportion of the unfished expected lifetime reproductive output. 

While the estimate of population scale (total spawning output) has remained relatively steady across the 2015, 2019, and 2025 assessments, unfished recruitment ($R_0$) and natural mortality ($M$), which are positively correlated, have declined from 2015 (female $M$ of 0.157 yr^-1^) to 2019 (0.144 yr^-1^) to 2025 (`r round(report$M_at_age[1, 4], 3)` yr^-1^). Thus, although the population scale is estimated consistently, additional years of data (in particular, age composition data) following the reopening of the fishery have led the model to estimate that the population is less productive, which results in lower yields at all three reference points (proxy based on spawning output, proxy based on fishing intensity, and model estimate of MSY).

{{< pagebreak >}} 


```{r}
#| label: tbl-e
#| echo: false
#| warning: false
#| tbl-cap: "Summary of reference points and management quantities for the base model."

table_e <- table_e |> mutate(across(where(is.numeric), \(x) {
  ifelse(
    x > 1, 
    format(round(x, digits = 0), big.mark = ","), 
    round(x, digits = 3)
    )
}))


flextable(table_e) %>%
  merge_h(part = "header") %>%
  merge_v(part = "header") %>%
  border(part = "all") %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = 1) %>%
  #font(fontname = "Times New Roman", part = "all") %>%
  bold(part = "header")%>%
  bold(i = c(6,11,16), part="body")%>%
  colformat_num( digits = 2)%>%
  fontsize(size = 9, part = "all")%>%
  set_table_properties(layout = "autofit")

```

{{< pagebreak >}} 

## Management performance{-}

Since annual catch limits for widow rockfish increased in 2017 and new fishing opportunities for use of midwater trawl gear became available, attainment of the annual catch limit has been high (@tbl-f). Specifically, attainment of the \gls{acl} has exceeded 70% every year since 2018, averaged 77% from 2017--2024, and was as high as 88% in 2022. 

```{r}
#| label: tbl-f
#| echo: false
#| warning: false
#| tbl-cap: "Recent trend in total catch and commercial landings (mt) relative to the management guidelines. Estimated total catch reflects the commercial landings plus the model estimated dead discarded biomass."

table_f <- read_rdata(here("report", "tables", "exec_summ_tables", "recent_management.rda"))$table

table_f <- table_f |> 
  mutate(across(contains("mt"), \(x) format(round(x, 0), big.mark = ",")))

flextable(table_f) |> 
  colformat_num(j = 1, big.mark = "", digits = 0) %>%
  exec_tab_style %>%
  width(j = 1, width = 0.5) %>%
  width(j = -1, width = 1)
```

## Harvest projections and decision table{-}

```{r}
#| include: false

##dec table params
dec_table_pars <- report$parameters|> #make sure to ppull from report, not input file so est values are used
  mutate(
         value = Value,
         sd = if_else(is.na(Parm_StDev),Pr_SD,Parm_StDev))|> #If est par, use param sd, otherwise use prior sd
  filter(Num %in% c(1,13,24))|> #natM Female, natM male, steepness
  select(value,sd)|>
  mutate(
    q125 = round(qnorm(0.125,mean = value,sd = sd),3), #apply the quantiles
    q875 = round(qnorm(0.875,mean = value,sd = sd),3),
    phase = -1,
    name = c("NatM_p_1_Fem_GP_1", #name the parms - easier to do manually as report and r4ss::read() have different parameter names
             "NatM_p_1_Mal_GP_1",
             "SR_BH_steep") 
  )

#dec tables stuff
dec_table_pars["NatM_uniform_Fem_GP_1","q125"]
colnames(dec_table_pars)

dec_table_res <- read.csv(here("data_derived", "decision_table", "dec_table_formatted.csv"))

##gmt stuff
gmt_catch <- data.frame(read.csv(here("data_provided", "gmt_forecast_catch", "GMT_forecast_catch.csv"))|>
  group_by(year)|>
  summarise(total_catch = sum(catch_mt)))

dec_table_summ <- list(
  "p45_2027" = dec_table_res|>filter(scenario == "P*0.45" & yr == 2027)|>pull(catch),
  "p45_2036" = dec_table_res|>filter(scenario == "P*0.45" & yr == 2036)|>pull(catch),
  "p40_2027" = dec_table_res|>filter(scenario == "P*0.40" & yr == 2027)|>pull(catch),
  "p40_2036" = dec_table_res|>filter(scenario == "P*0.40" & yr == 2036)|>pull(catch)
)

ctch_rng <- dec_table_res |> 
  filter(yr >= 2027 & scenario != "Constant catch") |> 
  slice(c(which.min(catch), which.max(catch)), .by = "scenario") |> 
  mutate(stat = ifelse(catch == max(catch), "max", "min"), .by = "scenario")
```

Projections for overfishing limit (OFL) (mt), ABC (mt), ACL (mt), the buffer between the OFL and ABC, estimated spawning output (billions of eggs), and fraction of unfished spawning output with adopted OFL and ACL and assumed catch for the
first two years of the projection period. The predicted OFL is the calculated total catch determined by FSPR=50% which are provided in @tbl-g.

{{< pagebreak >}} 

```{r}
#| label: tbl-g
#| echo: false
#| warning: false
#| tbl-cap: "Potential \\gls{ofl} (mt), \\gls{abc} (mt), \\gls{acl} (mt), the buffer between the \\gls{ofl} and \\gls{abc}, estimated spawning output (billions of eggs), and fraction of unfished spawning output with adopted \\gls{ofl} and \\gls{acl} and assumed catch for the first two years of the projection period. The predicted \\gls{ofl} is the calculated total catch determined by FSPR=50%."

table_g <- read_rdata(here::here("report", "tables", "exec_summ_tables", "projections.rda"))$table

# get buffers from forecast file
fcst <- r4ss::SS_readforecast(here(model_path, "forecast.ss"), verbose =  FALSE)
buffers <- fcst$Flimitfraction_m

# which rows in the table are missing buffer and ABC values?
NA_rows <- is.na(table_g[["Buffer"]])

# fill in Buffer column from forecast file values
table_g[["Buffer"]][NA_rows] <- buffers[["fraction"]][match(table_g[["Year"]][NA_rows], buffers[["year"]])]

# fill in ABC as product of Buffer and OFL
table_g[NA_rows, "ABC (mt)"] <- table_g[["Buffer"]][NA_rows] * table_g$"OFL (mt)"[NA_rows]

table_g <- table_g |> 
  mutate(
    across(contains("mt"), \(x) ifelse(
      is.na(x), NA, format(round(x, 0), big.mark = ",")
    )),
    `Fraction Unfished` = round(`Fraction Unfished`, 3)
  )

#replace NA with dashes
table_g[is.na(table_g)] <- "-"

#Replace OFL, bufferm ABC and ACL for 2025, 2026
table_g[c(1,2), c("OFL (mt)", "Buffer", "ABC (mt)", "ACL (mt)")] <- "-"

names(table_g)[names(table_g) == "Spawning Biomass (mt)"] <- "SB (mt)"

flextable(table_g) %>%
  colformat_num(j = 1, big.mark = "", digits = 0) %>%
  exec_tab_style() %>% 
  width(j = 1, width = 0.25) %>%
  width(j = 2:10, width = 0.55) %>%
  fontsize(size = 8, part = "all")

```

Uncertainty in natural mortality (estimated for both sexes) contributed greatly to uncertainty in the results. Natural mortality is proposed to be used as the axis of uncertainty to define low and high states of nature for the decision table. This differs from the 2019 assessment, which included two more factors: steepness (not estimated in the model) and 2013 recruitment strength. Steepness was fixed in the base model at the mean of a prior that has not changed since 2019, and the 2013 year class is no longer a major source of uncertainty, and there is no recent similarly large estimate of recruitment. The 12.5% and 87.5% quantiles for female and male natural mortality (independently) are proposed to be used as low and high values (`r dec_table_pars["NatM_uniform_Fem_GP_1","q125"]` yr^-1^ and `r dec_table_pars["NatM_uniform_Fem_GP_1","q875"]` yr^-1^ for females; `r dec_table_pars["NatM_uniform_Mal_GP_1","q125"]` yr^-1^ and `r dec_table_pars["NatM_uniform_Mal_GP_1","q875"]` yr^-1^ for males). The decision table with projections under alternative catch streams is to be developed during the review meeting.


```{r}
#| label: tbl-h
#| include: false
#| echo: false
#| warning: false
#| tbl-cap: "Summary table of 12-year projections beginning in 2025 for alternate states of nature based on the axis of uncertainty (a combination of M, h). Columns range over low, mid, and high state of nature, and rows range over different assumptions of total catch levels (discards + retained). Catches in 2025 and 2026 are allocated using the percentage of landings for each fleet in 2019--2024."

table_h <- read.csv(here("data_derived", "decision_table", "dec_table_formatted.csv"))

# Merge rows manually because merging via flextable causes 
# alignment issues when table spans multiple pages
table_h <- table_h |> 
  filter(scenario != "Constant catch") |> 
  mutate(across(starts_with("SpawnBio"), \(x) round(x, 0))) |> 
  add_rownames() |> 
  mutate(scenario = case_when(
    scenario == "P*0.45" & rowname == "14" ~ "ACLp*=0.45,", 
    scenario == "P*0.45" & rowname == "15" ~ "sigma=0.50",
    scenario == "P*0.40" & rowname == "2" ~ "ACLp*=0.40,",
    scenario == "P*0.40" & rowname == "3" ~ "sigma=0.50"
  )) |> 
  select(-rowname)

flextable(table_h) |>
  set_header_labels(values = c(scenario = "Management decision",
                               yr = "Year",
                               catch = "catch (mt)",
                               SpawnBio_low = "SB (mt)",
                               dep_low = "Depletion (%)",
                               SpawnBio_base = "SB (mt)",
                               dep_base = "Depletion (%)",
                               SpawnBio_high = "SB (mt)",
                               dep_high = "Depletion (%)")
                    ) |>
  add_header_row(values = c("", "Low", "Base", "High"), colwidths = c(3, 2, 2, 2)) |>
  add_header_row(values = c("", "State of nature"), colwidths = c(3, 6)) |>
  align(align = "center", part = "all") |>  # |> valign(j = 1, valign = "top") |>
  # font(fontname = "Times New Roman", part = "all") |>
  fontsize(size = 8, part = "all") |>
  padding(padding.top = -2, part = "all") |>
  bold(part = "header", i = 1:2) |>
  bold(part = "header", i = 3, j = 1) |>
  colformat_num(j = 2, big.mark = "", digits = 0) |>
  border_inner(part = "header") |> border_outer() |>
  vline(j = c(1, 3, 5, 7)) |> 
  #merge_at(i = 1:12, j = 1) |> merge_at(i = 13:24, j = 1) |>
  width(width = 0.35, j = 2) |> 
  width(width = 0.55, j = c(3, 4, 6, 8)) |> 
  width(width = 0.6, j = c(5, 7, 9)) |>
  width(width = 0.8, j = 1)|>
  hline(i = 12) |>
  hline_bottom(j = 9)|>
  line_spacing(space = 0.8, part = "all")


```

## Scientific uncertainty{-}

The model estimate of the log-scale standard deviation of the \gls{ofl} in 2025 is `r round(report$OFL_sigma, 3)`. This is less than the default SSC value of 0.5 for a category 1 assessment, so harvest projections assume an initial sigma of 0.5.


## Unresolved problems and major uncertainties{-}

```{r echo = FALSE, results = 'hide', message = FALSE, warning = FALSE }

steep_proflik <- read.csv(here::here("report", "tables", "SR_BH_steep_quant_table.csv"))
names(steep_proflik)[1] <- "var"

steep_yspr50 <- as.numeric(steep_proflik[steep_proflik$var == "Yield SPR 50",-1])
steep_yspr50 <- round(steep_yspr50[round(steep_yspr50) > 0])

```

Major sources of uncertainty include natural mortality, stock productivity, and spatial dynamics, which are discussed below. 

This assessment attempts to capture uncertainty in $M$ by estimating it inside the model (thereby propagating uncertainty in $M$ into uncertainty in spawning output and other derived quantities), and by incorporating variation in $M$ in a decision table. Model sensitivities and profiles over $M$ showed that current stock status was highly sensitive to the assumption about natural mortality.

Widow rockfish is a relatively long-lived fish, and its natural mortality ($M$) is likely to be lower than many managed fish stocks (e.g., gadoids). Ages above 50 years have been observed, and it is expected that natural mortality could be less than 0.10 yr^-1^ (the median of the prior for natural mortality used in this assessment). However, even with length and age data available back to the late 1970s, $M$ was estimated at `r round(report$parameters["NatM_uniform_Fem_GP_1",]$Value, 3)` for females and `r round(report$parameters["NatM_uniform_Mal_GP_1",]$Value, 3)` yr^-1^ for males, with a small amount of uncertainty (e.g., a `r paste0(round(100 * report$parameters["NatM_uniform_Fem_GP_1",]$Parm_StDev/report$parameters["NatM_uniform_Fem_GP_1",]$Value, 1), "%")` coefficient of variation for females). 

Notably, the estimated $M$ for both sexes from this assessment is lower than those from the 2015 and 2019 assessments, for which natural mortality was estimated above 0.15 yr^-1^ and 0.14 yr^-1^, respectively. The estimates of $M$ varied slightly depending on the weight given to age and length data, or removing recent years of data, but female $M$ was always estimated above 0.12 yr^-1^. The likelihood profile over natural mortality provides support for values up to or above 0.14 yr^-1^, but with greater curvature than in the 2019 assessment, suggesting additional data has reduced the support for higher natural mortality values. A contributing factor to this change is the increased mean age and frequency of older fish in the catch observed by the \gls{s-wcgbt} survey in recent years. The successful rebounding of the stock as a result of reduced fishing pressure from 2003--2016 allowed more individuals to reach older ages than observed in the earlier years of the \gls{s-wcgbt}. However, consideration should be given in future assessments to structural changes which may improve the fits to \gls{s-wcgbt} age data.

Steepness was fixed at 0.720 in the base model, but a likelihood profile showed that it would be estimated at a value less than that. Estimates of $M$ increased with lower steepness, while unfished spawning output increased and current spawning output decreased. Sustainable yields at the SPR~50%~ reference harvest rate ranged from approximately `r min(steep_yspr50)` to `r max(steep_yspr50)` mt depending on the value of steepness.

A greater understanding of the spatial dynamics of widow rockfish would help with interpretation of patterns in the existing data and help improve future assessment models. In particular, there are opposing trends in the age distribution sampled in recent years from the target fishery, which is concentrated in a subset of the range of the stock, and the coastwide age distribution \gls{s-wcgbt}. These differences could be due to some combination of localized depletion, spatial variability in age relative to benthic vs. pelagic habitat, spatial distribution relative to untrawlable habitat, and the rate of mixing among areas.

## Research and Data Needs{-}

There are many areas of research that could be improved to benefit the understanding and assessment of widow rockfish, including the following:

- **Natural mortality:** Uncertainty in natural mortality translates into uncertain estimates of status and sustainable fishing levels for widow rockfish. The collection of additional age data, re-reading of older age samples, reading old age samples that are as of yet unread, and improved understanding of the life history of widow rockfish may reduce that uncertainty. Investigating the ageing error and bias would help to understand the influences that the age data have on this assessment, particularly the influence of age data on natural mortality.

- **Sex-specific selectivity:** The midwater and bottom trawl length-composition data fits showed divergent residual patterns between male and female fish. The underlying mechanism driving this pattern is unclear and could be related to growth, sexing error, or to sex-specific selectivity (e.g., when widow rockfish aggregate, sexes possibly may be aggregating separately). Sex-specific selectivity for these two fleets could be explored or included to address this.

- **Coastwide understanding of stock structure, biology, connectivity, and distribution:** This is a stock assessment for widow rockfish off of the west coast of the U.S. and does not consider data from British Columbia or Alaska. Further investigating and comparing the data and predictions from British Columbia and Alaska to determine if there are similarities with the U.S. West Coast observations would help to define the connectivity between widow rockfish north and south of the U.S.-Canada border. Research into spatial dynamics within the U.S. West Coast (as noted above under Unresolved problems and major uncertainties) is also important.