# Data

Many sources of data were available for this assessment (@fig-data-plot-base), including indices of abundance, landings, discards, and length and age observations from fishery-dependent and fishery-independent sources. Data used in this assessment are described below. No new data sources were considered in this update assessment.

## Fishery-independent data

Data from three fishery-independent surveys were used in this assessment: 1) the SWFSC and NWFSC/PWCC Midwater Trawl Survey (hereafter, “juvenile survey”); 2) the Alaska Fisheries Science Center (AFSC)/NWFSC Triennial Shelf Trawl Survey (hereafter, “triennial survey”); and 3) the NWFSC West Coast Groundfish Bottom Trawl Survey (hereafter, “WCGBTS”). Depth and latitude strata used to analyze the catch-rates, length compositions, and age compositions are the same as in the 2019 update assessment [@adams_status_2019] and shown in @tbl-strat-surv.

### Triennial Survey

The Alaska Fisheries Science Center/Northwest Fisheries Science Center West Coast Triennial Shelf Survey (Triennial Survey) was first conducted by the Alaska Fisheries Science Center (AFSC) in 1977 and was conducted every three years, ending 2004. The survey’s design and sampling methods are most recently described in [@weinberg_estimation_2002].

The time series suggests a possible slightly increasing trend in biomass from 1980–1983, although is relatively flat until the end of the period in 2001 and 2004 when the index declines significantly. The index and length compositions for this historical Triennial survey were unchanged from the previous assessment. Please refer to the most recent benchmark assessment [@hicks_status_2015] and most recent update assessment [@adams_status_2019] for details on those data inputs.

### West Coast Groundfish Bottom Trawl Survey

The Northwest Fisheries Science Center West Coast Groundfish Bottom Trawl Survey (\gls{s-wcgbt}) has been conducted annually, since 2003. It is based on a random-grid design; covering the coastal waters between depths of 55–1,280 m [@bradburn_2003_2011]. No survey occurred in 2020 due to Coronavirus pandemic (COVID-19). Widow rockfish are not commonly caught in the WCGBTS. Higher catch rates occur north of 40° N latitude and catches are rare south of 36° N latitude. Few large fish are found shallower than 100 m and few small fish are found in the deeper water of the slope. There is no clear trend in length with latitude other than smaller fish tend to occur south of approximately 36° N latitude, and there appears to be some very small fish found near 39° N latitude.

For this assessment, geostatistical models of biomass density were fit to survey data using the R package \gls{sdmtmb} [@anderson_sdmtmb_2022]. This is an updated approach compared to the 2015 benchmark assessment (non-spatial delta-GLMM) and the 2019 update assessment (VAST delta-lognormal model). The \gls{sdmtmb} model used a 200 knot mesh of the survey area [@thorson_geostatistical_2015]. The prediction grid was truncated to only include available survey locations in depths between 55--500 m to limit extrapolating beyond the data and edge effects. Tweedie, delta-binomial, delta-gamma, and mixture distributions, which allow for extreme catch events, were investigated. The positive catch weight model includes survey pass ('first' for early season or 'second' for late season) and year. Vessel-year effects, which have traditionally been included in index standardization for this survey, were not included as the estimated variance for the random effect was close to zero. Vessel-year effects were more prominent when models did not include spatial effects and were included for each unique combination of vessel and year in the data to account for the random selection of commercial vessels used during sampling [@helser_generalized_2004; @thorson_accounting_2014]. 

Results are shown for the delta-gamma and delta-lognormal distributions, which reported the best diagnostics among the explored models (@fig-qq-sdmtmb). Both models converged (positive definite Hessian matrix) but predicted data from both models showed slightly right-heavy tails compared to the assumed likelihoods, with the gamma model having stronger divergence. Spatiotemporal estimates of biomass from the delta-lognormal model were then converted into annual indices using sdmtmb::get_index() function by integrating across the spatial domain of the survey [@anderson_sdmtmb_2022].

Overall, the delta-lognormal index estimates is more comparable to the 2019 spatiotemporal VAST-based index than the delta-gamma index, and less influenced by extreme catch events, particularly in 2013 and 2016; for these reasons, in addition to better model performance observed above, the delta-lognormal sdmTMB-based index was used for the base model in this assessment. The delta-lognormal mean value (2262.824) was slightly lower than the means of the index values used in the 2015 benchmark assessment (2701.12) and the 2019 update assessment (3301.765). However, since these are used as relative indices, these differences in mean values have no impact in themselves on the outcome of the assessment. Comparisons of the different error structures, design-based estimate and the VAST index used in 2019 are in @fig-absurv-compare. 

Length, age, and conditional age-at-length compositions were processed using the nwfscSurvey package in R publicly available on GitHub [@wetzel_nwfscsurvey_2025]. Length compositions were created by expanding to the tow and summing to give a strata specific composition [@tbl-sample-age-gear-state]. Strata definition was retained from the benchmark assessment. The strata compositions were combined to a coastwide composition using a design-based index of abundance from each stratum. The design based index is constructed by taking the average catch per unit effort (CPUE) defined as catch per area swept across tows in each stratum and year. The sum of strata specific composition data was then calculated, weighting by the average CPUE per stratum multiplied by the area of each stratum. Age distributions were included in the model as conditional-age-at-length (CAAL) observations. The marginal age-compositions were also included, but only for easier viewing of strong cohorts. The CAAL data were not expanded and were binned according to length, age, sex, and year.

The input sample sizes for length and marginal age-composition data were calculated based on [@stewart_bootstrapping_2014]. The input sample size of CAAL data was set at the number of fish at each length by sex and by year. Expanded length frequencies from this survey show intermittent years of small fish; the 2018–2024 period generally suggests most fish are around 40–45cm in length (@fig-exp-lencomp-WCBTS). Strong cohorts are not immediately apparent and it seems that ageing error may result in some variability between years. Conditional age-at-length proportions (@fig-cond-agelen-WCBTS) show relatively consistent length-at-age with few outliers.

### Juvenile Survey

An updated coastwide pre-recruit index of abundance for 2001-2024 for widow rockfish was created  using data from three midwater trawl surveys targeting young-of-the-year (YOY) rockfish (\gls{s-juv}), provided by Tanya Rogers (SWFSC, pers. comm.). All surveys used identical gear, enabling the construction of a consistent coastwide index spanning from 36°N to the U.S./Canada border since 2004. Sampling in 2020 was limited due to the COVID-19 pandemic and excluded from all models. In 2010 and 2012, coverage was incomplete, so these years excluded from the final model to align with the 2019 assessment. Following the 2015 and 2019 assessments, data from 2001–2003 were also excluded due to limited spatial coverage (36°30′ to 38°20′ N latitude). 

The index was built using a spatial GLM with the sdmTMB package [@anderson_sdmtmb_2022], modeling 100-day standardized catch-per-tow as a function of year (fixed effect), Julian date (GAM smoother, k = 4), spatial random field, and spatiotemporal random effects. Models with Tweedie, delta-lognormal, and delta-gamma error structures were compared; DHARMa residuals and simulation-based diagnostics indicated the Tweedie model performed best. The index shows a moderately strong increasing trend in juvenile abundance from 2017 to 2023, with a slight decline in 2024 (@fig-surv-abundest). Recent values remain high relative to the previous decade, and uncertainty estimates support the robustness of this trend.

## Fishery-dependent data

Widow rockfish is primarily caught by bottom trawl and midwater trawl gears (@tbl-all_removals). This species also commonly bycaught by fishery targeting Pacific hake/whiting (*Merluccius productus*, hereafter "hake"). Minimal amounts have been historically taken by non-trawl gears such as net and hook-and-line, which represented 1.4% and 1.1% of catches, respectively.

The fleet structure in this assessment was streamlined compared to the 2015 benchmark and the 2019 update assessments. The catches in the model are divided among three fishing fleets:

1. A coastwide shore-based bottom trawl fleet (1916–2024). This fleet also includes minimal catches made by net and hook-and-line gears, as well as widow rockfish catches made within historical foreign fishery targeting Pacific Ocean Perch between 1966–1976. 
2. A coastwide shore-based midwater trawl fleet that targets rockfish (1979–2024).
3. A mostly midwater trawl fleet that targets hake and includes at-sea fleet (1975–2024) and a domestic shore-based fleet (1991–2024).

Catches by fleet and year are provided in @tbl-all_removals. As in previous assessments, catches from Puget Sound and those from commercial shrimp trawls, commercial pots, and recreational fisheries were excluded (as these are generally minimal).

### Landings

Landings from the 1916-2018 period were carried forward into this assessment with two modifications. One was made to the midwater and bottom trawl catches from California. Because PacFIN appears to underestimate midwater trawl catches in California in 1979-1980 when midwater trawl fishery for widow rockfish developed (Edward Dick, SWFSC, pers. comm.) we adjusted midwater and bottom trawl catches from California in these years to reflect the ratio of California midwater to bottom trawl catches in 1981-1982. 

The second update was the use of new catch reconstruction for widow rockfish, provided by the Washington Department of Fish and Wildlife (WDFW). New Washington historical landings represent the best scientific information available. The three main sources used in this reconstruction included the US Fish Commission Report (UFSC), Washington Bound Volumes, and Washington Statistical Bulletin. The historical species composition was based on the various historical reports and interviews of fishermen and dockside samplers. The landings between 1981 and 2000 were also provided by WDFW, since WDFW developed and used an improved method for apportioning unidentified rockfish (URCK) category in fish tickets to the individual species landings. This improved approach relaxed the borrowing rules for missing data used in the WDFW species allocation algorithm that feeds into PacFIN. 

Recent catches (2019-onward) were extracted from PacFIN for commercial shorebased data and NORPAC for at-sea hake fishery bycatch, and were otherwise appended onto the 1916-2018 landings and apportioned among fleets using the same criteria as those documented in the 2015 benchmark assessment @hicks_status_2015. 

### Fishery catch-per-unit-effort

Three fishery-dependent CPUE indices were included as in the most recent update assessment [@adams_status_2019]. Indices were derived from 1) Oregon bottom trawl (1984-1999), 2) hake at-sea foreign (1977-1988), and 3) hake at-sea domestic fleets (1983-1998). These were not updated; please refer to the most recent benchmark assessment [@hicks_status_2015] and most recent update assessment [@adams_status_2019] for information.

### Fishery length and age data

Biological data from commercial fisheries that caught widow rockfish were extracted from PacFIN on March 25, 2025 and from the NORPAC database on February 3, 2025. Lengths and age samples taken during port sampling in California, Oregon, and Washington were used to generate length and age compositions. The data were classified into bottom trawl, midwater trawl, and hake trawl. For each fleet, the raw observations were expanded to the trip level, to account for differences in samples sizes relative to catch weights among trips (first stage expansion). The expanded length observations were then further expanded to state level, to account for differences in sampling intensity of widow rockfish landings among states combined into a single fleet (second stage expansion).

@tbl-num-land-length shows the number of trips sampled and @tbl-len-gear-state shows the number of lengths taken for each year and fleet for non-hake fleets from the three states.  @tbl-landings-hake shows these numbers for the shoreside and at-sea hake fisheries. Expanded length compositions for bottom trawl, midwater trawl and hake fleets are shown in @fig-lencomp-fleet1 to @fig-lencomp-fleet3. @tbl-sample-age-gear-state and @tbl-age-gear-state show the number of trips and number ages used in the. Age compositions for the three fleets are shown in @fig-agecomp-fleet1 and @fig-agecomp-fleet3. Occasional cohorts appear to move through the population, indicating that widow rockfish population dynamics may be characterized by episodic recruitment events.

### Discards

```{r}
#| include: false

disc<- read.csv(here::here("report", "tables", "discards.csv"))[["Discards"]]

disc_min <- format(min(disc), big.mark = ",", scientific = FALSE)
disc_max <- format(max(disc), big.mark = ",", scientific = FALSE)

```

Data on discards of widow rockfish are available from three different sources. Historical sources included @pikitch_evaluation_1988 and Enhanced Data Collection Project (EDCP, @sampson2002). The West Coast Groundfish Observer Program (WCGOP) provided information on recent discard of widow rockfish between 2002--2023. Since 2011, under trawl rationalization, 100% observer coverage is required for the limited entry trawl sectors, which resulted in a large increase in data and ability to determine discard behavior.

For this assessment, we estimated bottom trawl discards and midwater trawl discards externally, using historical sources as well as WCGOP data, and added calculated discard amounts to landings of the corresponding fleets by year. The methods used to reconstruct discard amounts in different time periods are detailed below. 

1. For the years 1982--1989, landings in the bottom trawl and midwater trawl fleets were scaled using fleet-specific discard rates estimated from the Pikitch study.
2. For the years 1990--2001, landings in the bottom trawl and midwater trawl fleets were scaled using fleet-specific discard rates estimated from the EDCP data.
3. For the years 2002--2023, GEMM estimates of fleet-specific discards amounts were added to the landed catch (bottom trawl + other, and midwater trawl).
4. The 2024 values were set to the average from the GEMM estimates for the period 2021--2023.

The treatment of the discard changed from the 2015 benchmark and 2019 update assessments. Then, discards were estimated with retention functions within the model, which was standard practice in 2015, following a long period of large discards due to trip limits. However, discard rates have been low for widow rockfish for more than 10 years so the benefit of using retention functions to extrapolate beyond the range of observed data is reduced and the biological data from the discards are no longer as likely to be informative about dynamics of the total population. As a part of model bridging, we illustrated that treating discard as catch (rather than estimating them in the model) did not have much impact on model results. Also, this approach avoids hard-to-diagnose issues that come from estimating retention curves (especially with limited amounts of data). For some species, the discard data may provide important information about recent recruitment because the smaller fish are being discarded. However, for this stock, removing the discard amounts and length-compositions from the likelihood had little impact on recruitment estimates.

Widow rockfish bycatch within the hake fishery represents total catch, with discards already accounted for.

### Biological data

The approach to the estimation of all biological parameters was the same as in the 2019 update assessment, @adams_status_2019.

#### Weight-length relationship 

Weight-at-length data were updated for this assessment. Following the approach used in the 2015 benchmark, data used to estimate the length-weight relationship for widow rockfish were gathered from commercial catch sampling schemes (PACFIN, ASHOP) and fishery-independent surveys (\gls{s-tri} and \gls{s-wcgbt}). 

The following relationships between weight and length for females and males were used in the current assessment:

```{r}
#| include: false

library(dplyr)
lw_pars  <- read.csv(here::here("data_derived", "length_weight", "weight_length_estimates.csv"))

f_lw <- c(
  round(lw_pars|>filter(Source == "All", sex == "female")|>pull(A)*100000,4),
  round(lw_pars|>filter(Source == "All", sex == "female")|>pull(B),4)
)

m_lw <- c(
  round(lw_pars|>filter(Source == "All", sex == "male")|>pull(A)*100000,4),
  round(lw_pars|>filter(Source == "All", sex == "male")|>pull(B),4)
)

```


$$
\text{Females: } \qquad weight = `r f_lw[[1]]` \times 10^{-5} \cdot length^{`r f_lw[[2]]`}
$$

$$
\text{Males: } \qquad weight = `r m_lw[[1]]` \times 10^{-5} \cdot length^{`r m_lw[[2]]`}
$$

where weight is measured in kilograms and length in cm. These parameters were used in the assessment as fixed.


#### Maturity schedule

Maturity parameters in this update assessment were carried over from 2015 benchmark and 2019 update assessments; please refer to the most recent benchmark assessment [@hicks_status_2015] and most recent update assessment [@adams_status_2019] for information.

#### Fecundity

The model uses widow rockfish specific fecundity-at-length relationships from the meta-analysis from @dick_meta-analysis_2017 instead of assuming spawning output being proportional to body weight as in 2015 and 2019 models. Accounting for size-dependent fecundity in rockfish stocks using the meta-analysis from @dick_meta-analysis_2017 has become standard practice in recent years and reflects the best available science. The fecundity parameters for widow rockfish are reported at https://github.com/EJDick-NOAA/Rockfish-Fecundity as follows:

a = 1.10961E-08, b = 4.545

The value for a above produces spawning output values in billions of eggs from SS3, given length in cm.

#### Natural Mortality

In this assessment, natural mortality (M) is estimated for females and males, while using the @hamel_development_2022 prior. The prior on M has been updated to reflect the most recent guidance from @hamel_development_2022; the log-mean therefore remains unchanged while the log-SD has been set to 0.31. Using a maximum age of 54 the point estimate and median of the prior on M is 0.10.

#### Length-at-age

Growth parameters were fully estimated within the assessment model, for females and males separately following the same formulation in 2015 and 2019 assessment models.

#### Ageing bias and imprecision

Ageing error matrices were unchanged from the previous assessment; please refer to the most recent benchmark assessment [@hicks_status_2015] and most recent update assessment [@adams_status_2019] for information.

## Environmental and Ecosystem Data

This assessment did not use any environmental or ecosystem data related to the stock.
