
```{r}
#| include: false

library(r4ss)

if (!exists("report")) {
  report <- r4ss::SS_output(here::here("models","2025 base model"))
}
```

# Assessment model

## History of modeling approaches

This section is not required for an update assessment; please refer to the most recent benchmark assessment [@hicks_status_2015] and most recent update assessment [@adams_status_2019] for information.

## Responses to most recent past STAR Panel recommendations

There are no recommendations from the most recent STAR Panel to address. 

## Responses to SSC Groundfish Subcommittee requests

There are no requests from the SSC Groundfish Subcommittee. 

## Model Structure and Assumptions

### Model Changes from the Last Assessment

The changes made to the 2019 update assessment [@adams_status_2019] occurred in two stages. A first round of changes were reviewed in August 2025. The model reviewed at the August 2025 Pacific Fishery Management Council (PFMC) Science and Statistical Committee (SSC) Groundfish Subcommittee (GFSC) meeting followed the 2015 benchmark assessment in its structure and assumptions, as required by the Terms of Reference for update assessments (with the exception of removing discard length composition data for the hook-and-line fleet, as discussed and approved at the review meeting).

A second round of changes followed after the August 2025 review meeting, when we explored a larger range of options which go beyond update assessment Terms of Reference and no longer match the 2015 benchmark. The second round of changes resulted in more parsimonious model that was raised to the current best practices in West Coast groundfish stock assessments.

The changes made to the model are listed in two separate groups, according to stages described above.

**Changes made from the 2019 update assessment for August 2025 update:**

- **Adding most recent data from commercial fleets** (as described in "Data" section of this report).
- **Updating \gls{s-wcgbt} index using current methodology** (as described in "Data" section of this report).
- **Updating the prior for natural mortality** based on @hamel_development_2022.
- **Updating length-weight parameters estimated** by including most recent data.
- **Extending the main period for estimating recruitment deviations and updating recruitment bias adjustment parameters** based on @methot_adjusting_2011.
- **Adding a block to the retention curve for the midwater trawl fleet and a block to hake fleet selectivity to account for recent changes in fleet behavior.** Adding a block to the retention curve for midwater trawl allowed for recent increases in midwater trawl discards (from 2017 forward), and improved fit to discard amounts modeled in the assessment. Adding a block to hake fleet selectivity accounted for a change in fish mean length (2020--2024) due to shifts in the spatial distribution of the hake fleet and to improve fit to length compositions [@holland_bycatch_2019]. Neither change had a discernible influence on the estimated SSB. See the "Model Bridging" section for additional detail.
- **Adding hook-and-line discards to landings in the hook-and-line fleet.** WCGOP corrected discard lengths for the hook-and-line fleet to omit nearshore fixed gear samples, which were erroneously included with the hook-and-line fleet in the previous benchmark [@hicks_status_2015] and update assessments [@adams_status_2019]. This resulted in changes to the discard length distribution and years for which data was available [@fig-disc-hnl-comparison]. Because of decreased sample sizes of hook-and-line discard length data available, this update assessment model was unable to reliably estimate retention parameters and exhibited substantial sensitivity to even slight changes in discard amounts within the hook-and-line fleet. Therefore, in this assessment, we added the hook-and-line discard to hook-and-line landings. Removal of the hook-and-line discard lengths translated into lower estimated recruitment in the 2010s than that of in 2019 update assessment (higher recruitment estimates in the 2019 assessment were previously informed by erroneous estimates of smaller fish, which were corrected in this assessment by combining the hock-and-line discards and landings). Lower recruitment in turn contributed to a decrease in stock size in recent years.


**Changes made from August 2025 model for the January 2026 model:**

- **Streamlining the fleet structure** (as described in "Data" section of this report). The landings from fleets with minimal and mostly historical catches (hook-and-line and net) were added to the bottom trawl landings with hook-and-line and net composition data removed from the model.
- **Use external estimates of discards rather than estimate within the model**. Methods, used to reconstruct discards in the bottom trawl and midwater trawl fleets, based on the Pikitch, EDCP and WCGOP data are described in "Data" section of this report. All retention parameters were removed from the model.
- **Updating the data weighting approach.** The data weighting approach was changed to match recent practices, including the use of the Francis rather than McAllister-Ianelli approach, and no longer including a 50% additional reduction in weights. The Francis method performed better in a simulation analysis [@punt_some_2017] and has been the preferred approach in most recent stock assessments. The 50% multiplier on the data weights used in the previous assessment was an approach more common in 2015 to account for the potential double use of data since length and age are observed from the same fish. However, the sum of the adjusted input sample sizes for length and age data combined are already far lower than the number of sampled fish because they are primarily based on trips or hauls and then further reduced by the data-weighting algorithm. This 50% adjustment is no longer common practice and has not been used any other recent assessments.
- **Use a fecundity relationship rather than assume the number of eggs is proportional to body weight.** Use fecunsity-at-length based on the meta-analysis from @dick_meta-analysis_2017 has become standard practice in recent years and reflects the best available science.
- **Various additional refinements made.** 
  - The landings from Washington were updated to use the new Washington catch reconstruction.
  - The historical catch within foreign fishery that targeted Pacific Ocean Perch during 1966--1976 period was moved from the hake fleet to the bottom trawl fleet. 
  - The bias adjustment ramp was updated after the retuning and also to modify the estimated values to have full bias adjustment starting in 1970.
  - The start of the final block on midwater trawl selectivity was moved from 2011 to 2017.

Comparison of the time series estimates from the 2019 update assessment model, August 2025 model and this model are shown in @fig-base_19_25_bratio to @fig-base_19_25_recdevs. See "Model Bridging" section for further details.

#### Model Bridging up to August 2025

The model bridging took place in two stages: first from the 2019 update assessment to the model presented in August 2025 to the Groundfish Subcommittee (GFSC) of the Pacific Fishery Management Council's Scientific and Statistical Committee (described below), and second after the GFSC recommended additional explorations that were outside the terms of reference for an update assessment (descibed in the following section).

The exploration of models began by bridging from the 2019 update assessment to SS3 version 3.30.23.2, which produced no discernible difference. We then focused on data bridging (@fig-bridge2019-data). Updating the catch series did not have a substantial effect on the historical biomass; the stock biomass increasing from 2000--2020 before decreasing up to the current period. 

Updates to discards, particularly the removal of hook and line discard amounts and length composition data, had the most significant impact on the absolute stock biomass (SSB): absolute SSB before 1980 was slightly lower than the 2019 update assessment estimates, and noticeably lower from the mid 2010s through the current year. Updates to model indices and composition data, following updates to discards, lead to minor changes in SSB estimates. Removal of hook-and-line discard data had the largest effect on recent relative SSB (fraction of unfished biomass) in recent years; updates to all other data sets, following updates to discards, lead to minor changes in recent relative SSB estimates (@fig-bridge2019-bratio). 

Additional model bridging changes included (1) updating the prior for natural mortality (M) to follow that recommended by @hamel_development_2022, (2) updating fixed parameters of the male and female length-weight curve by fitting to data from the \gls{s-wcgbt}, ASHOP, and the \gls{s-tri} outside the model, (3) re-fitting the bias adjustment ramp for recruitment deviates, (4) adding a block on retention in the midwater trawl fishery from 2011--2016 and allowing estimation of retention in the final years of the model, which was previously fixed at 0.99, (5) adding a block to hake fleet selectivity from 2020--2024, and (6) re-fitting the model using the MLE as initial values following a jittering analysis, which revealed the previous MLE was a local minimum. These changes collectively resulted in a very small decrease in relative SSB post-1990, and pairwise correlations among all shared parameters in these models were high (>0.999). Bridging from the previous prior on M to the @hamel_development_2022 prior is detailed in the “Priors” section. None of these model bridging steps had a substantial effect on the estimates of stock biomass (@fig-bridge2019-data). 

To estimate discards in the model, time blocks for changes in selectivity and retention in discard data were used. Except for the aforementioned addition of blocks to the midwater trawl and hake fleets, the same structure for time blocks was used as in the most recent update assessment [@adams_status_2019]. The addition of midwater trawl and hake fleets block was necessitated by poor fits to midwater trawl discards and the length composition data from the hake fleet in from 2020--2024, respectively. The poor fit to hake fleet length composition data may be due to the high variability in the distribution of the hake fleet between years, as the fleet moves to avoid bycatch [@holland_bycatch_2019]. Poor fit to midwater trawl discards is thought to be due to (1) the previous decision to fix late-year retention to 0.99 which represented an overestimate of the (very low) discards in 2011--2016, and (2) the reallocation of quota shares and subsequent increase in midwater trawl discards in 2017. These choices are described in more detail in the “Fits to Data” section.

#### Model Bridging after August 2025 Review

```{r}
#| include: false

# read ss_summary.sso files (much quicker than running r4ss::SS_output())
summary_Aug2025_base <- r4ss::SS_read_summary(here::here("models/Aug2025 base model/ss_summary.sso"))
summary_1.04 <- r4ss::SS_read_summary(here::here("models/supplemental_requests/1.04_add_fecundity/ss_summary.sso"))

get_min_bratio <- function(ss_summary) {
  bratiovals <- ss_summary$derived_quants |>
    dplyr::filter(grepl("Bratio_", rownames(ss_summary$derived_quants)))
  bratio_min <- bratiovals$Value |> 
    min()
  bratio_min_yr <- bratiovals |>
    dplyr::filter(Value == bratio_min) |>
    rownames() |>
    substring(8) |>
    as.numeric()
  glue::glue("{round(bratio_min, 3)} in {bratio_min_yr}")
}
```

For illustration of the additional bridging analysis after the August 2025 review, the changes to the August 2025 base model are grouped into the following steps:

1. Adopt the new Washington historical catch reconstruction
1. Simplify fleets and include discards with landings, refine selectivity block
1. Update the status-quo data weighting (McAllister-Ianelli, lambda = 0.5)
1. Change data weighting methods (Francis, lambda = 1)
1. Add the fecundity relationship for rockfish in general (unobserved sebastes)
1. Refine the recruitment bias adjustment and re-tune the weighting
1. Update the fecundity relationship with the species-specific relationship for widow rockfish

**TODO: REVISE TEXT BELOW WHICH WAS COPIED FROM THE OCTOBER 2025 REPORT: "Proposed new widow rockfish base model"**

None of the updates resulted in large changes in estimated population trajectories and model outputs, but several important quantities were sensitive to a few of the changes (TODO: add tables) 
<!-- (@tbl-bridge2, Figures -@fig-bridge2_bratio and -@fig-bridge2_recruits). -->

Adding the Washington historical catch reconstruction had a relatively small impact, as was already demonstrated through a sensitivity analysis in the August 2025 assessment report. 

Simplifying the fleet structure and combining discards with landings did not bring about large changes in model results either, but interacted with the Francis weighting approach applied later. Extending the block on midwater trawl selectivity did not cause any noticeable change.

The use of the Francis weighting approach (instead of McAllister-Ianelli) and using lambda = 1 (and not further down-weighting fishery composition data) resulted in the biggest change from the August 2025 base model, causing the estimates of natural mortality to go up, which translated into increases in the estimated productivity of the stock, including equilibrium catch and estimated OFLs. Relative to the August 2025 model, the updated data weighting method upweighted length data from the bottom trawl fleet and age data from all three commercial fleets (bottom trawl, midwater trawl, hake); all other composition data sources were downweighted (TODO: add table)<!-- (@tbl-compweight-comparison) -->. For bottom trawl, in particular, this upweighting is likely in part due to the removal of discard composition data, as Stock Synthesis does not apply separate data weights to discarded and retained composition data from the same fleet. The discard compositions are very noisy and difficult to fit, so their removal allowed for upweighting of the remaining data. The decrease in the weight of West Coast Groundfish Bottom Trawl Survey (WCGBTS) conditional age-at-length data by approximately one-third, paired with upweighting of other data, is likely the major driver of the increased estimate of natural mortality, given explorations in GFSC request 6. 

With the change in fecundity relationship, spawning output is now expressed in trillions of eggs and no longer directly comparable with spawning biomass expressed in metric tons in previous models. This change did not have much impact on age 4+ biomass, but the spawning output time series was impacted by the changes in relative contributions of older fish. In particular, the stock status (fraction of unfished spawning output) reached a low point of `r get_min_bratio(summary_1.04)` (below the 0.25 overfished threshold) as compared to a minimum of `r get_min_bratio(summary_Aug2025_base)` in the August 2025 base model. The addition of the fecundity relationship also resulted in slightly lower estimates of OFLs and fraction unfished at the end of the time series compared to the previous step in this bridging, though still higher than the August 2025 estimates.

After the change in fecundity relationships, an additional refinement to the recruitment bias adjustment was conducted and the model re-tuned (again with the Francis method) which had only a small impact on the model results.

Finally, an error in the choice of fecundity relationship was identified by the October 2025 review panel and replaced by the widow-rockfish-specific fecundity relationships. Due to an oversight, the October 2025 base model used the fecundity parameters associated with an unobserved rockfish species, rather than the relationship estimated for widow rockfish. The widow-specific parameters used were taken from https://github.com/EJDick-NOAA/Rockfish-Fecundity with an additional division by 1000 for the scalar parameter to convert the resulting spawning output from billions to trillions of eggs. The final parameters were 

### Modeling Platform and Structure

For this update assessment, new versions of the previously used software were used. The 2019 update assessment used Stock Synthesis v3.30.13 to estimate the model; and R4SS version 1.35.3 and R version 3.5.3 to investigate and plot the 2019 model fits. This update used Stock Synthesis v3.30.24.1 to estimate the model; and R4SS version 1.52.1 and R version 4.5.0 to investigate and plot the model fits. A summary of the data sources used in the model (details discussed above) is shown in @fig-data-plot-base. 

### Model Overview

The model is a two-sex, age-structured model starting in 1916 with an accumulated age group at 40 years. Sex-specific growth and natural mortality were estimated. The lengths in the population were tracked by 1 cm intervals and the length data were binned into 2 cm intervals. Ageing error was retained from the 2015 benchmark assessment [@hicks_status_2015]. TODO: UPDATE: Fecundity was assumed to be proportional to body weight, thus \gls{ssb} was used as the measure of spawning output.

#### Model Fleets and Areas

The assessment uses a single-area model, consistent with the previous benchmark and update assessments [@hicks_status_2015, @adams_status_2019]. Multiple fisheries encounter widow rockfish. The fishing fleets have been simplified from previous benchmark and update assessments [@hicks_status_2015, @adams_status_2019], where two fleets with small catches were combined with the bottom trawl fleet. These fleets, hook-and-line and net, represented 1.1% and 1.4% of the input catches, respectively. 

### Model Parameters 

#### Estimated and Fixed Parameters

There were `report$parameters$Active_Cnt |> max(na.rm = TRUE) # was 194 on 2025-09-18` estimated parameters in the base model (relative to 214 in the August 2025 base model). These included one parameter for recruitment (R0), 10 sex-specific parameters for sex-specific growth, two sex-specific natural mortality parameters, four parameters for the catchability (q) of the hake and the \gls{s-tri} of abundance (catchability for other indices were calculated analytically using the "float" option in SS3), four parameters each for extra variability for the hake, bottom trawl, juvenile and foreign-at-sea indices of abundance, 36 parameters for selectivity, 125 recruitment deviations (16 of which set up the initial age structure), and 12 forecast recruitment deviations.

Fixed parameters in the model included steepness, standard deviation of recruitment deviates, maturity at age, and length-weight parameters Consistent with the previous full assessment [@adams_status_2019], steepness was fixed at 0.72, matching the mean of the current west coast rockfish steepness prior [@thorson_steep_2019], described below in the "Priors" section. A likelihood profile were done for steepness. The standard deviation of recruitment deviates was fixed at 0.60. Maturity at age was fixed as described in the previous update assessment, @adams_status_2019. Length-weight parameters were fixed at estimates using length-weight observations from the \gls{s-wcgbt} (@fig-wl-fits and @tbl-params).

All selectivity curves in the final base model were length based and the same shape as the 2019 update, with the exception of the \gls{s-wcgbt} survey fleet. The final base model assumed asymptotic selectivity (using the double-normal formulation in SS3) for each fishery, except for the midwater trawl fishery, following the 2015 benchmark assessment [@hicks_status_2015]. The \gls{s-wcgbt} and \gls{s-tri} both used spline curves. The \gls{s-wcgbt} survey fleet selectivity was estimated to be slightly lower at lengths greater than 45 cm, compared to the selectivity estimated in 2019.

Time blocks were used for selectivity and retention parameters in the bottom trawl, midwater trawl, and hake fisheries as indicated in @tbl-speci-base-assess.

#### Priors

The prior on natural mortality (M) in the previous assessment was defined as a lognormal with mean on the log-scale of ln(5.4/Amax) and SD(ln(M)) = 0.44 following analysis of the data in @then_evaluating_2015 by Owen Hamel. In the current assessment the prior on M has been updated to reflect guidance from @hamel_development_2022; the log-mean therefore remains unchanged while the log-SD has been set to 0.31. Using a maximum age of 54 the point estimate and median of the prior on M is 0.10.

The prior for steepness (h) assumes a beta distribution with parameters based on an update of the Dorn rockfish prior (commonly used in past West Coast rockfish assessments) conducted by Jim Thorson (pers. comm, NWFSC, NOAA), which was reviewed and endorsed by the SSC in 2015. During the stock assessment review in 2015, it was decided that the steepness prior should be developed without the past widow rockfish data to avoid using the same data to inform both the model prior and likelihood. Without widow rockfish, the prior used for the 2015 assessment was a beta distribution with mean = 0.798 and SD = 0.132 (corresponding beta parameters $\alpha$ = 6.59 and $\beta$ = 1.667). The 2019 update assessments used the current West Coast rockfish steepness prior with mean = 0.72 and SD = 0.16 ($\alpha$ = 4.95, $\beta$ = 1.93) which was approved for use in all rockfish stock assessments for 2019. This update assessment uses the same prior as the 2019 update assessment [@adams_status_2019]. 

#### Recruitment deviations

Recruitment deviations from 1900--2024 were estimated to appropriately quantify uncertainty; inappropriate specifications can have a large effect on model uncertainty. The standard deviation of recruitment variability (sigma-R) was assumed to be 0.6 in the 2015 assessment, based on iteratively tuning to a value slightly less than the observed variability of recruitment deviations in the period 1975--2010 in 2015 (@fig-recfit-bias). The earliest length-composition data occur in 1976 and the earliest age data were in 1978. The most informed years for estimating recruitment deviations based on available composition data were from about the mid-1970s to about 2019. The period from 1900--1970 was fit using an early series with little or no bias adjustment, the main period of recruitment deviates was 1971--2020 using a bias adjustment ramp, and 2021 onward was fit using forecast recruitment deviates with little bias adjustment. @methot_adjusting_2011 summarize the reasoning behind varying levels of bias adjustment based on the information available to estimate the deviates. The bias adjustment settings were tuned to the estimates produced by `r4ss::SS_fitbiasramp()` with the exception that the first year with full bias adjustment was set to 1970 (instead of the estimate of 1975, which occurs after two large, reasonably well-informed recruitments in 1970 and 1971). 

#### Sample weights

The base model uses the @francis_data_2011 weighting method. This is a departure from the August 2025 base model which followed the most recent benchmark assessment [@hicks_status_2015] in using the @mcallister_bayesian_1997 method. An additional change from the previous assessments was the removal of the lambda factors for all fleets with both marginal length and marginal age compositions which down-weighted the likelihood contribution by 50% to account for the potential double use of data since length and age are observed from the same fish. The adjusted input sample sizes which are based on trips or hauls rather and then further reduced by the data-weigthing method, are far lower than the number of number of sampled fish, and this 50% adjustment has not been used in other recent assessments. 

### Key Assumptions and Structural Choices

**TODO: add text here to reflect further changes from August to new base.**

This section is not required for an update assessment; please refer to the most recent full assessment [@hicks_status_2015] for additional information.
