
```{r}
#| include: false

library(r4ss)

if (!exists("report")) {
  report <- r4ss::SS_output(here::here("models","2025 base model"))
}
```

# Assessment model

## History of modeling approaches

This section is not required for an update assessment; please refer to the most recent benchmark assessment [@hicks_status_2015] and most recent update assessment [@adams_status_2019] for information.

## Responses to most recent past STAR Panel recommendations

There are no recommendations from the most recent STAR Panel to address. 

## Responses to SSC Groundfish Subcommittee requests

Responses to the most recent requests from the Science and Statistical Committee (SSC) and Pacific Fishery Management Council (PFMC) are provided in a separate document. 

## Model Structure and Assumptions

### Changes from the Last Assessment and Model Bridging

The last benchmark stock assessment for widow rockfish was conducted in 2015 [@hicks_status_2015], and it was updated in 2019 [@adams_status_2019]. 

In this assessment, the changes made to the 2019 model [@adams_status_2019] occurred in two stages. A first round of changes were reviewed in August 2025. The model reviewed at the August 2025 Pacific Fishery Management Council (PFMC) Science and Statistical Committee (SSC) Groundfish Subcommittee (GFSC) meeting followed the 2015 benchmark assessment in its structure and assumptions, as required by the Terms of Reference for update assessments (with the exception of removing discard length composition data for the hook-and-line fleet, as discussed and approved at the review meeting).

A second round of changes followed after the August 2025 review meeting, when we explored a larger range of options which go beyond update assessment Terms of Reference and no longer match the 2015 benchmark. The second round of changes resulted in more parsimonious model that was raised to the current best practices in West Coast groundfish stock assessments.

The changes made to the model are listed in two separate groups, according to stages described above.

**Changes made from the 2019 update assessment for August 2025 update:**

The exploration of models began by bridging from the 2019 update assessment to current SS3 version, which produced no discernible difference. 

We then focused on data bridging (@fig-bridge2019-data). 

- **Adding most recent data from commercial fleets** (as described in "Data" section of this report). 
- **Updating \gls{s-wcgbt} index using current methodology** (as described in "Data" section of this report).
- **Updating the log-SD of the prior for natural mortality** based on @hamel_development_2022. The log-mean therefore remains unchanged while the log-SD has been decreased 0.44 to 0.31.
- **Updating length-weight parameters estimated** by including most recent data.
- **Extending the main period for estimating recruitment deviations and updating recruitment bias adjustment parameters** based on @methot_adjusting_2011.
- **Adding a block to hake fleet selectivity to account for recent changes in fleet behavior.** Adding a block to hake fleet selectivity accounted for a change in fish mean length (2020--2024) due to shifts in the spatial distribution of the hake fleet and to improve fit to length compositions [@holland_bycatch_2019]. This change had a discernible influence on the estimated spawning biomass. 
- **Adding hook-and-line discards to landings in the hook-and-line fleet.** WCGOP corrected discard lengths for the hook-and-line fleet to omit nearshore fixed gear samples, which were erroneously included with the hook-and-line fleet in the previous benchmark [@hicks_status_2015] and update assessments [@adams_status_2019]. This resulted in changes to the discard length distribution and years for which data was available [@fig-disc-hnl-comparison]. Because of decreased sample sizes of hook-and-line discard length data available, this assessment model was unable to reliably estimate retention parameters and exhibited substantial sensitivity to even slight changes in discard amounts within the hook-and-line fleet. Therefore, in this assessment, we added the hook-and-line discard to hook-and-line landings. 

Updating the catch series did not have a substantial effect on the historical biomass; the stock biomass increasing from 2000--2020 before decreasing up to the current period.

Updates to discards, particularly the removal of hook-and-line length composition data and adding hook-and-line discards to landings, had the most significant impact on the absolute stock biomass (SSB): absolute SSB before 1980 was slightly lower than the 2019 update assessment estimates, and noticeably lower from the mid 2010s through the current year. In the 2015 and 2019 assessments, the hook-and-line discard length data consisted primarily of small fish from nearshore fixed gear, and the model interpreted the presence of those small fish as a recruitment signal. With the erroneous data removed and the hook-and-line discards added to landings, the model estimated lower recruitment in the 2010s, which is translated into lower estimates of spawning biomass and stock productivity in recent years.
 
Updates to model indices and composition data, following updates to discards, lead to minor changes in SSB estimates. The rest of the changes lead to minor changes in recent relative SSB estimates (@fig-bridge2019-bratio) and collectively resulted in a very small decrease in relative SSB post-1990, and pairwise correlations among all shared parameters in these models were high (>0.999). None of these model bridging steps had a substantial effect on the estimates of stock biomass (@fig-bridge2019-data).

**Changes made from August 2025 model for the January 2026 model:**

After August 2025 review, we explored changes to the model that were outside the Terms of Reference for an update assessment. Those changes included:

- **Streamlining the fleet structure** (as described in "Data" section of this report). The landings from fleets with minimal and mostly historical catches (hook-and-line and net) were added to the bottom trawl landings with hook-and-line and net composition data removed from the model.
- **Use external estimates of discards rather than estimate within the model**. Methods, used to reconstruct discards in the bottom trawl and midwater trawl fleets, based on the Pikitch, EDCP and WCGOP data are described in "Data" section of this report. All retention parameters were removed from the model.
- **Updating the data weighting approach.** The data weighting approach was changed to match recent practices, including the use of the Francis rather than McAllister-Ianelli approach, and no longer including a 50% additional reduction in weights. The Francis method performed better in a simulation analysis [@punt_some_2017] and has been the preferred approach in most recent stock assessments. The 50% multiplier on the data weights used in the previous assessment was an approach more common in 2015 to account for the potential double use of data since length and age are observed from the same fish. However, the sum of the adjusted input sample sizes for length and age data combined are already far lower than the number of sampled fish because they are primarily based on trips or hauls and then further reduced by the data-weighting algorithm. This 50% adjustment is no longer common practice and has not been used any other recent assessments.
- **Use a fecundity relationship rather than assume the number of eggs is proportional to body weight.** Use fecundity-at-length based on the meta-analysis from @dick_meta-analysis_2017 has become standard practice in recent years and reflects the best available science.
- **Various additional refinements made.** 
  - The landings from Washington were updated to use the new Washington catch reconstruction.
  - The historical catch within foreign fishery that targeted Pacific Ocean Perch during 1966--1976 period was moved from the hake fleet to the bottom trawl fleet. 
  - The bias adjustment ramp was updated after the retuning and also to modify the estimated values to have full bias adjustment starting in 1970.
  - The start of the final block on midwater trawl selectivity was moved from 2011 to 2017.



```{r}
#| include: false

# read ss_summary.sso files (much quicker than running r4ss::SS_output())
summary_Aug2025_base <- r4ss::SS_read_summary(here::here("models/Aug2025 base model/ss_summary.sso"))
summary_1.04 <- r4ss::SS_read_summary(here::here("models/supplemental_requests/1.04_add_fecundity/ss_summary.sso"))

get_min_bratio <- function(ss_summary) {
  bratiovals <- ss_summary$derived_quants |>
    dplyr::filter(grepl("Bratio_", rownames(ss_summary$derived_quants)))
  bratio_min <- bratiovals$Value |> 
    min()
  bratio_min_yr <- bratiovals |>
    dplyr::filter(Value == bratio_min) |>
    rownames() |>
    substring(8) |>
    as.numeric()
  glue::glue("{round(bratio_min, 3)} in {bratio_min_yr}")
}
```

Stepwise bridging from the August 2025 model to January 2025 included the following:

1. Adopt the new Washington historical catch reconstruction
1. Simplify fleets and include discards with landings, refine selectivity block
1. Update the status-quo data weighting (McAllister-Ianelli, lambda = 0.5)
1. Change data weighting methods (Francis, lambda = 1)
1. Add the fecundity relationship for rockfish in general (unobserved sebastes)
1. Refine the recruitment bias adjustment and re-tune the weighting
1. Update the fecundity relationship with the species-specific relationship for widow rockfish

None of the updates resulted in large changes in estimated population trajectories and model outputs, but several important quantities were sensitive to a few of the changes (TODO: add tables and figures) 
<!-- (@tbl-bridge2, Figures -@fig-bridge2_bratio and -@fig-bridge2_recruits). -->

Adding the Washington historical catch reconstruction had a minimal impact. Simplifying the fleet structure and combining discards with landings did not reveal unexpected or unreasonable patterns with a disproportionate effect of changes on model results (as it was in case of HKL discard data). Simplifying the fleet structure also did not bring about large changes in model results either, but interacted with the Francis weighting approach applied later. Extending the block on midwater trawl selectivity did not cause any noticeable change.

The use of the Francis weighting approach (instead of McAllister-Ianelli) and using lambda = 1 (and not further down-weighting fishery composition data) resulted in the biggest change from the August 2025 base model, causing the estimates of natural mortality to go up, which translated into increases in the estimated productivity of the stock, including equilibrium catch and estimated OFLs. Relative to the August 2025 model, the updated data weighting method upweighted length data from the bottom trawl fleet and age data from all three commercial fleets (bottom trawl, midwater trawl, hake); all other composition data sources were downweighted (TODO: add table)<!-- (@tbl-compweight-comparison) -->. For bottom trawl, in particular, this upweighting is likely in part due to the removal of discard composition data, as Stock Synthesis does not apply separate data weights to discarded and retained composition data from the same fleet. The discard compositions are very noisy and difficult to fit, so their removal allowed for upweighting of the remaining data. The decrease in the weight of West Coast Groundfish Bottom Trawl Survey (WCGBTS) conditional age-at-length data by approximately one-third, paired with upweighting of other data, is likely the major driver of the increased estimate of natural mortality, given explorations in GFSC request 6. 

With the change in fecundity relationship, spawning output is now expressed in billions of eggs and no longer directly comparable with spawning biomass expressed in metric tons in previous models. This change did not have much impact on age 4+ biomass, but the spawning output time series was impacted by the changes in relative contributions of larger/older fish which are disproportionately removed from the population, which leads to lower estimates of fraction of unfished and estimated OFLs. In particular, the stock status (fraction of unfished spawning output) reached a low point of `r get_min_bratio(summary_1.04)` (below the 0.25 overfished threshold) as compared to a minimum of `r get_min_bratio(summary_Aug2025_base)` in the August 2025 base model. The addition of the fecundity relationship also resulted in slightly lower estimates of OFLs and fraction unfished at the end of the time series compared to the previous step in this bridging, though still higher than the August 2025 estimates.

After the change in fecundity relationships, an additional refinement to the recruitment bias adjustment was conducted and the model re-tuned (again with the Francis method) which had only a small impact on the model results.

Comparison of the time series estimates from the 2019 update assessment model and this model are shown in @fig-base_19_25_bratio to @fig-base_19_25_recdevs. 

### Modeling Platform and Structure

Stock Synthesis statistical catch-at-age modelling framework, version 3.30.24.1, was used for this assessment; and R4SS version 1.52.1 and R version 4.5.0 to investigate and plot the model fits. A summary of the data sources used in the model (details discussed above) is shown in @fig-data-plot-base.

### Model Overview

The model is a two-sex, age-structured model starting in 1916 with an accumulated age group at 40 years. Sex-specific growth and natural mortality were estimated. The lengths in the population were tracked by 1 cm intervals and the length data were binned into 2 cm intervals. Ageing error was retained from the 2015 benchmark assessment [@hicks_status_2015]. The model uses widow rockfish fecundity at length relationships instead of assuming spawning output being proportional to body weight as in the 2015 and 2019 models. With the change in fecundity relationship, spawning output is now expressed in billions of eggs and not in metric tons, as in 2015 and 2019 models.

#### Model Fleets and Areas

The assessment uses a single-area model, consistent with the previous benchmark and update assessments [@hicks_status_2015, @adams_status_2019]. The fleet structure in this assessment was streamlined compared to the 2015 benchmark and the 2019 update assessments [@hicks_status_2015, @adams_status_2019]. 








In this assessment, the catches are divided among three fishing fleets:

1. A coastwide shore-based bottom trawl fleet (1916–2024). This fleet also includes minimal catches made by net and hook-and-line gears, which represented 1.4% and 1.1% of the input catches, respectively. Removals from net and hook-and-line fisheries were previously treated as separate fleets. The bottom trawl fleet also includes widow rockfish catches made within historical foreign fishery targeting Pacific Ocean Perch between 1966–1976. 
2. A coastwide shore-based midwater trawl fleet that targets rockfish (1979–2024).
3. A mostly midwater trawl fleet that targets hake and includes at-sea fleet (1975–2024) and a domestic shore-based fleet (1991–2024).

Catch in each fleet represents both landings and discards. Discards in bottom trawl and midwater trawl fleets were estimated externally to the model and added to landings in this assessment.

The following fishery-independent surveys are included in the model: 

1. Alaska Fisheries Science Center/Northwest Fisheries Science Center West Coast Triennial Shelf Survey (Triennial Survey). 
2. Northwest Fisheries Science Center West Coast Groundfish Bottom Trawl Survey (WCGBTS) (2003-2024).
3. recruitment index from the National Marine Fisheries Service (NMFS) SWFSC and NWFSC/PWCC Midwater Trawl Survey (Juvenile Survey) (2004-2024).

### Model Parameters 

#### Estimated and Fixed Parameters

There were `report$parameters$Active_Cnt |> max(na.rm = TRUE) # was 194 on 2025-09-18` estimated parameters in the base model (relative to 214 in the August 2025 base model). These included one parameter for recruitment (R0), 10 sex-specific parameters for sex-specific growth, two sex-specific natural mortality parameters, four parameters for the catchability (q) of the hake and the \gls{s-tri} of abundance (catchability for other indices were calculated analytically using the "float" option in SS3), four parameters each for extra variability for the hake, bottom trawl, juvenile and foreign-at-sea indices of abundance, 36 parameters for selectivity, 125 recruitment deviations (16 of which set up the initial age structure), and 12 forecast recruitment deviations.

Fixed parameters in the model included steepness, standard deviation of recruitment deviates, maturity at age, fecundity at length, and length-weight parameters. Consistent with the previous assessment [@adams_status_2019], steepness was fixed at 0.72, matching the mean of the current west coast rockfish steepness prior [@thorson_steep_2019], described below in the "Priors" section. A likelihood profile were done for steepness. The standard deviation of recruitment deviates was fixed at 0.60. Maturity at age was fixed as described in the previous update assessment, @adams_status_2019. Fecundity at length parameters were fixed at values from @dick_meta-analysis_2017. Length-weight parameters were fixed at estimates using length-weight observations from the \gls{s-wcgbt} (@fig-wl-fits and @tbl-params). 

All selectivity curves in the base model were length based and had the same shape as the 2015 benchmark and the 2019 update assessments. The base model assumed asymptotic selectivity (using the double-normal formulation in SS3) for each fishery, except for the midwater trawl fishery [@hicks_status_2015]. The \gls{s-wcgbt} and \gls{s-tri} both used spline curves. The \gls{s-wcgbt} survey fleet selectivity was estimated to be slightly lower at lengths greater than 45 cm, compared to the selectivity estimated in 2019.

Time blocks were used for selectivity and retention parameters in the bottom trawl, midwater trawl, and hake fisheries as indicated in @tbl-speci-base-assess.

#### Priors

The prior on natural mortality (M) in the previous assessment was defined as a lognormal with mean on the log-scale of ln(5.4/Amax) and SD(ln(M)) = 0.44 following @hamel_method_2015 and @then_evaluating_2015. In the current assessment, the prior on M has been updated to reflect guidance from @hamel_development_2022. With this update, the log-mean remains unchanged while the log-SD has been set to 0.31. Using a maximum age of 54 the point estimate and median of the prior on M is 0.10.

The prior for steepness (h) has not changed since the 2019 update assessment [@adams_status_2019]. It assumes a beta distribution with parameters based on @thorson_steep_2019, with mean = 0.72 and SD = 0.16 ($\alpha$ = 4.95, $\beta$ = 1.93) which was approved for use in all rockfish stock assessments.

#### Recruitment deviations

Recruitment deviations from 1900--2024 were estimated to appropriately quantify uncertainty; inappropriate specifications can have a large effect on model uncertainty. The standard deviation of recruitment variability (sigma-R) was assumed to be 0.6 in the 2015 assessment, based on iteratively tuning to a value slightly less than the observed variability of recruitment deviations in the period 1975--2010 in 2015 (@fig-recfit-bias). The earliest length-composition data occur in 1976 and the earliest age data were in 1978. The most informed years for estimating recruitment deviations based on available composition data were from about the mid-1970s to about 2019. The period from 1900--1970 was fit using an early series with little or no bias adjustment, the main period of recruitment deviates was 1971--2020 using a bias adjustment ramp, and 2021 onward was fit using forecast recruitment deviates with little bias adjustment. @methot_adjusting_2011 summarize the reasoning behind varying levels of bias adjustment based on the information available to estimate the deviates. The bias adjustment settings were tuned to the estimates produced by `r4ss::SS_fitbiasramp()` with the exception that the first year with full bias adjustment was set to 1970 (instead of the estimate of 1975, which occurs after two large, reasonably well-informed recruitments in 1970 and 1971). 

#### Sample weights

The base model uses the @francis_data_2011 weighting method. This is a departure from the August 2025 base model, which followed the most recent benchmark assessment [@hicks_status_2015] in using the @mcallister_bayesian_1997 method. The data weighting approach was changed to match recent practices, including the use of the Francis rather than McAllister-Ianelli approach. The Francis method performed better in a simulation analysis [@punt_some_2017] and has been the preferred approach in most recent stock assessments.  

An additional change from the previous assessments was the removal of a 50% additional reduction in weights (via the lambda factors) for all fleets with both marginal length and marginal age compositions, which down-weighted the likelihood contribution by 50% to account for the potential double use of data since length and age are observed from the same fish. This 50% adjustment is no longer common practice and has not been used any other recent assessments, because the adjusted input sample sizes are based on trips or hauls rather the number of sampled fish, and the sum of the adjusted input sample sizes for length and age data combined are already far lower than the number of sampled fish; no further reduction by the data-weighting algorithm is necessary.
 
### Key Assumptions and Structural Choices

Following recommendation of Pacific Fishery Management Council (PFMC) Science and Statistical Committee (SSC) Groundfish Subcommittee (GFSC) meeting in August 2025, for this update assessment we explored a larger range of options which go beyond update assessment Terms of Reference. Following this recommendation, we conducted multiple model explorations, which resulted in more parsimonious model that was raised to the current best practices in West Coast groundfish stock assessments.

The major structural choice was related to fleet structure, which was simplified to focus on major fisheries that remove more that 97% of catch. Those included bottom trawl, midwater trawl and hake fleets. The fisheries with minimal and mostly historical catches, such as hook-and line and net, were added to the bottom trawl fleet. 

In the 2015 benchmark and 2019 update assessments catches were divided among  five fleets (bottom trawl, midwater trawl, catches within fleet targeting hake, net and hook-and line). Also, the 2015 benchmark and 2019 update assessments estimated discards using retention curves for the bottom trawl, midwater trawl, and hook-and-line fleets based on discard biomass and length composition data from the \gls{wcgop}. Changes to the underlying discard length composition data for the hook-and-line fleet from \gls{wcgop} illustrated that very large recruitment events previously estimated for 2013 and 2014 were mainly driven by very small amounts of composition data for a minor fleet responsible for less than one-tenth of a percent of the catch in the last decade. This led to implausible population dynamics, so the decision was made to consolidate fleets and add minimal (and primarily historical) catches from net and hook-and-line fisheries to the shore-based bottom trawl fleet. 

Another major structural change was related to treatment of the discard. In this assessment, the discards in bottom trawl and midwater trawl fleets were estimated externally, based on historical and the \gls{wcgop} data, and added to landings within the corresponding fleets. With that, all retention parameters became obsolete and thus were removed from the model. This change was motivated by the fact that discard rates have been low for widow rockfish for more than 10 years so the benefit of using retention functions to extrapolate beyond the range of observed data is reduced and the biological data from the discards are no longer as likely to be informative about dynamics of the total population. This change also helps avoid hard-to-diagnose issues that come from estimating retention curves (especially with limited amounts of data), as we encountered with the hook-and-line fleet, easrlier in the assessment process. Removing the discard amounts and length-compositions from the likelihood had little impact on recruitment estimates.

Other key assumptions were related to updating model to the current assessment practices, and included using up-to-date catches (specifically, historical catches from Washington), data weighting approach and treatment of fecundity. The stepwise changes made to the model and their justification are provided in "Changes from the Last Assessment and Model Bridging" section of this report.


with limited amounts of data), as we encountered with the hook-and-line fleet in the August 2025 base model. For some species, the discard data may provide important information about recent recruitment because the smaller fish are being discarded. However, for this stock, removing the discard amounts and length-compositions from the likelihood had little impact on recruitment estimates (See @fig-bridge2_recruits in bridging section).