
```{r}
#| include: false

library("r4ss")

if (!exists("report")) {
  report <- r4ss::SS_output(
    here("models", "2025 base model"), 
    printstats = FALSE, 
    verbose = FALSE,
    SpawnOutputLabel = "Spawning output (billions of eggs)"
  )
}

est_params <- report$estimated_non_dev_parameters

est_params["NatM_uniform_Fem_GP_1","Value"]

est_params["NatM_uniform_Mal_GP_1","Value"]

rownames(est_params)
#conf int for natM
natm_prior_ci <- setNames(exp(qnorm(
  c(0.025, 0.975), mean = est_params["NatM_uniform_Fem_GP_1", "Prior"], 
  sd = est_params["NatM_uniform_Fem_GP_1", "Pr_SD"]
)), c("lo", "hi"))

Bratio <- report$derived_quants[report$derived_quants$Label %in% paste0("Bratio_", 1916:2036),]
Bratio$Value <- 100 * round(Bratio$Value,3)
Bratio$Year <- as.integer(gsub("Bratio_", "", Bratio$Label))

```


## Base Model Results

As a supplement to the model results figures included in this report and described below, a full set of [diagnostic plots](https://github.com/mcgoodman/widow_rockfish_2025/tree/main/figures/2025%20base%20model%20r4ss%20plots) created by the r4ss package [@taylor_2021_r4ss] is available on the assessment GitHub repository ([github.com/mcgoodman/widow_rockfish_2025](https://github.com/mcgoodman/widow_rockfish_2025/)) along with the Stock Synthesis [input files](https://github.com/mcgoodman/widow_rockfish_2025/tree/main/models/2025%20base%20model).

### Parameter Estimates

```{r}
#| include: false

age_quants <- readRDS(here("data_derived", "NWFSCCombo", "nwfsc_age_quantiles.rds"))
```

The base model parameter estimates along with asymptotic standard errors are shown in @tbl-params and the likelihood components are shown in @tbl-likecomps. Estimates of key derived parameters and asymptotic 95% confidence intervals are shown in @tbl-key-deriv-params.

The estimates of natural mortality (`r round(est_params["NatM_uniform_Fem_GP_1","Value"], 3)` yr^-1^  and `r round(est_params["NatM_uniform_Mal_GP_1","Value"], 3)` yr^-1^ for females and males, respectively). These estimates continue gradual decrease in M from those of the 2019 update assessment (Female  0.1444, Male: 0.1549; @adams_status_2019, @fig-NatMort) and the 2015 benchmark assessment (Female: 0.1572, Male: 0.1705; @hicks_status_2015). However, they were higher than suggested by the medians of the prior distributions used in this assessment and the previous assessments. Fixing M at lower values than those estimates resulted in a pattern of reduced recruitment immediately before the fishery started (@fig-like_profile_recruits). This suggests that the model is attempting to reduce the number of observations of older fish in the data. The estimates of M fall within the 95% quantile range of the prior distribution (`r round(natm_prior_ci["lo"],3)`– `r round(natm_prior_ci["hi"],3)`), and are shown in @fig-NatMort. 

Updating age composition data resulted in the largest decrease in M among the various intermediate models bridging from the 2019 assessment. This is broadly consistent with observations of older individuals in updated age composition data. For instance, the maximum age for widow rockfish in the \gls{s-wcgbt} data prior to 2019 was `r age_quants$pre_2018[["100%"]]` yrs with a 99^th^ percentile of `r age_quants$pre_2018[["99%"]]`, while data collected since include individuals up to `r age_quants$post_2018[["100%"]]` with a 99^th^ percentile of `r age_quants$post_2018[["99%"]]`. Updating that natural mortality prior resulted in a comparatively small decrease in M (@fig-NatMort).

Estimating M is difficult in stock assessments, and the estimated values may represent model misspecification instead of the actual life-history trait. However, in alternative models to the base, the estimates of M were not higher than 0.13 yr^-1^ (@tbl-quant-sens-int), except when forcing asymptotic selectivity on the midwater trawl fleet (@tbl-quant-sens-int). Uncertainty in the estimated M was also much less than the range of the prior (@fig-NatMort).

Selectivity curves were estimated for commercial and survey fleets and parameter estimates are provided in @tbl-params. The estimated selectivity are shown in @fig-est-select-all. The selectivity curves showed a strong dome-shaped selectivity for the midwater fleet, while the selectivity for the hake fleet did not support dome-shaped selectivity (@fig-est-select-all). The estimated selectivity curves for the Triennial and WCGBTS were similar to each other except that the triennial survey selected larger fish (@fig-est-select-3478). The \gls{s-wcgbt} exhibited a more pronounced dome-shaped selectivity compared to the 2019 assessment. 

In the 2015 benchmark assessment, additional survey variability (observation error added directly to each year’s input variability) for the \gls{s-tri} and \gls{s-wcgbt} was not estimated in the model because initial runs estimated the it at zero [@hicks_status_2015]. To avoid bound issues in estimation of the Hessian, their values were fixed at zero. We retained the same modelling approach for the update assessment. The additional standard deviation added to the fishery-dependent indices was quite large: `r round(est_params["Q_extraSD_BottomTrawl(1)","Value"],2)` for the bottom trawl index and `r round(est_params["Q_extraSD_ForeignAtSea(9)","Value"],2)` for the foreign at-sea hake fleet. The additional variability on the juvenile survey was the highest, at `r round(est_params["Q_extraSD_JuvSurvey(6)","Value"],2)`, giving the index very little weight in the model. 

The estimates of maximum size for both females and males (@tbl-params) did not differ substantially from @adams_status_2019. Estimates of k were slightly different in the model, but that is expected when accounting for selectivity. Estimated growth curves are shown in (@fig-sizeatage-sdcv). 

Estimates of recruitment suggest that the widow rockfish population is characterized by variable recruitment with occasional strong recruitment events and periods of low recruitment (@fig-recdev-est, @tbl-key-deriv-params). There is little information regarding recruitment prior to 1965.

### Fits to the Data

Fits to data are provided for survey abundance indices, length composition data for the fisheries and surveys, marginal age compositions for the fisheries, and conditional age-at-length observations for the \gls{s-wcgbt}. 

Survey indices and total discards were fitted assuming a lognormal likelihood. The six indices of abundance (three survey series, and three fishery indices of abundance) are shown in @fig-surv-abundest. The \gls{s-tri} treatment was consistent with @adams_status_2019. Extra standard error was estimated for all of the series except for the two survey series (@tbl-params). None of the series showed patterns in residuals, and with the large amount of error, none of the series showed serious lack of fit. The recent \gls{s-wcgbt} showed a general increase from 2003 - 2015 followed by a general decrease from 2016 - 2024, which was also estimated in the base model (@fig-surv-abundest). The model did not fit the indices variability well, particularly years with relatively low (2015, 2021, 2022) or high (2016) estimates of abundance, as might be expected for a long-lived rockfish species.

Annual fits to the length-composition data are depicted using Pearson residuals-at-length for all fleets in Figures [-@fig-lenage-pears-res-trawl]. More detailed plots of fitted and observed proportions at length are shown in Appendix A. The combined \gls{s-wcgbt} length frequencies across years displayed a weakly bimodal distribution with a valley around 37 cm, which the model approximated well (@fig-comp-lenfit-aggreg). Fits to the \gls{s-tri} length frequencies (which have not been updated in this assessment) show underestimation of 35-45 cm males and overestimation of similar length females, as in @hicks_status_2015.

Age data were fitted as marginal age compositions for the fishing fleets and as conditional age-at-length for the \gls{s-wcgbt}, which were expanded by tow and then by strata, following the 2015 benchmark assessment (@hicks_status_2015). Pearson residuals for the commercial fleets are shown in Figures [-@fig-lenage-pears-res-trawl]. Aggregating across years shows that the fit to age comps was good (@fig-combine-agefreq-all). The observed and expected age-at-length are shown in @fig-condAAL-andreplot for the \gls{s-wcgbt} observations. Plots with the residuals for individual observations showed reasonably good fits to the conditional age-at-length data from the \gls{s-wcgbt} (@fig-condAAL-resids). 

### Population Trajectory

The predicted spawning output (billions of eggs) is given in @tbl-ts-popests-basecase and plotted in @fig-spawnout-95. The predicted spawning output from the base model generally showed a slight decline over the time series until 1966 when the foreign fleet began. A short, but sharp decline occurred, followed by a steep increase due to strong recruitment. The spawning output declined rapidly with the developing domestic midwater fishery in the late 1970s and early 1980s. The stock continued to decline until 2000 when a combination of strong recruitment and low catches resulted in a quick increase from 2000 - 2018. The stock declined rapidly from 2019 until the end of the time series, due to a combination of increased catches and low recruitment. 

The 2025 spawning output relative to unfished spawning output is above the target of 40% of unfinished spawning output (`r Bratio$Value[Bratio$Year == 2025]`%), with a low of `r min(Bratio$Value, na.rm = TRUE)`% in `r Bratio$Year[which.min(Bratio$Value)]` (@fig-rel-spawn-bio). Approximate confidence intervals based on the asymptotic variance estimates show that the uncertainty in the estimated spawning output is high, especially in the early years. It should be noted that while the stock is currently above the 40% target, the target does fall within the approximate 95% confidence interval (@fig-rel-spawn-bio).

Recruitment deviations (@fig-recdev-est and discussed in Section 3.3.1) provide a more realistic portrayal of uncertainty. Estimates of recruitment appear to be episodic and characterized by periods of low recruitment, with several very large, but uncertain, estimates. The 2019 update assessment estimated the 2013 recruitment as the strongest year class over the duration of the fishery, however the current assessment does not support this. It may be worthwhile to investigate the periods of strong and weak year classes further to see if it is an artifact of the data, a consistent autocorrelation, or a result of the environment. The input bias adjustment ramp matched the estimated ramp (@fig-recfit-bias).

The stock-recruit curve resulting from a fixed value of steepness is shown in @fig-recfit-bias with estimated annual recruitment also shown. The stock is predicted to have never fallen to low enough levels that the steepness is obvious. However, the lowest levels of predicted spawning output showed some of the smallest recruitment events and very few above average recruitment events. Steepness was not estimated in this model, but sensitivities to alternative values of steepness were explored via likelihood profile analysis and discussed below. 

