
```{r  echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library("here")

sens_table <- read.csv(here::here("figures", "sensitivities", "short_sens_table.csv"))
sens_base_ssb <- round(as.numeric(sens_table[28, 2:dim(sens_table)[2]]), 2)
mod_names <- c("base model", 
               "fixing natural mortality to 0.1 (2015 assessment prior)", 
               "fixing natural mortality at 0.124 yr^-1^ for females and 0.129 yr^-1^ for males (2011 assessment prior)", 
               "forcing asymptotic selectivity on the midwater trawl fleet", 
               "fitting logistic curves for WCGBTS selectivity", 
               "including previously excluded shrimp trawls",
               "updated Washington catch reconstruction",
               # "fixing steepness at 0.4", 
               # "fixing steepness at 0.6", 
               # "fixing steepness at 0.798 (2015 assessment value)", 
               "exclusion of the triennial survey",
               "weighting the composition data using the Francis method")

# to get capitalized mod_name
firstup <- function(x) {
  substr(x, 1, 1) <- toupper(substr(x, 1, 1))
  x
}
```

## Model Diagnostics

### Convergence

```{r}
#| include: false

library(here)
### jitter results manually added to text
#jitters <- readRDS(here("models", "jitters", "jitter_summary.rds"))

#report
if (!exists("report")) {
  report <- r4ss::SS_output(
    here("models", "2025 base model"), 
    printstats = FALSE, 
    verbose = FALSE,
    SpawnOutputLabel = "Spawning output (billions of eggs)"
  )
}
rep <- report
#rep <- r4ss::SS_output(here("models","2025 base model"))

#Depletion estimates
read_rdata <- function(path) {load(path); get(ls()[ls() != "path"])}
ref_points <- read_rdata(here("report", "tables", "exec_summ_tables", "reference_points.rda"))$table
dep <- ref_points[ref_points$`Reference Point` == "2025 Fraction Unfished",c("Lower Interval","Upper Interval")]
depletion_ints <- c("low" = round(dep[[1]]*100,2), "high" = round(dep[[2]]*100,2))

depletion_ints[["low"]]


## Liklihood profiles 
nat_M_male <- read.csv(here(
  "models","diagnostics","2025 base model_profile_NatM_uniform_Mal_GP_1", 
  "NatM_uniform_Mal_GP_1_results.csv"
))
nat_M_female <- read.csv(here(
  "models","diagnostics","2025 base model_profile_NatM_uniform_Fem_GP_1", 
  "NatM_uniform_Fem_GP_1_results.csv"
))

steepness <- read.csv(here(
  "models", "diagnostics", "2025 base model_profile_SR_BH_steep", 
  "SR_BH_steep_results.csv"
))

ln_R0 <- read.csv(here(
  "models", "diagnostics", "2025 base model_profile_SR_LN(R0)", 
  "SR_LN(R0)_results.csv"
))


male_m_min <- round(nat_M_male[which.min(nat_M_male$likelihood),"parameter_value"], 3)
female_m_min <- round(nat_M_female[which.min(nat_M_female$likelihood),"parameter_value"], 3)
sr_min <- round(ln_R0[which.min(ln_R0$likelihood),"parameter_value"][[1]], 2)
steep_min <- round(steepness[which.min(steepness$likelihood),"parameter_value"], 3)

#base model R0 est
r0_bm <- round(rep$parameters[rep$parameters$Label == "SR_LN(R0)","Value"],3)
```

A number of tests were performed to verify convergence of the model. Following conventional AD Model Builder methods [@fournier_ad_2012], we checked that the Hessian matrix for the base model was positive-definite. There were no difficulties in inverting the Hessian to obtain asymptotic variance estimates. The maximum component of the final gradient is very low.

To confirm that the reported estimates were from the global best fit, we evaluated the model’s ability to recover similar likelihood estimates when initialized from dispersed starting points (jitter option in SS3). Starting parameters were jittered using a setting of 0.1 for 100 iterations. This perturbs the initial values used for minimization with the intention of causing the search to traverse a broader region of the likelihood surface. The majority (57 out of 100) returned to the same objective function value as the proposed model. The remaining runs exhibited worse fit than the proposed model. The spread of this search indicates that the jitter was sufficient to search a large portion of the likelihood surface, and that the model is in a global optimum.

We also completed a set of standard model diagnostics that included retrospective analysis and likelihood profiles described below.

### Parameter Uncertainty

Parameter estimates are shown in @tbl-params and @tbl-ts-popests-basecase along with approximate asymptotic standard errors. The only parameters with an absolute value of correlation greater than 0.95 were the female and male natural mortality parameters, which is expected. Estimates of key derived quantities are given in @tbl-key-deriv-params along with approximate 95% asymptotic confidence intervals. There is a reasonable amount of uncertainty in the estimates of biomass. The confidence interval of the 2025 estimate of the fraction of unfished spawning output is `r depletion_ints[["low"]]`%–`r depletion_ints[["high"]]`% and mostly above the management target of 40% of the unfished spawning output.

### Sensitivity Analyses

Sensitivity analysis was performed to determine the model behavior under different assumptions than those of the base case model. Broad range of sensitivity runs were conducted at multiple stages of the assessment. Here, we present only several of them:

1. Fixed natural mortality at 0.1 for both sexes (the median of the prior used in the assessment).
2. Fixed natural mortality at 0.144 yr^-1^ for females and 0.155 yr^-1^ for males (2019 assessment estimated values).
3. Fixed natural mortality at 0.157 yr^-1^ for females and 0.17 yr^-1^ for males (2015 assessment estimated values).
4. Estimating natural mortality using the prior based on the maximum age of 50 years old.
5. Estimating natural mortality using the prior based on the maximum age of 45 years old.
6. Using double-normal selectivity curve for \gls{s-wcgbt}, fixing it asymptotic.
7. Using double-normal selectivity curve for \gls{s-wcgbt}, allowed it to be dome-shaped.
8. Adding small amount of recreational catches as reported by WCGOP.

Differences in likelihood and estimates of key parameters and derived quantities are shown in @tbl-quant-sens-int. Predicted trajectories of spawning output, fraction unfished, recruitments and recruitment deviations comparisons of model estimates for 2025 are shown in @fig-spawnbio-sens-uncrt,  @fig-depl-sens, @fig-rec-sens, and @fig-recdevs-sens.  

Fixing $M$ at values lower than the base case estimate resulted in decreases in estimated spawning output in recent years. Fixing $M$ at 0.1, the median of a prior informed by maximum age of 54 years, resulted in largest decrease in spawning output and associated management quantities. The unfished spawning output in sensitivity runs with $M$ fixed at the estimates from the 2019 and 2015 assessments is very similar among these runs. However, the lower the natural mortality is, the lower the final assessment year spawning output is, since the less productive stock is impacted more by fishery removals. The models with higher natural mortality need to scale up estimated recruitments, in order to provide for a sufficient amount of fish to sustain the catches over time, while having the same recruitment deviations. 

The model results were not sensitive to the alternative values of the median of the $M$ prior, and estimated $M$ values were similar in all runs. This is an indication that the data provide a stronger signal about $M$ than the prior. 

The last three runs (changing the selectivity pattern for \gls{s-wcgbt} and adding recreational catches) did not have a noticeable impact. 

### Retrospective Analysis

A 10-year retrospective analysis (@fig-retros) was conducted by running the model removing the most recent 1 to 10 years of data (i.e., "Data -1 Years" corresponds to data through 2023 instead of 2024). Population trends from all retrospective runs were very close, and no concerning patterns were observed in the retrospective analysis. There is a general trend of increasing scale and fraction unfished as years are removed. This result is expected because the fraction of older fish in the \gls{s-wcgbt} increased substantially in recent years, providing more information on overall and recent productivity, and sequentially peeling off individual years of \gls{s-wcgbt} age data leads to increasing estimates of natural mortality and unfished recruitment. Additionally, as years are pulled away the estimate of the size of the 2016 year class decreases towards average and the sizes of year classes since 2016 increase towards average. This is also expected as information about those recent year classes is removed.

Retrospective analysis of recruitment deviations over the last 10 years is shown in @fig-retros-squid, illustrating which year classes are estimated to be large or small in the current assessment, and also what the recruitment deviations would have been estimated at with years of data sequentially removed. It is worth pointing out that despite substantial changes made to the fleet structure of the model (compared to the 2019 assessment), with seven years of data removed, the model would have still estimated the 2013 year class as very large, which gets smaller with every year of data added. This suggests that multiple sources of earlier data inform the model about the potential of a high 2013 year class, and  caution should be taken when large recent year classes are estimated in stock assessments. 

Estimated relative spawning output from the base model was also compared with previous assessments (@fig-ssb-historical). The current assessment follows a similar trend to previous assessments over the overlapping years, and estimated spawning output falls within the mid-range of previous assessments.

### Likelihood Profiles and key parameters {#sec-profiles}

Likelihood profiles were conducted for unfished virgin recruitment ($R_0$), steepness ($h$) and sex-specific natural mortality ($M$) values simultaneously. These likelihood profiles were conducted by fixing the parameter at specific values and removing the prior on the parameter being profiled. Without the original prior distribution the MLE estimates from the base case will likely be different than the MLE in the likelihood profile, but this displays what information the data have. 

For profiles of natural mortality, the negative log-likelihood was minimized at a value of `r male_m_min` for males, and a value of `r female_m_min` for females. Profiles for natural mortality for each sex are illustrated in (@fig-like_profile_MF, @fig-like_profile_MM). Profiles over female natural mortality indicate that the recruitment deviation prior and age data most strongly influence the estimate of natural mortality, with survey indices also being somewhat influential.Length data provides very little information. Profiles over male natural mortality behaved similarly. For steepness, the negative log-likelihood was minimized at a steepness of `r steep_min`, however the 95% confidence interval extends over the entire range of possible steepness values. Profiles for steepness are illustrated in (@fig-like_profile_Steepness). For $R_0$, the negative log-liklihood was minimized at a value of `r sr_min`, which supports the base model estimates of $R_0$ at `r r0_bm`.Profiles for $R_0$ are illustrated in @fig-like_profile_Rzero.

