
```{r  echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library("here")

sens_table <- read.csv(here::here("figures", "sensitivities", "short_sens_table.csv"))
sens_base_ssb <- round(as.numeric(sens_table[28, 2:dim(sens_table)[2]]), 2)
mod_names <- c("base model", 
               "fixing natural mortality to 0.1 (2015 assessment prior)", 
               "fixing natural mortality at 0.124 yr^-1^ for females and 0.129 yr^-1^ for males (2011 assessment prior)", 
               "forcing asymptotic selectivity on the midwater trawl fleet", 
               "fitting logistic curves for WCGBTS selectivity", 
               "including previously excluded shrimp trawls",
               "updated Washington catch reconstruction",
               # "fixing steepness at 0.4", 
               # "fixing steepness at 0.6", 
               # "fixing steepness at 0.798 (2015 assessment value)", 
               "exclusion of the triennial survey",
               "weighting the composition data using the Francis method")

# to get capitalized mod_name
firstup <- function(x) {
  substr(x, 1, 1) <- toupper(substr(x, 1, 1))
  x
}
```

## Model Diagnostics

### Convergence

```{r}
#| include: false

library(here)
jitters <- readRDS(here("models", "jitters", "jitter_summary.rds"))

#report
if (!exists("report")) {
  report <- r4ss::SS_output(
    here("models", "2025 base model"), 
    printstats = FALSE, 
    verbose = FALSE,
    SpawnOutputLabel = "Spawning output (billions of eggs)"
  )
}
rep <- report
#rep <- r4ss::SS_output(here("models","2025 base model"))

#Depletion estimates
read_rdata <- function(path) {load(path); get(ls()[ls() != "path"])}
ref_points <- read_rdata(here("report", "tables", "exec_summ_tables", "reference_points.rda"))$table
dep <- ref_points[ref_points$`Reference Point` == "2025 Fraction Unfished",c("Lower Interval","Upper Interval")]
depletion_ints <- c("low" = round(dep[[1]]*100,2), "high" = round(dep[[2]]*100,2))

depletion_ints[["low"]]


## Liklihood profiles 
nat_M_male <- read.csv(here(
  "models","diagnostics","2025 base model_profile_NatM_uniform_Mal_GP_1", 
  "NatM_uniform_Mal_GP_1_results.csv"
))
nat_M_female <- read.csv(here(
  "models","diagnostics","2025 base model_profile_NatM_uniform_Fem_GP_1", 
  "NatM_uniform_Fem_GP_1_results.csv"
))

steepness <- read.csv(here(
  "models", "diagnostics", "2025 base model_profile_SR_BH_steep", 
  "SR_BH_steep_results.csv"
))

ln_R0 <- read.csv(here(
  "models", "diagnostics", "2025 base model_profile_SR_LN(R0)", 
  "SR_LN(R0)_results.csv"
))


male_m_min <- round(nat_M_male[which.min(nat_M_male$likelihood),"parameter_value"], 3)
female_m_min <- round(nat_M_female[which.min(nat_M_female$likelihood),"parameter_value"], 3)
sr_min <- round(ln_R0[which.min(ln_R0$likelihood),"parameter_value"][[1]], 2)
steep_min <- round(steepness[which.min(steepness$likelihood),"parameter_value"], 3)

#base model R0 est
r0_bm <- round(rep$parameters[rep$parameters$Label == "SR_LN(R0)","Value"],3)
```

A number of tests were performed to verify convergence of the model, facilitated by the \{nwfscDiags\} package in R [@wetzel2025nwfscdiag]. Following conventional AD Model Builder methods [@fournier_ad_2012], we checked that the Hessian matrix for the base model was positive-definite. There were no difficulties in inverting the Hessian to obtain asymptotic variance estimates. The maximum component of the final gradient is very low (`r format(base_out$maximum_gradient_component, digits = 3, scientific = TRUE)`). 

To confirm that the reported estimates were from the global best fit, we evaluated the model’s ability to recover similar likelihood estimates when initialized from dispersed starting points (jitter option in SS3). Starting parameters were jittered using a setting of 0.05 for 100 iterations. This perturbs the initial values used for minimization with the intention of causing the search to traverse a broader region of the likelihood surface. The majority (56 out of 100) returned to the same objective function value as the proposed model. The remaining runs exhibited worse fit than the proposed model. The spread of this search indicates that the jitter was sufficient to search a large portion of the likelihood surface, and that the model is in a global optimum.

We also completed a set of standard model diagnostics that included retrospective analysis and likelihood profiles.

### Parameter Uncertainty

Parameter estimates are shown in @tbl-params and @tbl-ts-popests-basecase along with approximate asymptotic standard errors. The only parameters with an absolute value of correlation greater than 0.95 were the female and male natural mortality parameters, which is expected. Estimates of key derived quantities are given in @tbl-key-deriv-params along with approximate 95% asymptotic confidence intervals. There is a reasonable amount of uncertainty in the estimates of biomass. The confidence interval of the 2025 estimate of depletion is `r depletion_ints[["low"]]`%–`r depletion_ints[["high"]]`% and mostly above the management target of 40% of the unfished spawning output.

### Sensitivity Analyses

Sensitivity analysis was performed to determine the model behavior under different assumptions than those of the base case model. `r length(mod_names)-1` sensitivity analyses were conducted to explore the potential differences in model structure and assumptions, including:

1. Fixed natural mortality at 0.1 for both sexes (2015 assessment prior)
2. Fixed natural mortality at 0.124 yr^-1^ for females and 0.129 yr^-1^ for males (2011 assessment prior)
3. Forcing asymptotic selectivity on the midwater trawl fleet
4. Fitting logistic curves for WCGBTS selectivities
5. Weighting the composition data using the @francis_data_2011 method
6. Updated Washington catch reconstruction
7. Inclusion of previously excluded shrimp trawl data
8. Exclusion of triennial survey data

Likelihood values and estimates of key parameters are shown in @tbl-quant-sens-int. Predicted \gls{ssb} trajectories, estimated recruitment deviations and comparisons of model estimates for 2025 are shown in @fig-recdevs-sens, @fig-spawnbio-sens-uncrt and @fig-sens-summary. The estimates of SSB in 2025 ranged from `r min(sens_base_ssb)` to `r max(sens_base_ssb)` thousand metric tons across the sensitivity runs, with `r mod_names[which(sens_base_ssb == min(sens_base_ssb))]` resulting in the lowest estimate and `r mod_names[which(sens_base_ssb == max(sens_base_ssb))]` resulting in the highest estimate. Generally, the trajectory of the \gls{ssb} was qualitatively similar across all tested models, e.g., peak around late 1970s and late 2010s, projected decrease in biomass in 2025 followed by some recovery into the 2030s; the quantitative magnitude of these trends did vary across cases. 

Fixing M at values lower than the base case estimate resulted in decreases in estimated \gls{ssb}. The alternative weighting using the Francis method generally increased the estimate of \gls{ssb} across the time series. Other changes, including forcing asymptotic selectivity on the midwater fleet, forcing logistic selectivity on the \gls{s-wcgbt}, and excluding the triennial survey had relatively small effects on \gls{ssb} [@fig-spawnbio-sens-uncrt] and recruitment [@fig-recdevs-sens]. Changes to the historical landings (updating WA catch reconstruction and including shrimp trawl data) had no discernible on the estimated \gls{ssb}. 


### Retrospective Analysis

A 10-year retrospective analysis (@fig-retros) was conducted by running the model removing the most recent 1 to 10 years of data (i.e., "Data -1 Years" corresponds to data through 2023 instead of 2024). Population trends from all retrospective runs were very close, and no concerning patterns were observed in the retrospective analysis. There is a general trend of increasing scale and fraction unfished as years are removed. This result is expected because the fraction of older fish in the \gls{s-wcgbt} increased substantially in recent years, and sequentially peeling off individual years of \gls{s-wcgbt} age data leads to increasing estimates of natural mortality and unfished recruitment. Additionally, as years are pulled away the estimate of the size of the 2016 year class decreases towards average and the sizes of year classes since 2016 increase towards average. This is also expected as information about those recent year classes is removed.

Also, estimated relative spawning output from the base model was compared with previous assessments [@fig-ssb-historical]. The current assessment follows a similar trend to previous assessments over the overlapping years, and estimated spawning output falls within the mid-range of previous assessments.

### Likelihood Profiles and key parameters

Likelihood profiles were conducted for unfished virgin recruitment (*R0*), steepness (*h*) and sex-specific natural mortality (*M*) values simultaneously. These likelihood profiles were conducted by fixing the parameter at specific values and removing the prior on the parameter being profiled. Without the original prior distribution the MLE estimates from the base case will likely be different than the MLE in the likelihood profile, but this displays what information the data have. 

For profiles of natural mortality, the negative log-likelihood was minimized at a value of `r male_m_min` for males, and a value of `r female_m_min` for females. Profiles for natural mortality for each sex are illustrated in (@fig-like_profile_MF, @fig-like_profile_MM). Profiles over female natural mortality indicate that the recruitment deviation prior and age data most strongly influence the estimate of natural mortality, with survey indices also being somewhat influential.Length data provides very little information. Profiles over male natural mortality behaved similarly. For steepness, the negative log-likelihood was minimized at a steepness of `r steep_min`, however the 95% confidence interval extends over the entire range of possible steepness values. Profiles for steepness are illustrated in (@fig-like_profile_Steepness). For R0, the negative log-liklihood was minimized at a value of `r sr_min`, which supports the base model estimates of R0 at `r r0_bm`.Profiles for R0 are illustrated in @fig-like_profile_Rzero.
