
```{r  echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library("here")

sens_table <- read.csv(here::here("figures", "sensitivities", "short_sens_table.csv"))
sens_base_ssb <- round(as.numeric(sens_table[28, 2:dim(sens_table)[2]]), 2)
mod_names <- c("base model", 
               "fixing natural mortality to 0.1 (2015 assessment prior)", 
               "fixing natural mortality at 0.124 yr^-1^ for females and 0.129 yr^-1^ for males (2011 assessment prior)", 
               "forcing asymptotic selectivity on the midwater trawl fleet", 
               "fitting logistic curves for WCGBTS selectivity", 
               "including previously excluded shrimp trawls",
               "updated Washington catch reconstruction",
               # "fixing steepness at 0.4", 
               # "fixing steepness at 0.6", 
               # "fixing steepness at 0.798 (2015 assessment value)", 
               "exclusion of the triennial survey",
               "weighting the composition data using the Francis method")

# to get capitalized mod_name
firstup <- function(x) {
  substr(x, 1, 1) <- toupper(substr(x, 1, 1))
  x
}
```

## Model Diagnostics

<!-- Remove intro text: 
Three types of uncertainty are presented for the assessment of widow rockfish. First, uncertainty in the parameter estimates was determined using approximate asymptotic estimates of the standard error. These estimates were based on the maximum likelihood theory that the inverse of the Hessian matrix (the second derivative of the log-likelihood function with respect to the parameter vector) approaches the true uncertainty of the parameter estimates as the sample size approaches infinity. This approach takes into account the uncertainty in the data and supplies correlation estimates between parameters, but does not capture possible skewness in the error distribution of the parameters and may not accurately estimate the standard error in some cases (see [@stewart_comparison_2013]).

The second type of uncertainty that is presented is related to modeling and structural error. This uncertainty cannot be captured in the base model as it is related to errors in the assumptions used in specifying the base model. Therefore, sensitivity analyses were conducted where assumptions were modified to reveal the effect they have on the model results.

Lastly, a major axis of uncertainty was determined from a parameter or structural assumption that results in the greatest change in stock status and advice, and projections were made for different states of nature based upon that parameter or structural assumption.
-->

### Convergence

```{r}
#| include: false

library(here)
jitters <- readRDS(here("models", "jitters", "jitter_summary.rds"))

#Depletion estimates
read_rdata <- function(path) {load(path); get(ls()[ls() != "path"])}
ref_points <- read_rdata(here("report", "tables", "exec_summ_tables", "reference_points.rda"))$table
dep <- ref_points[ref_points$`Reference Point` == "2025 Fraction Unfished",c("Lower Interval","Upper Interval")]
depletion_ints <- c("low" = round(dep[[1]]*100,2), "high" = round(dep[[2]]*100,2))

depletion_ints[["low"]]


## Liklihood profiles 
nat_M_male <- read.csv(here("models","diagnostics","2025 base model_profile_NatM_uniform_Mal_GP_1","NatM_uniform_Mal_GP_1_quant_table.csv"))
nat_M_female <- read.csv(here("models","diagnostics","2025 base model_profile_NatM_uniform_Fem_GP_1","NatM_uniform_Fem_GP_1_quant_table.csv"))
steepness <- read.csv(here("models","diagnostics","2025 base model_profile_SR_BH_steep","SR_BH_steep_quant_table.csv"))
ln_R0 <- read.csv(here("models","diagnostics","2025 base model_profile_SR_LN(R0)","SR_LN(R0)_quant_table.csv"))


```


Model convergence was determined by examining the final gradient, checking that the Hessian was positive (semi-)definite, and initializing the minimizer from perturbed  values around the maximum likelihood estimates (MLE) to determine if the model found a better minimum (“jittering”). Initial jittering analyses indicated that the model had converged to a false (local) minimum, though the difference in log-likelihoods at the previous estimates and the new MLE was small (<0.5). After accepting the new MLE as the base model, jittering was repeated 100 times with a jitter coefficient of 0.10 and (using a convergence threshold of 0.01) a better minimum was not found. `r paste0(jitters$njitters_at_base, "%")` of the jittered models achieved the minimum negative log-likelihood and `r paste0(jitters$njitters_within_2, "%")` were within two likelihood units. Through the jittering done as explained above and likelihood profiles, we are confident that the base case as presented represents the best fit to the data given the assumptions made. There were no difficulties in inverting the Hessian to obtain the parameter variance-covariance matrix. Likelihood profile runs which fixed M, R0 and h at more extreme values did not initially converge. This was addressed by using the parameter estimates from models with less extreme values as starting values in subsequent models. Convergence was defined as the lowest negative log-likelihood achieved with jittering where the Hessian matrix was invertible. 

### Parameter Uncertainty

Parameter estimates are shown in @tbl-params and @tbl-ts-popests-basecase along with approximate asymptotic standard errors. The only parameters with an absolute value of correlation greater than 0.95 were the female and male natural mortality parameters, which is expected. Estimates of key derived quantities are given in @tbl-key-deriv-params along with approximate 95% asymptotic confidence intervals. There is a reasonable amount of uncertainty in the estimates of biomass. The confidence interval of the 2025 estimate of depletion is `r depletion_ints[["low"]]`%–`r depletion_ints[["high"]]h`% and mostly above the management target of 40% of the unfished \gls{ssb}.

### Sensitivity Analyses

Sensitivity analysis was performed to determine the model behavior under different assumptions than those of the base case model. `r length(mod_names)-1` sensitivity analyses were conducted to explore the potential differences in model structure and assumptions, including:

1. Fixed natural mortality at 0.1 for both sexes (2015 assessment prior)
2. Fixed natural mortality at 0.124 yr^-1^ for females and 0.129 yr^-1^ for males (2011 assessment prior)
3. Forcing asymptotic selectivity on the midwater trawl fleet
4. Fitting logistic curves for WCGBTS selectivities
5. Weighting the composition data using the @francis_data_2011 method
6. Updated Washington catch reconstruction
7. Inclusion of previously excluded shrimp trawl data
8. Exclusion of triennial survey data

Likelihood values and estimates of key parameters are shown in @tbl-quant-sens-int. Predicted \gls{ssb} trajectories, estimated recruitment deviations and comparisons of model estimates for 2025 are shown in @fig-recdevs-sens, @fig-spawnbio-sens-uncrt and @fig-sens-summary. The estimates of SSB in 2025 ranged from `r min(sens_base_ssb)` to `r max(sens_base_ssb)` thousand metric tons across the sensitivity runs, with `r mod_names[which(sens_base_ssb == min(sens_base_ssb))]` resulting in the lowest estimate and `r mod_names[which(sens_base_ssb == max(sens_base_ssb))]` resulting in the highest estimate. Generally, the trajectory of the \gls{ssb} was qualitatively similar across all tested models, e.g., peak around late 1970s and late 2010s, projected decrease in biomass in 2025 followed by some recovery into the 2030s; the quantitative magnitude of these trends did vary across cases. 

Fixing M at values lower than the base case estimate resulted in decreases in estimated \gls{ssb}. 

Forcing asymptotic selectivity on the midwater fleet decreased estimated SSB in the early years, but had little effect in recent years. Forcing logistic selectivity on the \gls{s-wcgbt} increased peak SSB estimates in 2017 but likely had little effect on recent estimates. Excluding the triennial survey decreased estimated \gls{ssb} dramatically and decreased recruitment deviations in the 2019-2025 period [@fig-recdevs-sens]. Updating WA catch reconstruction and adding shrimp trawl data to the historical landings had almost no impact on the estimated \gls{ssb}. 

The alternative weighting using the Francis method generally increased the estimate of \gls{ssb} across the time series.

### Retrospective Analysis

First, a 5-year retrospective analysis was conducted by running the model using data only through 2020, 2021, 2022, 2023, and 2024 progressively (@fig-retros). The initial scale of the \gls{ssb} was effectively unchanged for all retrospective runs. Removing 4–5 years of data led to slightly higher estimated \gls{ssb} over the last 15 years.In contrast, removing only 1–2 years resulted lower estimated \gls{ssb} over the period. Removing 1-5 years of data resulted in increased estimated  recruitment over the past 5 years, but decreased estimates in 10 years prior to this. Despite these minor differences, population trends from all retrospective runs were very close, and there were no consistent patterns as years were removed. No concerning patterns were observed in the retrospective analysis.

Estimated \gls{ssb} from the base model was comapred with previous assessments. For assessment prior to 2009 (2000, 2003, 2005, 2009), \gls{ssb} was compared with spawning output, and \gls{ssb} for models thereafter (2011, 2015, 2019) [@fig-ssb-historical].The current update follows a similar trend to prvious assessments over the model period, and estimated \gls{ssb} falls within the mid-range of previous assessments.


### Likelihood Profiles and key parameters

Likelihood profiles were conducted for unfished virgin recruitment (*R0*), steepness (*h*) and sex-specific natural mortality (*M*) values simultaneously. These likelihood profiles were conducted by fixing the parameter at specific values and removing the prior on the parameter being profiled. Without the original prior distribution the MLE estimates from the base case will likely be different than the MLE in the likelihood profile, but this displays what information the data have. 

For profiles of natural mortality, the negative log-likelihood was minimized at a value of 0.132 for males, and a value of 0.121 for females. Profiles for natural mortality for each sex are illustrated in (@fig-like_profile_MF, @fig-like_profile_MM). For steepness, the negative log-likelihood was minimized at a steepness of 0.715, however the 95% confidence interval extends over the entire range of possible steepness values. Profiles for steepness are illustrated in (@fig-like_profile_Steepness). For R0, the total likelihood supported the estimated value, though there was variable support for each likelihood component across the range of R0 evaluated. As R0 increased, natural mortality also increased and the relative \gls{ssb} in 2024 was less depleted. Profiles for R0 are illustrated in @fig-like_profile_Rzero.
