
<!-- -->
<!-- Load the report from the completed model -->
```{r  echo=FALSE, results='hide', message=FALSE, warning=FALSE }

# This will need to be adjusted to point at the final model

#renv::activate()
model_path <- here::here("models", "2025 base model")

report <- r4ss::SS_output(dir = model_path, verbose = FALSE, printstats = FALSE)
#r4ss::SSexecutivesummary(replist = report) 

catch_sum <- aggregate(report$catch$Exp, by = list(Yr = report$catch$Yr), sum)

##Now parse exec tables into the items we wan
```




# Management 

## Reference Points

```{r  echo=FALSE, results='hide', message=FALSE, warning=FALSE}

load(file = here::here("report", "tables", "reference_points.rda"))
df <- as.data.frame(reference_points)
colnames(df) <- c("Label", "Est", "Lwr", "Upr")

ref_pts <- setNames(
  lapply(seq_len(nrow(df)), function(i) {
    vals <- df[i, c("Est", "Lwr", "Upr")]
    vals <- format(round(as.numeric(vals)), big.mark = ",", scientific = FALSE)
    setNames(as.list(vals), c("Est", "Lwr", "Upr"))
  }),
  nm = gsub("\\\\%", "%", df$Label)
)

```


Reference points were calculated using the estimated selectivities and catch distribution among fleets in the most recent year of the model (2024). Sustainable total yields (landings plus discards) were `r ref_pts[["Yield with SPR50 at SB SPR (mt)"]]$Est` mt when using an SPR50% reference harvest rate and with a 95% confidence interval of `r ref_pts[["Yield with SPR50 at SB SPR (mt)"]]$Lwr` to `r ref_pts[["Yield with SPR50 at SB SPR (mt)"]]$Upr` mt. The spawning biomass equivalent to 40% of the unfished spawning output (SB40%) was `r ref_pts[["Proxy Spawning Biomass (mt) SB40%"]]$Est` mt. Catches between the late 1990's and 2016 were well below the point estimate of potential long-term yields calculated using an SPR50% reference point, and the population is estimated to have increased continuously throughout that time period. However, catches from 2017 through 2024 were above the point estimate of potential long-term yields using an SPR50% reference point, exceeding the upper bound of the confidence interval in all years since 2018, and exceeding the point estimate by on average `r paste0(round(100 * (mean(catch_sum$x[catch_sum$Yr >= 2018]/ df$Est[df$Label == "Yield with SPR50 at SB SPR (mt)"]) - 1)), "%")`.

The predicted spawning biomass from the base model generally showed a slight decline until the late 1970s, followed by steep increase above unfished equilibrium biomass and reaching a peak in `r report$timeseries$Yr[which.max(report$timeseries$SpawnBio)]`. This was followed by a steep decrease up to the mid-1980s, and then a more gradual decease through 2000 (Figure 61). Between 2001 and 2016, the spawning biomass increased continuously due to small catches, and several years of high recruitment (though with lower than average recruitment in other recent years). The spawning biomass relative to unfished equilibrium spawning biomass climbed above the target of 40% of unfished spawning biomass in the early 2000's. It is estimated to still be above the target, though the lower bound of the 95% confidence interval lies between the target and minimum stock size threshold (Figure 63). The fishing intensity (relative 1-SPR) exceeded the current estimates of the harvest rate limit (SPR50%) throughout the 1980s and early 1990s, and has again since 2018 (Figure 73). Recent exploitation rates on Widow Rockfish were predicted to be much less than target levels. In recent years, the stock has experienced exploitation rates that have been above the target level while the biomass level has remained above the target level (Figure 74).

The equilibrium yield plot is shown in Figure 75, based on a steepness value fixed at 0.720. The predicted maximum sustainable yield under the assumptions of this assessment occurs near 25% of equilibrium unfished spawning biomass.


## Unresolved problems and major uncertainties


## Harvest Projections and Decision Tables

```{r}
load(file=here::here("report","tables","projections.rda"))
projections  <- as.data.frame(projections$table)

projections <- projections|>dplyr::mutate(across(2:ncol(projections), ~ round(.x)))

```




Projections with catches based on the predicted annual catch limit (ACL) using the SPR rate of 50%, the
40:10 control rule, and a 0.25 P* adjustment using a sigma of 0.50 from 2021 onward suggest that the
spawning biomass will decrease over the projection period for all states of nature. Predicted ACL catches
range from 10,961 mt in 2021 to 5,944 mt in 2030. 

***nOT SURE WHERE acl CATCHES ARE ***

## Evaluation of Scientific Uncertainty

## Regional management considerations


Widow Rockfish have shown latitudinal differences in life-history parameters, which has led past assessment authors to pursue a two-area model. Modelling a stock with two areas is difficult because it requires many assumptions about recruitment distribution, movement, and connectivity, while also splitting data into two areas that reduces sample sizes when compared to a coastwide model. The upside is that it can result in a better model that more accurately predicts regional status. This assessment is a coastwide model because not enough is known about the assumptions that would have to be made for a two-area model. 

It is still important to consider regional differences when making management decisions. Following recent cohorts through time with survey data showed that older fish showed up in the north after younger fish were observed in the south (Figure 2). This may indicate connectivity between the north and the south and that this is truly one stock. However, more investigation is needed.

Widow Rockfish are managed on a coastwide basis and observed more often in the NWFSC WCGBT bottom trawl survey north of latitude 40° 10′ N. Bottom trawl catches in California have historically been as large as in Oregon and larger than in Washington, but recently catches in California have been small. Rockfish Conservation Areas (RCAs) cover a significant proportion of Widow Rockfish habitat, but a midwater trawl fishery is beginning to re-develop that can fish in these areas. Future assessments and management of Widow Rockfish may want to monitor where catches are being taken to make sure that specific areas are not being overexploited. In addition, research on the connectivity along the coast as well as regional differences would help to inform the potential for overfishing specific areas.
>>>>>>> 1ed44512b0ba062144bae79bac8c50155d6b6fec

## Research and Data Needs

<!-- 1. Describe progress on Research and Data Needs items identified in the most recent previous stock assessment document and associated STAR panel report.  -->
<!-- 2. Describe new research and data needs and specify their priority (high, medium, low).  -->

There are many areas of research that could be improved to benefit the understanding and assessment of
Widow Rockfish. Below, we specifically identify five topics that we believe are most important (order
does not indicate importance).

* **Historical landings and discards:** The historical landings and discards are uncertain for Widow
Rockfish and improvements would increase the certainty that fishing removals are applied
appropriately. Because landings are assumed to be known exactly in the assessment model,
uncertainty in the predictions does not include uncertainty in the landings. A thorough look at
historical landings, species compositions, and discarding practices would reduce the potential
uncertainty that is not entirely accounted for.

* **Natural mortality:** Uncertainty in natural mortality translates into uncertain estimates of status
and sustainable fishing levels for Widow Rockfish. The collection of additional age data, re-
reading of older age samples, reading old age samples that are unread, and improved
understanding of the life-history of Widow Rockfish may reduce that uncertainty.

* **Maturity and fecundity:**: There are few studies on the maturity of Widow Rockfish and even
less recent information. There have been no studies that reported results of a histological
analysis. Further research on the maturity and fecundity of Widow Rockfish, the potential
differences between areas, the possibility of changes over time would greatly improve the
assessment of these species.
* **Age data and error:** There is a considerable amount of error in the age data and potential for
bias. Investigating the ageing error and bias would help to understand the influences that the age
data have on this assessment. Disagreement between surface reads of age versus break-and-burn
reads is worth more investigation.
* **Basin-wide understanding of stock structure, biology, connectivity, and distribution:** This is
a stock assessment for Widow Rockfish off of the west coast of the U.S. and does not consider
data from British Columbia or Alaska. Further investigating and comparing the data and
predictions from British Columbia and Alaska to determine if there are similarities with the U.S.
West Coast observations would help to define the connectivity between Widow Rockfish north
and south of the U.S.-Canada border.