
<!-- -->
<!-- Load the report from the completed model -->

```{r echo = FALSE, results = 'hide', message = FALSE, warning = FALSE }

# This will need to be adjusted to point at the final model

library("here")
read_rdata <- function(path) {load(path); get(ls()[ls() != "path"])}

#renv::activate()
model_path <- here::here("models", "2025 base model")

report <- r4ss::SS_output(dir = model_path, verbose = FALSE, printstats = FALSE)
#r4ss::SSexecutivesummary(replist = report) 

catch_sum <- aggregate(report$catch$Exp, by = list(Yr = report$catch$Yr), sum)

##Now parse exec tables into the items we wan
```


# Management 

## Reference Points

```{r echo = FALSE, results = 'hide', message = FALSE, warning = FALSE }

load(file = here::here("report", "tables", "exec_summ_tables", "reference_points.rda"))
df <- as.data.frame(reference_points$table)
colnames(df) <- c("Label", "Est", "Lwr", "Upr")

ref_pts <- setNames(
  lapply(seq_len(nrow(df)), function(i) {
    vals <- df[i, c("Est", "Lwr", "Upr")]
    vals <- format(round(as.numeric(vals)), big.mark = ",", scientific = FALSE)
    setNames(as.list(vals), c("Est", "Lwr", "Upr"))
  }),
  nm = gsub("\\\\%", "%", df$Label)
)

```

Reference points were calculated using the estimated selectivities and catch distribution among fleets in the most recent year of the model (2024). Sustainable total yields (landings plus discards) were `r ref_pts[["Yield with SPR50 at SB SPR (mt)"]]$Est` mt when using an SPR50% reference harvest rate and with a 95% confidence interval of `r ref_pts[["Yield with SPR50 at SB SPR (mt)"]]$Lwr` to `r ref_pts[["Yield with SPR50 at SB SPR (mt)"]]$Upr` mt. The spawning biomass equivalent to 40% of the unfished spawning output (SB40%) was `r ref_pts[["Proxy Spawning Biomass (mt) SB40%"]]$Est` mt. Catches between the late 1990's and 2016 were well below the point estimate of potential long-term yields calculated using an SPR50% reference point, and the population is estimated to have increased continuously throughout that time period. However, catches from 2017 through 2024 were above the point estimate of potential long-term yields using an SPR50% reference point, exceeding the upper bound of the confidence interval in all years since 2018, and exceeding the point estimate by on average `r paste0(round(100 * (mean(catch_sum$x[catch_sum$Yr >= 2018]/ df$Est[df$Label == "Yield with SPR50 at SB SPR (mt)"]) - 1)), "%")`.

The predicted spawning biomass from the base model generally showed a slight decline until the late 1970s, followed by steep increase above unfished equilibrium biomass and reaching a peak in `r report$timeseries$Yr[which.max(report$timeseries$SpawnBio)]`. This was followed by a steep decrease up to the mid-1980s, and then a more gradual decease through 2000 (@fig-rel-spawn-bio). Between 2001 and 2016, the spawning biomass increased continuously due to small catches, and several years of high recruitment (though with lower than average recruitment in other recent years). The spawning biomass relative to unfished equilibrium spawning biomass climbed above the target of 40% of unfished spawning biomass in the early 2000's. It is estimated to still be above the target, though the lower bound of the 95% confidence interval lies between the target and minimum stock size threshold (@fig-rel-spawn-bio). The fishing intensity (relative 1-SPR) exceeded the current estimates of the harvest rate limit (SPR50%) throughout the 1980s and early 1990s, and has again since 2018 (@fig-pred-one-SPR). Exploitation rates on widow rockfish between 2001 and 2016 were predicted to be far below target levels. In recent years, the stock has experienced exploitation rates that have been above the target level while the biomass level has remained above the target level (@fig-relative-one-SPR).

The equilibrium yield plot is shown in @fig-est-eqi-yield, based on a steepness value fixed at 0.720. The predicted maximum sustainable yield under the assumptions of this assessment occurs near 27% of equilibrium unfished spawning biomass, however this represents only 12% more yield, with considerably more risk, than under the 50% SPR policy, which occurs near 45% of unfished spawning biomass. 


## Unresolved problems and major uncertainties

```{r echo = FALSE, results = 'hide', message = FALSE, warning = FALSE }

steep_proflik <- read.csv(here::here("report", "tables", "SR_BH_steep_quant_table.csv"))
names(steep_proflik)[1] <- "var"

steep_yspr50 <- as.numeric(steep_proflik[steep_proflik$var == "Yield SPR 50",-1])
steep_yspr50 <- round(steep_yspr50[round(steep_yspr50) > 0])

```

Major sources of uncertainty include landings, discards, natural mortality, and recruitment, which are discussed below. 

Because widow rockfish is a marketable species, historical discard rates were likely lower than less desirable or smaller species. This assessment assumed that discarding was nearly negligible before management restrictions began in 1982. Once trip limits were introduced, discarding tended to be an all or none event, and detecting large, but rare, discard events with far less than 100% observer coverage has a low probability. From 2002 onward, the WCGOP has provided data on discards from vessels that were randomly selected for observer coverage, thus some uncertainty is present in the total amount discarded. The implementation of trawl rationalization in 2011 resulted in almost 100% observer coverage for the trawl fleet and very little incentive to discard widow rockfish. The open access fixed-gear fleet is not monitored by the full observer coverage required under trawl rationalization. Discard mortality is assumed to be 100%, which may overestimate actual mortality (Jarvis & Lowe, 2007), but given the low number of discards, is likely to have a minimal effect on assessment results.

There may also be uncertainty in the ability of bottom trawl surveys to be a reliable measure of widow abundance, which spend a significant portion of their time in mid-water (Wilkins 1986). Multiple surveys are used in the assessment, but further consideration of additional surveys is reasonable.

Widow rockfish is a relatively long-lived fish, and natural mortality is likely to be lower than many species of fish, such as gadoids. Ages above 50 years have been observed and it is expected that natural mortality could be less than 0.10 yr^-1^ (the median of the prior for natual mortality used in this assessment). However, even with length and age data available back to the late 1970s, natural mortality was estimated at `r round(report$parameters["NatM_uniform_Fem_GP_1",]$Value, 3)` for females and `r round(report$parameters["NatM_uniform_Mal_GP_1",]$Value, 3)` yr^-1^ for males, with a small amount of uncertainty (e.g., a `r paste0(round(100 * report$parameters["NatM_uniform_Fem_GP_1",]$Parm_StDev/report$parameters["NatM_uniform_Fem_GP_1",]$Value, 1), "%")` coefficient of variation for females). This assessment attempts to capture that uncertainty by estimating natural mortality (M) and integrating that uncertainty into the derived biomass estimates, as well as additional uncertainty by including levels outside of the predicted interval in a decision table. One contributing factor to this change is the increase in the mean age of the catch observed by the \gls{s-wcgbt} survey in recent years. These observations are poorly fit by the model, and considertation should be given in future to structural changes which may improve the fit to \gls{s-wcgbt} CAAL data. One possible explanation as to the increased observation of older fish is that successfully rebounding of the stock as a result of reduced fishing pressure form 2003-2016 allowed year classes to grow older and are now being fully observed by \gls{s-wcgbt}.

Model sensitivities and profiles over M showed that current stock status was highly sensitive to the assumption about natural mortality. Notably, the estimated natural mortality values for both sexes from this assessment are lower than those from the 2015 and 2019 assessments, for which natural mortality was estimated above 0.15 yr^-1^ and 0.14 yr^-1^, respectively.  The estimates of M varied slightly depending on the weight given to age and length data, or removing recent years of data, but M was always estimated above 0.12 yr^-1^. The likelihood profile over natural mortality provides support for values up to or above 0.14 yr^-1^, but with greater curvature than in the 2019 assessment, suggesting additional data has reduced the support for higher natural mortality values.

Steepness was fixed at 0.720 in the base model, but a likelihood profile showed that it would be estimated at a value less than that. Estimates of M increased with lower steepness, while unfished equilibrium spawning biomass increased and current spawning biomass decreased. Sustainable yields at the SPR50% reference harvest rate ranged from approximately `r min(steep_yspr50)` to `r max(steep_yspr50)` mt depending on the value of steepness.



## Harvest Projections and Decision Tables

```{r echo = FALSE, results = 'hide', message = FALSE, warning = FALSE }

projections <- read_rdata(here("report", "tables", "exec_summ_tables", "projections.rda"))$table
projections <- projections |> dplyr::mutate(across(is.numeric, ~ round(.x)))

```



```{r}
#| include: false

##dec table params
dec_table_pars <- report$parameters|> #make sure to ppull from report, not input file so est values are used
  mutate(
         value = Value,
         sd = if_else(is.na(Parm_StDev),Pr_SD,Parm_StDev))|> #If est par, use param sd, otherwise use prior sd
  filter(Num %in% c(1,13,24))|> #natM Female, natM male, steepness
  select(value,sd)|>
  mutate(
    q125 = round(qnorm(0.125,mean = value,sd = sd),3), #apply the quantiles
    q875 = round(qnorm(0.875,mean = value,sd = sd),3),
    phase = -1,
    name = c("NatM_p_1_Fem_GP_1", #name the parms - easier to do manually as report and r4ss::read() have different parameter names
             "NatM_p_1_Mal_GP_1",
             "SR_BH_steep") 
  )

#dec tables stuff
dec_table_pars["NatM_uniform_Fem_GP_1","q125"]
colnames(dec_table_pars)

dec_table_res <- read.csv(here("data_derived", "decision_table", "dec_table_formatted.csv"))

##gmt stuff
gmt_catch <- data.frame(read.csv(here("data_provided", "gmt_forecast_catch", "GMT_forecast_catch.csv"))|>
  group_by(year)|>
  summarise(total_catch = sum(catch_mt)))

dec_table_summ <- list(
  "p45_2027" = dec_table_res|>filter(scenario == "P*0.45" & yr == 2027)|>pull(catch),
  "p45_2036" = dec_table_res|>filter(scenario == "P*0.45" & yr == 2036)|>pull(catch),
  "p40_2027" = dec_table_res|>filter(scenario == "P*0.40" & yr == 2027)|>pull(catch),
  "p40_2036" = dec_table_res|>filter(scenario == "P*0.40" & yr == 2036)|>pull(catch)
)
```


Two categories of parameters that greatly contributed to uncertainty in the results were natural mortality (an important estimated parameter) and steepness (not estimated in the model). A combination of these two factors was used as the axis of uncertainty to define low and high states of nature. The 12.5% and 87.5% quantiles for female and male natural mortality (independently) were chosen as low and high values (`r dec_table_pars["NatM_uniform_Fem_GP_1","q125"]` yr^-1^ and `r dec_table_pars["NatM_uniform_Fem_GP_1","q875"]` yr^-1^ for females; `r dec_table_pars["NatM_uniform_Mal_GP_1","q125"]` yr^-1^ and `r dec_table_pars["NatM_uniform_Mal_GP_1","q875"]` yr^-1^ for males). Steepness is probably the most important factor since it was fixed in the base model and is not incorporated in the estimation uncertainty. The 12.5% and 87.5% quantiles from the steepness prior (without widow rockfish data) were used to define the low and high values of steepness (`r dec_table_pars["SR_BH_steep","q125"]` and `r dec_table_pars["SR_BH_steep","q875"]`). The low combination of these two factors defined the low state of nature and the high combination of these two factors defined the high state of nature. The predictions of spawning biomass in 2025 from the low and high states of nature are close to the 12.5% and 87.5% lognormal quantiles from the base model.

Previous assessments included recent large recruitments as a third axis of uncertainty. As no large recruitments occured in the recent period, this axis was not included in the harvest projection.

This assessment synthesizes many sources of data and estimates recruitment variability, thus it is classified as a Category 1 stock assessment. Therefore, the sigma to determine the catch reduction to account for scientific uncertainty is 0.50.

A twelve year projection of the base model with two catch streams based on ACL adjustments of 0.45 P* and 0.40 P* were conducted (@tbl-dec-table).

Projections with catches based on the predicted annual catch limit (ACL) using the SPR rate of 50%, the 40:10 control rule, and a 0.45 P* adjustment using a sigma of 0.50 from 2027 onward suggest that the spawning biomass will decrease over the projection period for all states of nature (@tbl-dec-table). Predicted ACL catches range from `r format(dec_table_summ$p45_2027, big.mark = ",", digits = 1)` mt in 2027 to `r format(dec_table_summ$p45_2036, big.mark = ",", digits = 1)` mt in 2036. 

Projections with catches based on the predicted annual catch limit (ACL) using the SPR rate of 50%, the 40:10 control rule, and a 0.40  P* adjustment using a sigma of 0.50 from 2027 onward suggest that the spawning biomass will decrease over the projection period for all states of nature. Predicted ACL catches range from `r format(dec_table_summ$p40_2027, big.mark = ",", digits = 1)` mt in 2027 to `r format(dec_table_summ$p40_2036, big.mark = ",", digits = 1)` mt in 2036 @tbl-dec-table, @fig-dec-table-plot). 

This assessment synthesizes many sources of data and estimates recruitment variability, thus it is classified as a Category 1 stock assessment. Therefore, the sigma for P* to determine the catch reduction to account for scientific uncertainty is 0.50.

## Evaluation of Scientific Uncertainty

Spawning biomass is estimated to be at `r format(SSB["SSB_2025","Value"], big.mark = ",", digits = 1)` mt in 2025, with a sigma of `r format(SSB["SSB_2025","StdDev"]/SSB["SSB_2025","Value"], big.mark = ",", digits = 4)`. OFL is estimated to be `r format(report$derived_quants["OFLCatch_2025","Value"], big.mark = ",", digits = 1)` mt in 2025 with a coefficient of variation of `r format(report$OFL_sigma, big.mark = ",", digits = 4)`.

## Regional management considerations

Widow rockfish are managed on a coastwide basis and observed more often in the WCGBTS north of latitude 40° 10′ N. Bottom trawl catches in California have historically been as large as in Oregon and larger than in Washington, but recently catches in California have been small. Future assessments and management of widow rockfish may want to monitor where catches are being taken to make sure that specific areas are not being overexploited. In addition, research on the connectivity along the coast as well as regional differences would help to inform the potential for overfishing specific areas.

## Research and Data Needs

<!-- 1. Describe progress on Research and Data Needs items identified in the most recent previous stock assessment document and associated STAR panel report.  -->
<!-- 2. Describe new research and data needs and specify their priority (high, medium, low).  -->

There are many areas of research that could be improved to benefit the understanding and assessment of widow rockfish:

- **Natural mortality:** Uncertainty in natural mortality translates into uncertain estimates of status and sustainable fishing levels for widow rockfish. The collection of additional age data, re-reading of older age samples, reading old age samples that are unread, and improved understanding of the life-history of widow rockfish may reduce that uncertainty. Investigating the ageing error and bias would help to understand the influences that the age data have on this assessment, particularly the influence of age data on natural mortality.

- **Historical landings and discards:** Although progress has been made in reconstructing historical landings of rockfish on the U.S. West Coast, the historical landings and discards continue to be uncertain for widow rockfish and improvements would increase the certainty that fishing removals are applied appropriately. Because landings are assumed to be known exactly in the assessment model, uncertainty in the predictions does not include uncertainty in the landings. A thorough look at historical landings, species compositions, and discarding practices would reduce the potential uncertainty that is not entirely accounted for. In addition, discard composition data have become available for the midwater trawl fleet in recent years and should be included in future assessments.

<!-- - **Maturity and fecundity::** There are few studies on the maturity of widow rockfish and even less recent information. There have been no studies that reported results of a histological analysis. Further research on the maturity and fecundity of widow rockfish, the potential differences between areas, the possibility of changes over time would greatly improve the assessment of these species. -->

<!-- - **Age data and error:** There is a considerable amount of error in the age data and potential for bias. Investigating the ageing error and bias would help to understand the influences that the age data have on this assessment. Disagreement between surface reads of age versus break-and-burn reads is worth more investigation. -->

- **Sex-specific selectivity:** The midwater and bottom trawl length-composition data fits showed divergent residual patterns between male and female fish. The underlying mechanism driving this pattern is unclear, and could be related to growth, sexing error, or to sex-specific selectivity (e.g., when widow rockfish aggregate, sexes possibly may be aggregating separately). Sex-specific selectivity for these two fleets could be explored or included to address this.

- **Coastwide understanding of stock structure, biology, connectivity, and distribution:** This is a stock assessment for widow rockfish off of the west coast of the U.S. and does not consider data from British Columbia or Alaska. Further investigating and comparing the data and predictions from British Columbia and Alaska to determine if there are similarities with the U.S. West Coast observations would help to define the connectivity between widow rockfish north and south of the U.S.-Canada border.