<!-- -->
<!-- Load the report from the completed model -->
```{r  echo=FALSE, results='hide', message=FALSE, warning=FALSE }

# This will need to be adjusted to point at the final model

#renv::activate()
model_path <- here::here("models", "2019 base model","Base_45_new")

report <- r4ss::SS_output(dir = model_path)
#r4ss::SSexecutivesummary(replist = report) 


##Now parse exec tables into the items we wan
```




# Management 

## Reference Points

```{r  echo=FALSE, results='hide', message=FALSE, warning=FALSE}



load(file=here::here("report","tables","reference_points.rda"))
df <- as.data.frame(reference_points)
colnames(df) <- c("Label", "Estimate", "Lower.Interval", "Upper.Interval")

reference_points <- setNames(
  lapply(seq_len(nrow(df)), function(i) {
    vals <- df[i, c("Estimate", "Lower.Interval", "Upper.Interval")]
    vals <- format(round(as.numeric(vals)), big.mark = ",", scientific = FALSE)
    setNames(as.list(vals), c("Estimate", "Lower.Interval", "Upper.Interval"))
  }),
  nm = gsub("\\\\%", "%", df$Label)
)


```


<!-- 1. Unfished spawning stock biomass, summary age biomass, and recruitment, along with unfished spawning stock output.  -->
<!-- 2. Reference points based on B40% for rockfish and round fish and on B25% for flatfish (spawning biomass and/or output, SPR, exploitation rate, equilibrium yield).  -->
<!-- 3.   Reference points based on default SPR proxy (spawning biomass and/or output, SPR, exploitation rate, equilibrium yield).  -->
<!-- 4. Reference points based on MSY (if estimated) (spawning biomass and/or output, SPR, exploitation rate, equilibrium yield).  -->
<!-- 5. Equilibrium yield curve showing various BMSY proxies.   -->
 
<!--  Reference points (see Table e in the Executive Summary) were calculated using the estimated selectivities -->
<!-- and catch distribution among fleets in the most recent year of the model (`r report[["endyr"]]`). Sustainable total yields -->
<!-- (landings plus discards) were `r reference_points[["Yield with SPR50 at SB SPR (mt)"]][["Estimate"]]` mt when using an SPR50% reference harvest rate and with a 95% -->
<!-- confidence interval of 5,447 `r reference_points[["Yield with SPR50 at SB SPR (mt)"]][["Lower.Interval"]]` -->
<!-- to 9,033 `r reference_points[["Yield with SPR50 at SB SPR (mt)"]][["Upper.Interval"]]` -->
<!-- mt based on estimates of uncertainty. The spawning biomass -->
<!-- equivalent to 40% of the unfished spawning output (SB40%) was 35,198 `r reference_points[["Proxy Spawning Biomass (mt) SB40%"]][["Estimate"]]` -->
<!-- mt. Prior to `r report[["endyr"]]`, the most recent -->
catches (landings plus discards) have been below the point estimate of potential long-term yields
calculated using an SPR50% reference point and the population has been increasing over the last decade.
However, catches in 2018 were above the point estimate of potential long-term yields calculated using an
SPR50 reference point.

The predicted spawning biomass from the base model generally showed a slight decline until the late
1970s, steep increase above unfished equilibrium levels, then a steep decline until the mid-1980s followed
by less of a decline until 2001 (Figure 61). Since 2001, the spawning biomass has been increasing due to
small catches, and recently, above average recruitment. The 2018 spawning biomass relative to unfished
equilibrium spawning biomass is above the target of 40% of unfished spawning biomass (Figure 63). The
fishing intensity (relative 1-SPR) exceeded the current estimates of the harvest rate limit (SPR50%)
throughout the 1980s and early 1990s, as seen in Figure 73. Recent exploitation rates on Widow Rockfish
were predicted to be much less than target levels. In recent years, the stock has experienced exploitation
rates that have been below the target level while the biomass level has remained above the target level
(Figure 74).

The equilibrium yield plot is shown in Figure 75, based on a steepness value fixed at 0.720. The
predicted maximum sustainable yield under the assumptions of this assessment occurs near 25% of
equilibrium unfished spawning biomass.

## Unresolved problems and major uncertainties


## Harvest Projections and Decision Tables
```{r}
load(file=here::here("report","tables","projections.rda"))
projections  <- as.data.frame(projections$table)


projections <- projections|>dplyr::mutate(across(2:ncol(projections), ~ round(.x)))


```




Projections with catches based on the predicted annual catch limit (ACL) using the SPR rate of 50%, the
40:10 control rule, and a 0.25 P* adjustment using a sigma of 0.50 from 2021 onward suggest that the
spawning biomass will decrease over the projection period for all states of nature. Predicted ACL catches
range from 10,961 mt in 2021 to 5,944 mt in 2030. 

***nOT SURE WHERE acl CATCHES ARE ***

## Evaluation of Scientific Uncertainty

## Regional management considerations


Widow Rockfish have shown latitudinal differences in life-history parameters, which has led past assessment authors to pursue a two-area model. Modelling a stock with two areas is difficult because it requires many assumptions about recruitment distribution, movement, and connectivity, while also splitting data into two areas that reduces sample sizes when compared to a coastwide model. The upside is that it can result in a better model that more accurately predicts regional status. This assessment is a coastwide model because not enough is known about the assumptions that would have to be made for a two-area model. 

It is still important to consider regional differences when making management decisions. Following recent cohorts through time with survey data showed that older fish showed up in the north after younger fish were observed in the south (Figure 2). This may indicate connectivity between the north and the south and that this is truly one stock. However, more investigation is needed.

Widow Rockfish are managed on a coastwide basis and observed more often in the NWFSC WCGBT bottom trawl survey north of latitude 40° 10′ N. Bottom trawl catches in California have historically been as large as in Oregon and larger than in Washington, but recently catches in California have been small. Rockfish Conservation Areas (RCAs) cover a significant proportion of Widow Rockfish habitat, but a midwater trawl fishery is beginning to re-develop that can fish in these areas. Future assessments and management of Widow Rockfish may want to monitor where catches are being taken to make sure that specific areas are not being overexploited. In addition, research on the connectivity along the coast as well as regional differences would help to inform the potential for overfishing specific areas.
>>>>>>> 1ed44512b0ba062144bae79bac8c50155d6b6fec

## Research and Data Needs

<!-- 1. Describe progress on Research and Data Needs items identified in the most recent previous stock assessment document and associated STAR panel report.  -->
<!-- 2. Describe new research and data needs and specify their priority (high, medium, low).  -->