
```{r}
#| include: false

library("here")
read_rdata <- function(path) {load(path); get(ls()[ls() != "path"])}

model_path <- here::here("models", "2025 base model")

report <- r4ss::SS_output(dir = model_path, verbose = FALSE, printstats = FALSE)

catch_sum <- aggregate(report$catch$Exp, by = list(Yr = report$catch$Yr), sum)

```

# Management 

## Reference Points

```{r}
#| include: false

load(file = here::here("report", "tables", "exec_summ_tables", "reference_points.rda"))
df <- as.data.frame(reference_points$table)
colnames(df) <- c("Label", "Est", "Lwr", "Upr")

ref_pts <- setNames(
  lapply(seq_len(nrow(df)), function(i) {
    vals <- df[i, c("Est", "Lwr", "Upr")]
    vals <- format(round(as.numeric(vals)), big.mark = ",", scientific = FALSE)
    setNames(as.list(vals), c("Est", "Lwr", "Upr"))
  }),
  nm = gsub("\\\\%", "%", df$Label)
)

```

Reference points were calculated using the estimated selectivity parameters and catch distribution among fleets in the five most recent years of the model (2019-2024). Sustainable total yields (landings plus discards) were `r ref_pts[["Yield with SPR50 at SB SPR (mt)"]]$Est` mt when using an SPR~50%~ reference harvest rate with a 95% confidence interval of `r ref_pts[["Yield with SPR50 at SB SPR (mt)"]]$Lwr` to `r ref_pts[["Yield with SPR50 at SB SPR (mt)"]]$Upr` mt. The \gls{ssb} equivalent to 40% of the unfished spawning output (SB~40%~) was `r ref_pts[["Proxy Spawning Biomass (mt) SB40%"]]$Est` mt. Catches between the late 1990s and 2016 were well below the point estimate of potential long-term yields calculated using an SPR~50%~ reference point, and the population is estimated to have increased continuously throughout that time period. However, catches from 2017 through 2024 were above the point estimate of potential long-term yields using an SPR~50%~ reference point (by an average `r paste0(round(100 * (mean(catch_sum$x[catch_sum$Yr >= 2017]/ df$Est[df$Label == "Yield with SPR50 at SB SPR (mt)"]) - 1)), "%")`.), exceeding the upper bound of the confidence interval in all years since 2018.

The predicted \gls{ssb} from the base model generally showed a slight decline until the late 1970s, followed by steep increase above unfished equilibrium biomass and reaching a peak in `r report$timeseries$Yr[which.max(report$timeseries$SpawnBio)]`. This was followed by a steep decrease up to the mid-1980s, and then a more gradual decease through 2000 (@fig-rel-spawn-bio). Between 2001 and 2016, the \gls{ssb} increased continuously due to small catches and several years of high recruitment (though with lower than average recruitment in other recent years). The \gls{ssb} relative to unfished equilibrium \gls{ssb} climbed above the target of 40% of unfished \gls{ssb} in the early 2000s. It is estimated to still be above the target, though the lower bound of the 95% confidence interval lies between the target and minimum stock size threshold (@fig-rel-spawn-bio). The fishing intensity (relative 1-SPR) exceeded the current estimates of the harvest rate limit (SPR~50%~) throughout the 1980s and early 1990s, and has again since 2018 (@fig-pred-one-SPR). Exploitation rates on widow rockfish between 2001 and 2016 were predicted to be far below target levels. In recent years, the stock has experienced exploitation rates that have been above the target level while the biomass level has remained above the target level (@fig-relative-one-SPR).

The equilibrium yield plot is shown in @fig-est-eqi-yield, based on a steepness value fixed at 0.720. The predicted maximum sustainable yield under the assumptions of this assessment occurs near 27% of equilibrium unfished \gls{ssb}, however this represents only 12% higher yield with considerably more risk than under the 50% SPR policy, which occurs near 45% of unfished \gls{ssb}. 

## Unresolved problems and major uncertainties

```{r}
#| include: false

steep_proflik <- read.csv(here::here("report", "tables", "SR_BH_steep_quant_table.csv"))
names(steep_proflik)[1] <- "var"

steep_yspr50 <- as.numeric(steep_proflik[steep_proflik$var == "Yield SPR 50",-1])
steep_yspr50 <- round(steep_yspr50[round(steep_yspr50) > 0])

```

Major sources of uncertainty include landings, discards, natural mortality, and recruitment, which are discussed below. 

Because widow rockfish is a marketable species, historical discard rates were likely lower than less desirable or smaller species. This assessment assumed that discarding was nearly negligible before management restrictions began in 1982. Once trip limits were introduced, discarding tended to be an all or none event; rare, but large, discard events are unlikely to have been detected given low observer coverage. From 2002 onward, the \gls{wcgop} has provided data on discards from vessels that were randomly selected for observer coverage, thus some uncertainty is present in the total amount discarded. The implementation of trawl rationalization in 2011 resulted in almost 100% observer coverage for the trawl fleet and very little incentive to discard widow rockfish. The open access fixed-gear fleet is not monitored by the full observer coverage required under trawl rationalization. Discard mortality is assumed to be 100%, which may overestimate actual mortality (Jarvis & Lowe, 2007), but given the low number of discards, is likely to have a minimal effect on assessment results.

There may also be uncertainty in the ability of bottom trawl surveys to reliably estimate abundances of widow rockfish, which spend a significant portion of their time in mid-water [@wilkins_devewpment_1986]. Multiple surveys are used in the assessment, but further consideration of additional surveys is reasonable.

This assessment attempts to capture uncertainty in M by estimating it inside the model (thereby propagating uncertainty in M into uncertainty in \gls{ssb} and other derived quantities), and by incorporating variation in M in a decision table. Model sensitivities and profiles over M showed that current stock status was highly sensitive to the assumption about natural mortality.

Widow rockfish is a relatively long-lived fish, and their natural mortality (M) is likely to be lower than many managed fish stocks (e.g., gadoids). Ages above 50 years have been observed and it is expected that natural mortality could be less than 0.10 yr^-1^ (the median of the prior for natural mortality used in this assessment). However, even with length and age data available back to the late 1970s, M was estimated at `r round(report$parameters["NatM_uniform_Fem_GP_1",]$Value, 3)` for females and `r round(report$parameters["NatM_uniform_Mal_GP_1",]$Value, 3)` yr^-1^ for males, with a small amount of uncertainty (e.g., a `r paste0(round(100 * report$parameters["NatM_uniform_Fem_GP_1",]$Parm_StDev/report$parameters["NatM_uniform_Fem_GP_1",]$Value, 1), "%")` coefficient of variation for females). 

Notably, the estimated M for both sexes from this assessment are lower than those from the 2015 and 2019 assessments, for which natural mortality was estimated above 0.15 yr^-1^ and 0.14 yr^-1^, respectively. The estimates of M varied slightly depending on the weight given to age and length data, or removing recent years of data, but M was always estimated above 0.12 yr^-1^. The likelihood profile over natural mortality provides support for values up to or above 0.14 yr^-1^, but with greater curvature than in the 2019 assessment, suggesting additional data has reduced the support for higher natural mortality values. A contributing factor to this change may be the increased mean age and frequency of older fish in catch observed by the \gls{s-wcgbt} survey in recent years. The successful rebounding of the stock as a result of reduced fishing pressure from 2003-2016 may have allowed year classes to age and become fully observed by \gls{s-wcgbt}. However, consideration should be given in future assessments to structural changes which may improve the fits to \gls{s-wcgbt} CAAL data, which are poor in some years.

Steepness was fixed at 0.720 in the base model, but a likelihood profile showed that it would be estimated at a value less than that. Estimates of M increased with lower steepness, while unfished \gls{ssb} increased and current \gls{ssb} decreased. Sustainable yields at the SPR~50%~ reference harvest rate ranged from approximately `r min(steep_yspr50)` to `r max(steep_yspr50)` mt depending on the value of steepness.

## Harvest Projections and Decision Tables

```{r}
#| include: false

projections <- read_rdata(here("report", "tables", "exec_summ_tables", "projections.rda"))$table
projections <- projections |> dplyr::mutate(across(is.numeric, ~ round(.x)))

```



```{r}
#| include: false

##dec table params
dec_table_pars <- report$parameters|> #make sure to ppull from report, not input file so est values are used
  mutate(
         value = Value,
         sd = if_else(is.na(Parm_StDev),Pr_SD,Parm_StDev))|> #If est par, use param sd, otherwise use prior sd
  filter(Num %in% c(1,13,24))|> #natM Female, natM male, steepness
  select(value,sd)|>
  mutate(
    q125 = round(qnorm(0.125,mean = value,sd = sd),3), #apply the quantiles
    q875 = round(qnorm(0.875,mean = value,sd = sd),3),
    phase = -1,
    name = c("NatM_p_1_Fem_GP_1", #name the parms - easier to do manually as report and r4ss::read() have different parameter names
             "NatM_p_1_Mal_GP_1",
             "SR_BH_steep") 
  )

#dec tables stuff
dec_table_pars["NatM_uniform_Fem_GP_1","q125"]
colnames(dec_table_pars)

dec_table_res <- read.csv(here("data_derived", "decision_table", "dec_table_formatted.csv"))

##gmt stuff
gmt_catch <- data.frame(read.csv(here("data_provided", "gmt_forecast_catch", "GMT_forecast_catch.csv"))|>
  group_by(year)|>
  summarise(total_catch = sum(catch_mt)))

dec_table_summ <- list(
  "p45_2027" = dec_table_res|>filter(scenario == "P*0.45" & yr == 2027)|>pull(catch),
  "p45_2036" = dec_table_res|>filter(scenario == "P*0.45" & yr == 2036)|>pull(catch),
  "p40_2027" = dec_table_res|>filter(scenario == "P*0.40" & yr == 2027)|>pull(catch),
  "p40_2036" = dec_table_res|>filter(scenario == "P*0.40" & yr == 2036)|>pull(catch)
)
```


Uncertainty in both natural mortality (estimated for both sexes) and steepness (not estimated in the model) contributed greatly to uncertainty in the results. A combination of these two factors was used as the axis of uncertainty to define low and high states of nature.  The 12.5% and 87.5% quantiles for female and male natural mortality (independently) were chosen as low and high values (`r dec_table_pars["NatM_uniform_Fem_GP_1","q125"]` yr^-1^ and `r dec_table_pars["NatM_uniform_Fem_GP_1","q875"]` yr^-1^ for females; `r dec_table_pars["NatM_uniform_Mal_GP_1","q125"]` yr^-1^ and `r dec_table_pars["NatM_uniform_Mal_GP_1","q875"]` yr^-1^ for males). Steepness is probably the most important factor since it was fixed in the base model and is not incorporated in the estimation uncertainty. The 12.5% and 87.5% quantiles from the steepness prior (without widow rockfish data) were used to define the low and high values of steepness (`r dec_table_pars["SR_BH_steep","q125"]` and `r dec_table_pars["SR_BH_steep","q875"]`). The low combination of these two factors defined the low state of nature and the high combination of these two factors defined the high state of nature. The predictions of \gls{ssb} in 2025 from the low and high states of nature are close to the 12.5% and 87.5% quantiles from the base model.

Previous assessments included recent large recruitment events as a third axis of uncertainty. As no similarly large and uncertain recruitment events occurred in the recent period, this axis was not included in the harvest projection.

This assessment synthesizes many sources of data and estimates recruitment variability, thus it is classified as a Category 1 stock assessment. Therefore, the sigma to determine the catch reduction to account for scientific uncertainty is 0.50.

A twelve year projection of the base model with two catch streams based on \gls{acl} adjustments of 0.45 P* and 0.40 P* were conducted (@tbl-dec-table).

Projections with catches based on the predicted \gls{acl} using the \gls{spr} rate of 50%, the 40:10 control rule, and a 0.45 P* adjustment using a sigma of 0.50 from 2027 onward suggest that the \gls{ssb} is likely to initially decrease under all states of nature before partially rebounding under the base case and high state of nature (@tbl-dec-table). Predicted ACL catches range from `r format(dec_table_summ$p45_2027, big.mark = ",", digits = 1)` mt in 2027 to `r format(dec_table_summ$p45_2036, big.mark = ",", digits = 1)` mt in 2036. 

Projections with catches based on the predicted \gls{acl} using the \gls{spr} rate of 50%, the 40:10 control rule, and a 0.40  P* adjustment using a sigma of 0.50 from 2027 onward suggest that the \gls{ssb} is likely to initially decrease under all states of nature before partially rebounding under the base case and high state of nature. Predicted ACL catches range from `r format(dec_table_summ$p40_2027, big.mark = ",", digits = 1)` mt in 2027 to `r format(dec_table_summ$p40_2036, big.mark = ",", digits = 1)` mt in 2036 @tbl-dec-table, @fig-dec-table-plot). 

## Evaluation of Scientific Uncertainty

\gls{ssb} is estimated to be at `r format(SSB["SSB_2025","Value"], big.mark = ",", digits = 1)` mt in 2025, with a sigma of `r format(SSB["SSB_2025","StdDev"]/SSB["SSB_2025","Value"], big.mark = ",", digits = 4)`. \gls{ofl} is estimated to be `r format(report$derived_quants["OFLCatch_2025","Value"], big.mark = ",", digits = 1)` mt in 2025 with a coefficient of variation of `r format(report$OFL_sigma, big.mark = ",", digits = 4)`.

## Regional management considerations

Widow rockfish are managed on a coastwide basis and observed more often in the \gls{s-wcgbt} north of latitude 40° 10′ N. Bottom trawl catches in California have historically been as large as in Oregon and larger than in Washington, but recently catches in California have been small. Future assessments and management of widow rockfish may want to monitor where catches are being taken to make sure that specific areas are not being overexploited. In addition, research on the connectivity along the coast as well as regional differences would help to inform the potential for overfishing specific areas.

## Research and Data Needs

There are many areas of research that could be improved to benefit the understanding and assessment of widow rockfish:

- **Natural mortality:** Uncertainty in natural mortality translates into uncertain estimates of status and sustainable fishing levels for widow rockfish. The collection of additional age data, re-reading of older age samples, reading old age samples that are unread, and improved understanding of the life-history of widow rockfish may reduce that uncertainty. Investigating the ageing error and bias would help to understand the influences that the age data have on this assessment, particularly the influence of age data on natural mortality.

- **Historical landings and discards:** Although progress has been made in reconstructing historical catches of rockfish on the U.S. West Coast, historical landings and discards continue to be uncertain for widow rockfish and improvements would increase the certainty that fishing removals are applied appropriately. Because landings are assumed to be known exactly in the assessment model, uncertainty in the predictions does not include uncertainty in the landings. A thorough look at historical landings, species compositions, and discarding practices would reduce the potential uncertainty that is not entirely accounted for. In addition, discard composition data have become available for the midwater trawl fleet in recent years and should be included in future assessments.

- **Sex-specific selectivity:** The midwater and bottom trawl length-composition data fits showed divergent residual patterns between male and female fish. The underlying mechanism driving this pattern is unclear, and could be related to growth, sexing error, or to sex-specific selectivity (e.g., when widow rockfish aggregate, sexes possibly may be aggregating separately). Sex-specific selectivity for these two fleets could be explored or included to address this.

- **Coastwide understanding of stock structure, biology, connectivity, and distribution:** This is a stock assessment for widow rockfish off of the west coast of the U.S. and does not consider data from British Columbia or Alaska. Further investigating and comparing the data and predictions from British Columbia and Alaska to determine if there are similarities with the U.S. West Coast observations would help to define the connectivity between widow rockfish north and south of the U.S.-Canada border.