---
title: 'Status of widow rockfish stock off the U.S. West Coast in 2025'
author:
  - name: 'Michael Kinneen'
    affiliations:
      - name: 'School of Aquatic & Fishery Sciences (SAFS), University of Washington'
        address: '1122 NE Boat St'
        city: 'Seattle, WA'
        postal-code: '98195'

  - name: 'Maurice Codespoti Goodman'
    affiliations:
      - name: 'School of Aquatic & Fishery Sciences (SAFS), University of Washington'
        address: '1122 NE Boat St'
        city: 'Seattle, WA'
        postal-code: '98195'

  - name: 'Anna Sulc'
    affiliations:
      - name: 'School of Aquatic & Fishery Sciences (SAFS), University of Washington'
        address: '1122 NE Boat St'
        city: 'Seattle, WA'
        postal-code: '98195'

  - name: 'Laurinne Balstad'
    affiliations:
      - name: 'Department of Environmental Science and Policy, University of California Davis'
        address: 'Wickson Hall, 350 N Quad'
        city: 'Davis, CA'
        postal-code: '95616'

  - name: 'Raquel Ruiz Diaz'
    affiliations:
      - name: 'School of Aquatic & Fishery Sciences (SAFS), University of Washington'
        address: '1122 NE Boat St'
        city: 'Seattle, WA'
        postal-code: '98195'

  - name: 'Kristina Randrup'
    affiliations:
      - name: 'School of Aquatic & Fishery Sciences (SAFS), University of Washington'
        address: '1122 NE Boat St'
        city: 'Seattle, WA'
        postal-code: '98195'

  - name: 'William Patrone'
    affiliations:
      - name: 'School of Aquatic & Fishery Sciences (SAFS), University of Washington'
        address: '1122 NE Boat St'
        city: 'Seattle, WA'
        postal-code: '98195'

  - name: 'Laura Spencer'
    affiliations:
      - name: 'School of Aquatic & Fishery Sciences (SAFS), University of Washington'
        address: '1122 NE Boat St'
        city: 'Seattle, WA'
        postal-code: '98195'

  - name: 'Alaia Morell'
    affiliations:
      - name: 'UW Puget Sound Institute, Center for Urban Waters'
        address: '326 East D Street'
        city: 'Tacoma, WA'
        postal-code: '98421'

  - name: 'Alberto Rovellini'
    affiliations:
      - name: 'School of Aquatic & Fishery Sciences (SAFS), University of Washington'
        address: '1122 NE Boat St'
        city: 'Seattle, WA'
        postal-code: '98195'

  - name: 'Allison Dedrick'
    affiliations:
      - name: 'California Department of Fish and Wildlife'
        address: '1740 North Market Boulevard'
        city: 'Sacramento, CA'
        postal-code: '95834'

  - name: 'Nick Grunloh'
    affiliations:
      - name: 'Southwest Fisheries Science Center'
        address: '8901 La Jolla Shores Drive'
        city: 'La Jolla, CA'
        postal-code: '92037'

  - name: 'Madison Bargas'
    affiliations:
      - name: 'Coastal Oregon Marine Experiment Station, Oregon State University'
        address: '430 Strand Agriculture Hall'
        city: 'Corvallis, OR'
        postal-code: '97331'

  - name: 'Stephanie Hopkins'
    affiliations:
      - name: "University of California, Santa Cruz, Institute of Marine Sciences’ Fisheries Collaborative Program"
        address: '1156 High Street'
        city: 'Santa Cruz, CA'
        postal-code: '95064'

  - name: 'Vladlena Gertseva'
    affiliations:
      - name: 'NOAA Fisheries Northwest Fisheries Science Center'
        address: '2725 Montlake Boulevard East'
        city: 'Seattle, WA'
        postal-code: '98112'
        
  - name: 'Kiva L. Oken'
    affiliations:
      - name: 'NOAA Fisheries Northwest Fisheries Science Center'
        address: '2725 Montlake Boulevard East'
        city: 'Seattle, WA'
        postal-code: '98112'

  - name: 'Ian G. Taylor'
    affiliations:
      - name: 'NOAA Fisheries Northwest Fisheries Science Center'
        address: '2725 Montlake Boulevard East'
        city: 'Seattle, WA'
        postal-code: '98112'

  - name: 'Melissa A. Haltuch'
    affiliations:
      - name: 'NOAA Alaska Fisheries Science Center'
        address: '7600 Sand Point Way N.E., Building 4'
        city: 'Seattle, WA'
        postal-code: '98115'
        
  - name: 'Owen Hamel'
    affiliations:
      - name: 'NOAA Fisheries Northwest Fisheries Science Center'
        address: '2725 Montlake Boulevard East'
        city: 'Seattle, WA'
        postal-code: '98112'


date: today
lang: en
keep-tex: true
mainfont: 'Latin Modern Sans'
cover: support_files/Widow_rockfish.png
pdf-engine: lualatex
format:
  pdf:
    documentclass: scrartcl
    toc: true
    number-sections: true
    template-partials:
      - 'support_files/before-body.tex'
      - 'support_files/_titlepage.tex'
    include-in-header:
      - 'support_files/in-header.tex'
      - 'support_files/pfmc.tex'
    include-after-body:
      text: |
        \newpage{}
        \printnoidxglossaries
    lof: false
    sansfont: 'Latin Modern Sans'
    lot: false
    titlepage-geometry:
      - top=1in
      - bottom=1in
      - right=1in
      - left=1in
    fig-pos: 'H'
    reference-location: section
output-file: 'WIDOW_SAR_2025'
params:
  office: 'NWFSC'
  species: 'Widow Rockfish'
  spp_latin: 'Sebastes entomelas'
  region: 'U.S. West Coast'
bibliography:
  - asar_references.bib
csl: 'canadian-journal-of-fisheries-and-aquatic-sciences.csl'
---

```{r}
#| label: 'read_base_model'
#| eval: true
#| echo: false
#| message: false
#| warning: false

library(here)
library(dplyr)

# source(here('Rscripts/table_sens.R'))

# TODO list of things to update when changing the base model:
# - [ ] run Rscripts/make_r4ss_figs_tables.R
# - [ ] run Rscripts/run_diagnostics.R
# - [ ] run Rscripts/sensitivities.R
# - [ ] update any hardwired values in the text (or make them dynamic)
#   - [ ] text on 2019 discards "inadvertently left out of the base model"
#   - [ ] 
# base_mod <- '5.09_no_extra_se'
# mod_in <- r4ss::SS_read(here('model_runs', base_mod))
# mod_out <- r4ss::SS_output(here('model_runs', base_mod), verbose = FALSE, printstats = FALSE)

```


{{< pagebreak >}} 

Please cite this publication as: 

Kinneen, M., Goodman, M. C., Sulc, A., Balstad, L., Diaz, R., Randrup, K., Patrone, W., Spencer, L., Morell, A., Rovellini, A., Dedrick, A., Grunloh, N., Bargas, M., Hopkins, S., Gersteva, V., Oken, K., Taylor, I., Haltuch, M., \& Hamel, O. (2025) Status of widow rockfish stock off the U.S. West Coast in 2025. Pacific Fishery Management Council, Portland, Oregon.

{{< pagebreak >}} 

\pagenumbering{roman}
\setcounter{page}{1}

\renewcommand{\thetable}{\roman{table}}
\renewcommand{\thefigure}{\roman{figure}}

```{r, results='asis'}
#| label: 'disclaimer'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('00a_disclaimer.qmd', quiet = TRUE)
cat(a, sep = '\n')
```

 {{< pagebreak >}} 

```{r, results='asis'}
#| label: 'executive_summary'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('01_executive_summary.qmd', quiet = TRUE)
cat(a, sep = '\n')
```


 {{< pagebreak >}}

\setlength{\parskip}{5mm plus1mm minus1mm}
\pagenumbering{arabic}
\setcounter{page}{1}
\setcounter{section}{0}
\renewcommand{\thefigure}{\arabic{figure}}
\renewcommand{\thetable}{\arabic{table}}
\setcounter{table}{0}
\setcounter{figure}{0}


```{r, results='asis'}
#| label: 'introduction'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('02_introduction.qmd', quiet = TRUE)
cat(a, sep = '\n')
```

 {{< pagebreak >}}

```{r, results='asis'}
#| label: 'data'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('03_data.qmd', quiet = TRUE)
cat(a, sep = '\n')
```

 {{< pagebreak >}}

```{r, results='asis'}
#| label: 'assessment-configuration'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('04a_model_summary.qmd', quiet = TRUE)
cat(a, sep = '\n')
```

```{r, results='asis'}
#| label: 'assessment-sensitivity'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('04b_model_results.qmd', quiet = TRUE)
cat(a, sep = '\n')
```

```{r, results='asis'}
#| label: 'assessment-benchmarks'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('04c_model_diagnostics.qmd', quiet = TRUE)
cat(a, sep = '\n')
```

 {{< pagebreak >}}

```{r, results='asis'}
#| label: 'discussion'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('05_management.qmd', quiet = TRUE)
cat(a, sep = '\n')
```

 {{< pagebreak >}}

```{r, results='asis'}
#| label: 'acknowledgments'
#| eval: false
#| echo: false
#| warning: false
a <- knitr::knit_child('06_acknowledgments.qmd', quiet = TRUE)
cat(a, sep = '\n')
```


 {{< pagebreak >}}

```{r, results='asis'}
#| label: 'tables'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('08_tables.qmd', quiet = TRUE)
cat(a, sep = '\n')
```

 {{< pagebreak >}}

```{r, results='asis'}
#| label: 'figures'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('09_figures.qmd', quiet = TRUE)
cat(a, sep = '\n')
```

 {{< pagebreak >}}

```{r, results='asis'}
#| label: 'references'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('07_references.qmd', quiet = TRUE)
cat(a, sep = '\n')
```

 <!-- {{< pagebreak >}} -->

```{r, results='asis'}
#| label: 'appendix'
#| eval: true
#| echo: false
#| warning: false
a <- knitr::knit_child('10_appendix.qmd', quiet = TRUE)
cat(a, sep = '\n')
```
