<!-- I (Kiva) suggest we free ourselves from the standard stock assessment structure and intersperse figures and text in a logical flow. -->
The bridging steps to arrive from August 2025 base model to the proposed model included:
* Adopt the new Washington historical catch reconstruction
* Simplify fleets and include discards with landings
* Update the status-quo data weighting (McAllister-Ianelli, lambda = 0.5)
1. Change data weighting methods (Francis, lambda = 1)
1. Add the fecundity relationship
1. Refine the recruitment bias adjustment and re-tune the weighting

The relatively small impact of adding the Washington historical catch reconstruction was already demonstrated through a sensitivity analysis in the August 2025 assessment report. 

Simplifying the fleet structure and combining discards with landings did not create large changes in model results, but interacted with the Francis weighting approach applied later.

The use of the Francis weighting approach with lambda = 1 resulted in the biggest change from the August 2025 base model ...

The fecundity relationship ...

Refining the recruitment bias adjustment and re-tuning the weights with the Francis method were necessary final steps but has only a small impact on the model results.  

```{r}
#| echo: false
eval_tables <- TRUE # not sure why this is needed, but helps with something about how tables are added below
source(here::here("R", "table_sens.r"))
```

```{r}
#| label: tbl-bridge2
#| warning: false
#| echo: false
#| eval: !expr eval_tables 
#| tbl-cap: "Differences in negative log-likelihood, estimates of key parameters, and estimates of derived quantities between the August base model and the bridging models (columns). Red values indicate negative log-likelihoods that were lower (fit better to that component) than the base model. Note that the likelihoods are not comparable in most cases due to changes in data."
table_sens(here::here("report", "tables", "bridging2_table.csv"))
```

```{r}
#| label: fig-bridge2_bratio
#| echo: false
#| warning: false
#| fig-cap: "Time series of fraction of unfished spawning stock biomass estimates and asymptotic 95% confidence intervals estimates bridging from August 2025 base model to the current base models."
#| out.width: "90%"
#| fig-align: "center"

knitr::include_graphics(
    here::here("figures", "bridging", "bridging2_compare3_Bratio.png")
)
```

```{r}
#| label: fig-bridge2_smrybio
#| echo: false
#| warning: false
#| fig-cap: "Time series of age 4+ biomass estimates bridging from the August 2025 base model to the current base models."
#| out.width: "90%"
#| fig-align: "center"

knitr::include_graphics(
    here::here("figures", "bridging", "bridging2_compare18_smrybio.png")
)
```

```{r}
#| label: fig-bridge2_recruits
#| echo: false
#| warning: false
#| fig-cap: "Time series of recruitment bridging from the August 2025 base model to the current base models."
#| out.width: "90%"
#| fig-align: "center"

knitr::include_graphics(
    here::here("figures", "bridging", "bridging2_compare9_recruits.png")
)
```


```{r}
#| label: fig-bridge2_recdevs
#| echo: false
#| warning: false
#| fig-cap: "Time series of recruitment deviations bridging from the August 2025 base model to the current base models."
#| out.width: "90%"
#| fig-align: "center"

knitr::include_graphics(
    here::here("figures", "bridging", "bridging2_compare11_recdevs.png")
)
```