---
title: "Explore discard data"
author: "Alaia"
date: "2025-02-05"
output: html_document
---
WCGOP data need to be downloaded in data_provided folder from google drive
# Install and load package
```{r}
library(ggplot2)
library(reshape2)
library(here)
library(dplyr)
require(tidyverse)
library(ggpubr)
library(remotes)
library(r4ss)
```

# Visualizing the Data 
## Load and plot WCGOP data 

```{r}
data_discard_non_catch <- read.csv(here("data_provided/wcgop", "discard_rates_noncatch_share.csv")) %>%
  select(year, fleet, obs_discard) %>%
  rename(observed_discard_mt = obs_discard) %>%
  mutate(fleet = recode(fleet, 
         "bottomtrawl-coastwide" = "Bottom trawl",
         "hook-and-line-coastwide" = "Hook & Line"))

data_discard_catch<- read.csv(here("data_provided/wcgop", "discard_rates_combined_catch_share.csv")) %>%
  select("year", "fleet", "observed_discard_mt")%>%
  mutate(fleet = recode(fleet, 
         "bottomtrawl-coastwide" = "Bottom trawl",
         "hook-and-line-coastwide" = "Hook & Line",
         "midwaterrockfish-coastwide" = "Midwater trawl",
         "midwaterhake-coastwide" = "Midwater trawl"))

data_discard <- rbind(data_discard_non_catch, data_discard_catch)

ggplot(data_discard, aes(x = year, y = observed_discard_mt, fill = fleet)) +
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"))+
  geom_col() +  
  theme_minimal() +
    labs(x = "Year", y = "Discards (mt)", 
       title = paste0("Discards (WCGOP)"))+
  theme(strip.text = element_text(face = "bold"))

```




## Discards length composition 
```{r}
data <- read.csv(here("data_provided/wcgop", "biological_discard_lengths.csv"))
data <- data[,-c(2,4,5,6,32:56)]
names(data) <- c("year", "fleet" , seq(8,56,2))
data <- melt(data, c("fleet", "year"))
data$variable = as.numeric(as.character(data$variable))

fleet_type <- unique(data$fleet)

ggplot(data[data$fleet==fleet_type[1],], aes(x = variable, y = value)) +
  geom_col() +
  facet_wrap(~ year, scales = "free_y") +
  labs(x = "Length bin (cm)", y = "%", 
       title = paste0("Discards by length bin by year for ",fleet_type[1])) +
  theme_minimal() +
  theme(legend.position = "none", 
        strip.text = element_text(face = "bold")) 

ggplot(data[data$fleet==fleet_type[2],], aes(x = variable, y = value)) +
  geom_col() +
  facet_wrap(~ year, scales = "free_y") +
  labs(x = "Length bin (cm)", y = "%", 
       title = paste0("Discards by length bin by year for ",fleet_type[2]))+
  theme_minimal() +
  theme(legend.position = "none", 
        strip.text = element_text(face = "bold")) 
# )
```

```{r}

# Bottom trawl
{plot(data[data$fleet==fleet_type[1],]$year, data[data$fleet==fleet_type[1],]$variable, xlab = "Year", ylab = "Length (cm)", main = "WCGOP discards by length by year (Bottom Trawl)",
     xlim = range(data[data$fleet==fleet_type[1],]$year), ylim = c(8, 62), type = "n")  
symbols(data[data$fleet==fleet_type[1],]$year, data[data$fleet==fleet_type[1],]$variable, circles = sqrt(data[data$fleet==fleet_type[1],]$value/pi), inches = 0.1, add = TRUE, bg = "white")

legend_x <- 2007
legend_y <- 60 
legend_sizes <- c(0.01, 0.4, 0.8) #Identical to 2019 updates
legend_circles <- sqrt(legend_sizes/pi)
symbols(legend_x-0.5 + (1:length(legend_sizes)) * 3,rep(legend_y+0.5, length(legend_sizes)) ,
        circles = legend_circles, inches = 0.1, add = TRUE, bg = "white")
text(legend_x + (1:length(legend_sizes)) * 3,rep(legend_y + 1, length(legend_sizes)) ,
     labels = legend_sizes, adj = 0)}



# Hook & Line
{plot(data[data$fleet==fleet_type[2],]$year, data[data$fleet==fleet_type[2],]$variable, xlab = "Year", ylab = "Length (cm)", main = "WCGOP discards by length by year (hook and line)",
     xlim = range(data[data$fleet==fleet_type[2],]$year), ylim = c(8, 62), type = "n")  
symbols(data[data$fleet==fleet_type[2],]$year, data[data$fleet==fleet_type[2],]$variable, circles = sqrt(data[data$fleet==fleet_type[2],]$value/pi), inches = 0.1, add = TRUE, bg = "white")

legend_x <- 2007
legend_y <- 60 
legend_sizes <- c(0.01, 0.4, 0.8) #Identical to 2019 updates
legend_circles <- sqrt(legend_sizes/pi)
symbols(legend_x-0.5 + (1:length(legend_sizes)) * 3,rep(legend_y+0.5, length(legend_sizes)) ,
        circles = legend_circles, inches = 0.1, add = TRUE, bg = "white")
text(legend_x + (1:length(legend_sizes)) * 3,rep(legend_y + 1, length(legend_sizes)) ,
     labels = legend_sizes, adj = 0)}

```

## Look at ALL WCGOP discard data, regardless of fleet 

```{r}
data_agg <- data %>%
  group_by(year, variable) %>%
  mutate(value = sum(value)) %>%
  distinct(year, variable, value)%>%
  group_by(year)%>%
  mutate(value = value  /sum(value)*100)

ggplot(data_agg, aes(x = variable, y = value)) +
  geom_col() +
  facet_wrap(~ year, scales = "free_y") +
  labs(x = "Length bin (cm)", y = "%", 
       title = paste0("WCGOP discards by length bin by year"))+
  theme_minimal() +
  theme(legend.position = "none", 
        strip.text = element_text(face = "bold")) 
```


```{r}
# Size comp by year
dev.new(width=5, height=20, unit="in")
{plot(data_agg$year, data_agg$variable, xlab = "Year", ylab = "Length (cm)", main = "WCGOP discards by length by year",
     xlim = range(data_agg$year), ylim = c(8, 62), type = "n")  
symbols(data_agg$year, data_agg$variable, circles = sqrt(data_agg$value/pi), inches = 0.1, add = TRUE, bg = "white")


# Legend

legend_x <- 2007
legend_y <- 60 
legend_sizes <- c(0.01, 0.4, 0.8) #Identical to 2019 updates
legend_circles <- sqrt(legend_sizes/pi)
symbols(legend_x-0.5 + (1:length(legend_sizes)) * 3,rep(legend_y+0.5, length(legend_sizes)) ,
        circles = legend_circles, inches = 0.1, add = TRUE, bg = "white")
text(legend_x + (1:length(legend_sizes)) * 3,rep(legend_y + 1, length(legend_sizes)) ,
     labels = legend_sizes, adj = 0)}
```

### Prepare length composition data for .dat file 

**Column order:** yr month fleet sex part Nsamp datavector(female-male)

Year = from 1985 - 2023. Data prior to WCGOP will not change (use the same as 2019 update assessment). Data beginning 2004 

Fleet codes: 
- 1 = BottomTrawl <--- length data used in 2019 
- 2 = MidwaterTrawl 
- 3 = Hake <-- 
- 5 = HnL <--- length data used in 2010 
Sex codes: 
 - 0 = combined <-- most of WCGOP data 
 - 1 = use female only
 - 2 = use male only
 - 3 = use both as joint sexlength distribution, a couple years 
Partition codes  
 - 0 = combined
 - 1 = discard
 - 2 = retained

25 #_N_LengthBins; then enter lower edge of each length bin
8 10 12 14 16 18 20 22 24 26 28 30 32 34 36 38 40 42 44 46 48 50 52 54 56


## OPTION 1: 
- Length comps from 1980's from the 2019 .dat file 
- Length comps 2004+ from new WCGOP data provided this year (2025)

```{r}
ssdat.2019 <- SS_readdat(file = "../data_provided/2019_assessment/2019widow.dat")

# # Discard biomass 
# ssdat.2019$discard_fleet_info #fleets = 1 (bottom trawl), 2 (midwater trawl), 5 (hook & line) 
# ssdat.2019$discard_data #biomass 

# Discard length comps - in proportion 
length.comps.1980s <- ssdat.2019$lencomp %>% filter(part==1) %>% filter(year<2004 & year>0) #remove negative years

# 2019 update assessment .dat file has Botton Trawl and Hook & Line 
ssdat.2019$lencomp %>% 
#  filter(part==2) %>% #retained 
#  filter(part==1) %>% #discarded
#  filter(part==0) %>% #combined -- fleet == 7 & 8 only, survey data 
  filter(part%in%c(1,2)) %>% 
  mutate(part=case_when(part==1~"discarded", part==2~"retained")) %>% 
  filter(year>2004) %>% 
  mutate(fleet=as.factor(as.character(fleet))) %>% 
  mutate(fleet_name=case_when(
    fleet==1~"Bottom",fleet==2~"Midwater", fleet==3~"Hake", fleet==5~"Hook&Line")) %>% 
  ggplot(aes(x=fleet_name, y=Nsamp)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size=2) +
#  theme(axis.text.x = element_text(angle = 90)) + 
  facet_wrap(~part) + ggtitle("discard lengths number samples, 2019 data file")
#  ggtitle("Number of fish measured for length comp data by fleet, WCGOP (=>2004)\n2019 .dat file")


```
Read in sample and length comps data from WCGOP, 2025 version 

Here are discard samples & aggregation summary info  - also only bottom trawn and hook&line data

```{r}
length.n <- read.csv(here("data_provided/wcgop", "../wcgop/biological_sample_sizes_length.csv"))
length.n %>% 
  ggplot(aes(x=gear_groups,y=input_n)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size=2) + 
  ggtitle("Number of fish measured for length comp data by fleet, WCGOP (=>2004)\n2025 WCGOP data provided")
```


```{r}
length.comps.wcgop <- read.csv(here("data_provided/wcgop", "../wcgop/biological_discard_lengths.csv"), col.names = names(length.comps.1980s)) %>% #use same names as .dat from 2019 for easy row binding
  mutate(month=7) %>% 
   mutate(across(matches("^[f|m]\\d+"), ~ . / 100)) %>% # get into proportions to match the 2019 .dat file (currently percentage)
  mutate(fleet=case_when(
    fleet=="bottomtrawl-coastwide"~1,
    fleet=="midwaterrockfish-coastwide"~2,
    fleet=="midwaterhake-coastwide"~3,
    fleet=="hook-and-line-coastwide"~5,
  ))

length.comps <- bind_rows(length.comps.1980s, length.comps.wcgop)
head(length.comps)

#This file contains discard length comps which we will use for the new model 
write.csv(length.comps, file = here("data_derived/discards/discard_length_comps.csv"))
```

Plot all length comps 


```{r}
# 
length.comps.agg <- length.comps %>%
  pivot_longer(cols = f8:m56, names_to = "variable", values_to = "prop") %>%
  mutate(count=round(Nsamp*prop)) %>%
  select(year, sex, variable, count) %>% 
  mutate(variable=as.numeric(gsub("m|f","",variable))) %>% 
  group_by(year,variable,sex) %>%
  mutate(value = sum(count)) %>%
  distinct(year, sex, variable, value) %>% 
  left_join(length.comps %>% group_by(year) %>% summarize(n_total=sum(Nsamp))) %>% 
  mutate(prop_total=value/n_total)

dev.new(width=5, height=20, unit="in")

# Define the years to display on the x-axis
years_to_display <- c(1985, 2004, 2007, 2010, 2013, 2016, 2019, 2022)

{plot(length.comps.agg$year, length.comps.agg$variable, xlab = "Year", ylab = "Length (cm)", 
      main = "Discards by length by year, new WCGOP data",
      xlim = range(length.comps.agg$year), ylim = c(8, 62), type = "n", 
      xaxt = "n")  # Suppress default x-axis

# Add custom x-axis with only the selected years
axis(1, at = years_to_display, labels = years_to_display)

# Add symbols for data points
symbols(length.comps.agg$year, length.comps.agg$variable, 
        circles = sqrt(length.comps.agg$prop_total/pi), inches = 0.1, 
        add = TRUE, bg = "white")

# Legend
legend_x <- 1985
legend_y <- 60
legend_sizes <- c(0.01, 0.4, 0.8)  # Identical to 2019 updates
legend_circles <- sqrt(legend_sizes/pi)
symbols(legend_x - 0.5 + (1:length(legend_sizes)) * 3, 
        rep(legend_y + 0.5, length(legend_sizes)), 
        circles = legend_circles, inches = 0.1, add = TRUE, bg = "white")
text(legend_x + (1:length(legend_sizes)) * 3, 
     rep(legend_y + 1, length(legend_sizes)), 
     labels = legend_sizes, adj = 0)
}

# tidyverse option
length.comps.agg %>% mutate(sex=as.factor(as.character(sex))) %>% 
  mutate(sex_desc=as.factor(case_when(
    sex==0~"combined",
    sex==1~"female",
    sex==2~"male",
    sex==3~"joint distribution"))) %>% 
  filter(prop_total>0) %>% 
  ggplot(aes(x = year, y = variable, col = sex_desc, size = prop_total)) +
  
  # Add vertical dotted lines at each unique x-axis year
  geom_vline(data = length.comps.agg %>% distinct(year), 
             aes(xintercept = year), 
             linetype = "dotted", color = "gray50", alpha = 0.5) +

  # Plot points
  geom_point(shape = 1, position = position_dodge(0.5)) + 
  scale_color_manual(values = c("red", "blue", "purple", "darkgreen")) + 
  #scale_size_manual(breaks = c(0.01,0.4,0.8)) + 
  theme_pubclean() + 
  ylim(c(10, 60)) + 
  ylab("Length (cm)") + 
  xlab("Year") + 
  ggtitle("Discard by lengths by year, sex, new WCGOP data") + 
  scale_x_continuous(breaks = years_to_display) + 
  theme(legend.position = "right") + 
  guides(color = guide_legend(title = NULL), size = guide_legend(title = NULL))

```

## OPTION 2: 
- Length comps through 2017 from the 2019 .dat file 
- Add only new length comps (2018+) from new WCGOP data provided this year (2025)

Regarding the -2018 values from the 2019 .dat file, this is that the SS3 manual says:

"If the value of year is negative, then that observation is not transferred into the working array. This feature is the easiest way to include observations in a data file but not to use them in a particular model scenario." 

So, the previous authors wanted to keep the 2018 data since they had it in hand, but not include it in the run. 

```{r}
length.comps.2019assess <- ssdat.2019$lencomp %>% filter(part==1) %>% filter(year>0) %>%  
  add_row(length.comps.wcgop %>% filter(year>2017)) 

length.comps.updateonly <- bind_rows(length.comps.1980s, length.comps.wcgop)

# An alternative discard length comps file, should we want to only add the new years from the new WCGOP data file 
write.csv(length.comps.updateonly, file = here("data_derived/discards/discard_length_comps_add-years-only.csv"))
```

Plot all length comps 


```{r}
# 
length.comps.agg.2019assess <- length.comps.2019assess %>%
  pivot_longer(cols = f8:m56, names_to = "variable", values_to = "prop") %>%
  mutate(count=round(Nsamp*prop)) %>%
  select(year, sex, variable, count) %>% 
  mutate(variable=as.numeric(gsub("m|f","",variable))) %>% 
  group_by(year,variable,sex) %>%
  mutate(value = sum(count)) %>%
  distinct(year, sex, variable, value) %>% 
  left_join(length.comps %>% group_by(year) %>% summarize(n_total=sum(Nsamp))) %>% 
  mutate(prop_total=value/n_total)

dev.new(width=5, height=20, unit="in")

# Define the years to display on the x-axis
years_to_display <- c(1985, 2004, 2007, 2010, 2013, 2016, 2019, 2022)

{plot(length.comps.agg.2019assess$year, length.comps.agg.2019assess$variable, xlab = "Year", ylab = "Length (cm)", 
      main = "Discards by length by year, 2018+ WCGOP data",
      xlim = range(length.comps.agg.2019assess$year), ylim = c(8, 62), type = "n", 
      xaxt = "n")  # Suppress default x-axis

# Add custom x-axis with only the selected years
axis(1, at = years_to_display, labels = years_to_display)

# Add symbols for data points
symbols(length.comps.agg.2019assess$year, length.comps.agg.2019assess$variable, 
        circles = sqrt(length.comps.agg.2019assess$prop_total/pi), inches = 0.1, 
        add = TRUE, bg = "white")

# Legend
legend_x <- 1985
legend_y <- 60
legend_sizes <- c(0.01, 0.4, 0.8)  # Identical to 2019 updates
legend_circles <- sqrt(legend_sizes/pi)
symbols(legend_x - 0.5 + (1:length(legend_sizes)) * 3, 
        rep(legend_y + 0.5, length(legend_sizes)), 
        circles = legend_circles, inches = 0.1, add = TRUE, bg = "white")
text(legend_x + (1:length(legend_sizes)) * 3, 
     rep(legend_y + 1, length(legend_sizes)), 
     labels = legend_sizes, adj = 0)
}

# tidyverse option
length.comps.agg.2019assess %>% mutate(sex=as.factor(as.character(sex))) %>% 
  mutate(sex_desc=as.factor(case_when(
    sex==0~"combined",
    sex==1~"female",
    sex==2~"male",
    sex==3~"joint distribution"))) %>% 
  filter(prop_total>0) %>% 
  ggplot(aes(x = year, y = variable, col = sex_desc, size = prop_total)) +
  
  # Add vertical dotted lines at each unique x-axis year
  geom_vline(data = length.comps.agg %>% distinct(year), 
             aes(xintercept = year), 
             linetype = "dotted", color = "gray50", alpha = 0.5) +

  # Plot points
  geom_point(shape = 1, position = position_dodge(0.5)) + 
  scale_color_manual(values = c("red", "blue", "purple", "darkgreen")) + 
  #scale_size_manual(breaks = c(0.01,0.4,0.8)) + 
  theme_pubclean() + 
  ylim(c(10, 60)) + 
  ylab("Length (cm)") + 
  xlab("Year") + 
  ggtitle("Discard by lengths by year, sex, new WCGOP data") + 
  scale_x_continuous(breaks = years_to_display) + 
  theme(legend.position = "right") + 
  guides(color = guide_legend(title = NULL), size = guide_legend(title = NULL))

```
